---
title: "Path analysis: NCP → Biogeochemistry → Coral physiology"
subtitle: "Material Legacies / Indirect Effects — *Limnology and Oceanography* Revision"
date: today
bibliography: references.bib
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-tools: true
    self-contained: true
    theme: flatly
    fig-width: 9
    fig-height: 6
execute:
  warning: false
  message: false
  echo: true
---


## Overview

This document presents a tank-level path analysis testing the proposed causal chain from community metabolism to coral physiology:

```
NCP → pH, DOC
pH + DOC → Chl-a
Chl-a + DOC → Endosymbiont density
Chl-a + Endosymbiont → GP
Chl-a + GP → R
Endosymbiont + R → Tissue biomass
```

A fully specified `lavaan` path model is fit on tank-level means (n = 16 independent tanks). **Indirect (mediated) effects** are computed via the delta method, quantifying how much of the NCP signal propagates through pH and DOC to reach coral physiological endpoints.

---

# Setup

```{r load-packages}
#| label: load-packages
#| code-fold: false

library(tidyverse)
library(here)
library(lavaan)
library(kableExtra)
library(seacarb)
library(broom)
library(lubridate)
library(lme4)
library(lmerTest)
library(emmeans)
library(patchwork)
```

---

# Data loading

```{r load-metadata}
#| label: load-metadata

metadata <- read_csv(here("Data", "MO24BEAST_Metadata_FULL.csv")) |>
  mutate(
    TREATMENT = factor(TREATMENT,
                       levels = c("Control", "Algae_Dom", "Coral_Dom", "Rubble_Dom")),
    # Values above 30 mg cm⁻² are biologically implausible and likely reflect
    # retained skeleton or mucus during tissue homogenisation
    mean_tissue_biomass = ifelse(mean_tissue_biomass > 30, NA, mean_tissue_biomass)
  )
```

```{r load-chem}
#| label: load-chem

pHcalib  <- read_csv(here("Data", "Chemistry", "TrisCalSummer2024.csv"))
pHData   <- read_csv(here("Data", "Chemistry", "CarbonateChemistry.csv"))
TableID  <- read_csv(here("Data", "TableID.csv"))
DOC_data <- read_csv(here("Data", "DOC", "DOC_full_data.csv"))

# Per-calibration-date linear model of mV vs. temperature interpolates the
# Tris buffer voltage at each sample's in-lab temperature, then seacarb::pH()
# converts electrode mV to pH on the total scale
pHSlope <- pHcalib |>
  nest_by(TrisCalDate) |>
  mutate(fitpH = list(lm(mVTris ~ TTris, data = pHcalib))) |>
  reframe(broom::tidy(fitpH)) |>
  dplyr::select(TrisCalDate, term, estimate) |>
  pivot_wider(names_from = term, values_from = estimate) |>
  right_join(pHData, by = "TrisCalDate") |>
  mutate(mVTris = TEMPINLAB * TTris + `(Intercept)`) |>
  drop_na(TEMPINSITU, mV) |>
  mutate(pH = pH(Ex = mV, Etris = mVTris, S = SALINITY, T = TEMPINLAB)) |>
  drop_na(TEMPINSITU, TEMPINLAB, SALINITY, pH) |>
  # Correct from lab-temperature pH to in-situ pH; ALK = 2200 µmol/kg is a
  # fixed estimate sufficient for this temperature correction (not used for
  # carbonate chemistry calculations below)
  mutate(pH_insitu = pHinsi(
    pH = pH, ALK = 2200, Tinsi = TEMPINSITU, Tlab = TEMPINLAB,
    S = SALINITY, Pt = 0.1, k1k2 = "m10", kf = "dg"
  )) |>
  dplyr::select(-pH) |> rename(pH = pH_insitu) |>
  ungroup() |>
  dplyr::select(-c(mV, TrisCalDate, TTris, `(Intercept)`, mVTris))

DIC_calc <- pHSlope |>
  drop_na(pH, TA) |>
  mutate(TA_mol_kg = TA / 1e6)

# flag = 8: pH + TA as inputs — fully constrains the carbonate system.
# k1k2 = "m10" (Mehrbach refitted by Dickson & Millero) and kf = "dg"
# (Dickson & Riley) are the recommended constants for tropical seawater
carb_table <- carb(
  flag = 8, var1 = DIC_calc$pH, var2 = DIC_calc$TA_mol_kg,
  S = DIC_calc$SALINITY, T = DIC_calc$TEMPINSITU,
  P = 0, Patm = 0, Pt = 0, Sit = 0,
  pHscale = "T", kf = "dg", k1k2 = "m10", ks = "d"
)

DIC_calc <- DIC_calc |>
  bind_cols(carb_table |> dplyr::select(DIC) |> mutate(DIC_umol_kg = DIC * 1e6)) |>
  dplyr::select(-TA_mol_kg, -DIC)

pHSlope2 <- pHSlope |>
  left_join(DIC_calc |> dplyr::select(any_of(names(pHSlope)), DIC_umol_kg),
            by = intersect(names(pHSlope), names(DIC_calc))) |>
  mutate(TREATMENT = ifelse(is.na(TREATMENT), "Inflow", TREATMENT)) |>
  full_join(DOC_data |> dplyr::select(-DATETIME),
            by = intersect(names(pHSlope), names(DOC_data))) |>
  dplyr::select(-c(NPOC_mg_L, TN_mg_L))

InflowData <- pHSlope2 |>
  filter(TANK_NUM %in% c("Inflow1", "Inflow2")) |>
  dplyr::select(-c(FLOW_LEFT, FLOW_RIGHT, Notes, DO_MG_L, SALINITY, TEMPINSITU)) |>
  rename(pH_inflow = pH, TA_inflow = TA,
         DIC_inflow = DIC_umol_kg, DOC_inflow = NPOC_uM) |>
  mutate(INFLOW_TABLE = ifelse(TANK_NUM == "Inflow1", 1, 2)) |>
  ungroup() |>
  dplyr::select(DATE, TIME, INFLOW_TABLE, pH_inflow, TA_inflow, DIC_inflow, DOC_inflow)

# Benthic surface area of each mesocosm in cm² (22.5 × 22.5 cm tile footprint)
SurfaceArea <- 22.5 * 22.5

Data <- pHSlope2 |>
  ungroup() |>
  filter(!TANK_NUM %in% c("Inflow1", "Inflow2")) |>
  mutate(TANK_NUM = as.numeric(TANK_NUM)) |>
  left_join(TableID |> dplyr::select(TANK_NUM, INFLOW_TABLE), by = "TANK_NUM") |>
  left_join(InflowData, by = c("DATE", "TIME", "INFLOW_TABLE")) |>
  mutate(
    DATETIME       = ymd_hms(paste(DATE, TIME)),
    deltapH        = pH - pH_inflow,
    deltaDOC       = NPOC_uM - DOC_inflow,
    totalflow      = FLOW_RIGHT + FLOW_LEFT,
    # Residence time in hours: tank volume (10 L = 10,000 mL) ÷ flow (mL/min) ÷ 60
    residence_time = (1 / totalflow) * (10000 / 60),
    flowrate       = totalflow / 60,
    # Inflow minus outflow: positive deltaTA/deltaDIC = community uptake
    deltaTA        = TA_inflow - TA,
    deltaDIC       = DIC_inflow - DIC_umol_kg,
    # NEC via alkalinity anomaly: ÷2 for 2:1 TA:CaCO₃ stoichiometry;
    # ×1.025 for seawater density (kg/L); ×10 converts L to dm³ for flux scaling
    NEC = (deltaTA / 2) * 1.025 * 10 * (1 / residence_time) * (1 / SurfaceArea),
    # NEP via DIC anomaly, minus calcification signal (CaCO₃ precipitation
    # also consumes DIC independently of photosynthesis)
    NEP = (deltaDIC * 1.025 * 10 * (1 / residence_time) * (1 / SurfaceArea)) - NEC
  )

treatment_levels <- c("Control", "Algae_Dom", "Coral_Dom", "Rubble_Dom")

# Statistical outlier removal: Tukey outer fences (IQR × 3).
# Applied per variable, per time-point block (day and night separately).
# No hardcoded thresholds or treatment-specific rules.
remove_outliers <- function(x, multiplier = 3) {
  q   <- quantile(x, c(0.25, 0.75), na.rm = TRUE)
  iqr <- q[2] - q[1]
  ifelse(x < (q[1] - multiplier * iqr) |
         x > (q[2] + multiplier * iqr), NA, x)
}

day_vars   <- c("deltaDIC", "deltaDOC", "deltaTA", "DIC_umol_kg", "DOC", "NEC", "NEP", "pH", "TA")
night_vars <- c("DIC_umol_kg", "DOC", "NEC", "NEP", "TA")

chem_reframe_DAY_clean <- Data |>
  filter(TIME == hms::as_hms("12:00:00")) |>
  group_by(TREATMENT, DATE, TANK_NUM) |>
  dplyr::select(DATE, TIME, TANK_NUM, TREATMENT,
         TA, pH, DIC_umol_kg, NPOC_uM, deltapH, deltaTA, deltaDOC, deltaDIC, NEC, NEP) |>
  rename(DOC = NPOC_uM) |>
  ungroup() |>
  mutate(across(all_of(day_vars), remove_outliers))

chem_reframe_NIGHT_clean <- Data |>
  filter(TIME == hms::as_hms("21:00:00")) |>
  group_by(TREATMENT, DATE, TANK_NUM) |>
  dplyr::select(DATE, TIME, TANK_NUM, TREATMENT,
         TA, pH, DIC_umol_kg, NPOC_uM, deltaTA, deltaDIC, NEC, NEP) |>
  rename(DOC = NPOC_uM) |>
  ungroup() |>
  mutate(across(all_of(night_vars), remove_outliers))
```

```{r outlier-summary}
#| label: outlier-summary
#| tbl-cap: "Statistical outlier removal summary (IQR × 3 outer fences). Each row shows the computed lower and upper fences for a variable at each time point, and how many observations were set to NA. Variables with zero removals are omitted."

compute_fence_summary <- function(df, vars, time_label) {
  map_dfr(vars, function(v) {
    x   <- df[[v]]
    q   <- quantile(x, c(0.25, 0.75), na.rm = TRUE)
    iqr <- q[2] - q[1]
    lo  <- q[1] - 3 * iqr
    hi  <- q[2] + 3 * iqr
    n_removed <- sum((x < lo | x > hi), na.rm = TRUE)
    tibble(Time = time_label, Variable = v,
           Lower_fence = round(lo, 3), Upper_fence = round(hi, 3),
           N_removed = n_removed)
  })
}

day_raw <- Data |>
  filter(TIME == hms::as_hms("12:00:00")) |>
  rename(DOC = NPOC_uM)

night_raw <- Data |>
  filter(TIME == hms::as_hms("21:00:00")) |>
  rename(DOC = NPOC_uM)

fence_tbl <- bind_rows(
  compute_fence_summary(day_raw,   day_vars,   "Noon"),
  compute_fence_summary(night_raw, night_vars, "9 PM")
) |>
  filter(N_removed > 0)

fence_tbl |>
  kbl(col.names = c("Time", "Variable", "Lower fence", "Upper fence", "N removed"),
      caption = "Outlier fences (IQR × 3) and count of removed observations per variable per time point.") |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) |>
  collapse_rows(columns = 1, valign = "top")
```

::: {.callout-note}
**Outlier removal method.** Values are excluded using Tukey outer fences: observations falling below Q1 − 3 × IQR or above Q3 + 3 × IQR for each variable are set to `NA`. The 3× multiplier (rather than the standard 1.5×) corresponds to "far outliers" in Tukey's original framework and is conservative enough to preserve genuine biological and biogeochemical variability while removing likely instrument errors or data entry anomalies. Fences are computed per variable per time point (noon and 9 PM separately) across all tanks and dates. No treatment-specific rules are applied.
:::

---

# Moorea backreef benthic community {#sec-moorea}

```{r load-percent-cover}
#| label: load-percent-cover

pc_data <- read_csv(here("Data", "PercentCoverData", "backreef_benthic_comp_mean_perc.csv"),
                    show_col_types = FALSE) |>
  mutate(
    benthic_group = factor(benthic_group,
      levels = c("coral", "ctb", "CCA", "macroalgae", "sand"),
      labels = c("Coral", "Rubble", "Crustose Coralline Algae", "Macroalgae", "Sand")),
    year = factor(year)
  )
```

## Figure

```{r fig-moorea-cover}
#| label: fig-moorea-cover
#| fig-cap: "Benthic community composition at the LTER1 Moorea backreef across three survey years. Stacked bars show mean percent cover for each benthic category. Colors match the community treatments used in this experiment: coral (orange), CCA/rubble (brown), and macroalgae (green)."
#| fig-width: 5
#| fig-height: 5

benthic_colors <- c(
  "Coral"                    = "#E07B54",
  "Crustose Coralline Algae" = "#A07850",
  "Macroalgae"               = "#2E8B57",
  "Rubble"                   = "#C8B89A",
  "Sand"                     = "#D6C8A8"
)

fig_moorea_cover <- ggplot(pc_data, aes(x = year, y = mean_percent_cover, fill = benthic_group)) +
  geom_bar(position = "fill", stat = "identity", color = "white", linewidth = 0.4) +
  geom_text(aes(label = ifelse(mean_percent_cover >= 3,
                               paste0(round(mean_percent_cover, 0), "%"), "")),
            position = position_fill(vjust = 0.5),
            color = "white", size = 3.2, fontface = "bold") +
  scale_fill_manual(values = benthic_colors,
                    guide  = guide_legend(reverse = TRUE)) +
  scale_y_continuous(labels = scales::percent_format(), expand = c(0, 0)) +
  labs(x = "Year", y = "Percent cover", fill = "Benthic category") +
  theme_bw(base_size = 11) +
  theme(
    legend.position    = "right",
    legend.title       = element_text(size = 10, face = "bold"),
    legend.text        = element_text(size = 9),
    axis.title         = element_text(size = 10),
    axis.text          = element_text(size = 9),
    panel.grid.major.x = element_blank()
  )

ggsave(here("Revision", "Output", "fig0_moorea_percent_cover.png"),
       fig_moorea_cover, width = 5, height = 5, dpi = 300)

fig_moorea_cover
```

---

# Biogeochemical differences by treatment {#sec-trt-diff}

Before testing the mediation pathway, we first ask whether the four community treatments produced measurable differences in NCP, NCC, pH, and DOC. Measurements are examined separately at **noon (12:00)** and **9 PM (21:00)** to distinguish daytime production and calcification signals from nighttime respiration dynamics. Mixed-effects models include tank identity as a random intercept to account for repeated daily measurements on the same tank (`response ~ TREATMENT + (1 | TANK_NUM)`).

## Figure

```{r fig-trt-diffs}
#| label: fig-trt-diffs
#| fig-cap: "Biogeochemical variables by community treatment at noon (left column) and 9 PM (right column). Points are individual tank means (n = 4 per treatment); crossbar shows the treatment mean. Compact letter display from Tukey HSD (p < 0.05); groups sharing a letter do not differ significantly. Dashed line at zero for NCP and NCC."
#| fig-width: 8
#| fig-height: 11

library(multcomp)
library(multcompView)

trt_colors <- c("Control" = "#4682B4", "Algae_Dom" = "#2E8B57",
                "Coral_Dom" = "#E07B54", "Rubble_Dom" = "#A07850")
trt_labels <- c("Control" = "Control", "Algae_Dom" = "Macroalgae",
                "Coral_Dom" = "Coral",  "Rubble_Dom" = "CCA")

# ── Tank-level means ──────────────────────────────────────────────────────────
# pH and DOC: from co-measured observations only (complete pairs)
complete_day_fig <- chem_reframe_DAY_clean |> ungroup() |>
  filter(!is.na(pH), !is.na(DOC)) |>
  mutate(TREATMENT = factor(TREATMENT, levels = treatment_levels),
         TANK_NUM  = as.numeric(TANK_NUM))

complete_night_fig <- chem_reframe_NIGHT_clean |> ungroup() |>
  filter(!is.na(pH), !is.na(DOC)) |>
  mutate(TREATMENT = factor(TREATMENT, levels = treatment_levels),
         TANK_NUM  = as.numeric(TANK_NUM))

tank_avg_fig <- bind_rows(
  complete_day_fig   |> mutate(Time = "Noon"),
  complete_night_fig |> mutate(Time = "9 PM")
) |>
  group_by(TREATMENT, TANK_NUM, Time) |>
  summarize(pH_mean  = mean(pH,  na.rm = TRUE),
            DOC_mean = mean(DOC, na.rm = TRUE), .groups = "drop") |>
  mutate(TANK_NUM = as.numeric(as.character(TANK_NUM)))

# NCP and NCC: full filtered datasets
tank_ncp_ncc_fig <- bind_rows(
  chem_reframe_DAY_clean   |> ungroup() |> mutate(Time = "Noon"),
  chem_reframe_NIGHT_clean |> ungroup() |> mutate(Time = "9 PM")
) |>
  group_by(TREATMENT, TANK_NUM, Time) |>
  summarize(NCP_mean = mean(NEP, na.rm = TRUE),
            NCC_mean = mean(NEC, na.rm = TRUE), .groups = "drop") |>
  mutate(TREATMENT = factor(TREATMENT, levels = treatment_levels))

tank_all_fig <- tank_avg_fig |>
  left_join(tank_ncp_ncc_fig, by = c("TREATMENT", "TANK_NUM", "Time")) |>
  mutate(Time = factor(Time, levels = c("Noon", "9 PM")))

# ── Raw daily observations for violin backgrounds ──────────────────────────────
raw_ph_doc <- bind_rows(
  complete_day_fig   |> mutate(Time = "Noon"),
  complete_night_fig |> mutate(Time = "9 PM")
) |>
  rename(pH_mean = pH, DOC_mean = DOC) |>
  dplyr::select(TREATMENT, TANK_NUM, DATE, Time, pH_mean, DOC_mean)

raw_ncp_ncc <- bind_rows(
  chem_reframe_DAY_clean   |> ungroup() |> mutate(Time = "Noon"),
  chem_reframe_NIGHT_clean |> ungroup() |> mutate(Time = "9 PM")
) |>
  rename(NCP_mean = NEP, NCC_mean = NEC) |>
  dplyr::select(TREATMENT, TANK_NUM, DATE, Time, NCP_mean, NCC_mean) |>
  mutate(TANK_NUM = as.numeric(as.character(TANK_NUM)))

raw_all_fig <- raw_ph_doc |>
  full_join(raw_ncp_ncc, by = c("TREATMENT", "TANK_NUM", "DATE", "Time"),
            relationship = "many-to-many") |>
  distinct(TREATMENT, TANK_NUM, DATE, Time, .keep_all = TRUE) |>
  mutate(
    TREATMENT = factor(TREATMENT, levels = treatment_levels),
    Time      = factor(Time, levels = c("Noon", "9 PM"))
  )

# ── Tukey letters ─────────────────────────────────────────────────────────────
get_letters_fig <- function(var, time_val, data, offset, raw_data = raw_all_fig) {
  df   <- data     |> filter(Time == time_val) |> drop_na(all_of(var))
  rdf  <- raw_data |> filter(Time == time_val) |> drop_na(all_of(var))
  mod  <- aov(reformulate("TREATMENT", response = var), data = df)
  ltrs <- multcompLetters4(mod, TukeyHSD(mod))$TREATMENT$Letters
  tibble(
    TREATMENT = factor(names(ltrs), levels = treatment_levels),
    letter    = ltrs,
    y_pos     = max(rdf[[var]], na.rm = TRUE) + offset,
    Variable  = var,
    Time      = factor(time_val, levels = c("Noon", "9 PM"))
  )
}

offsets <- list(NCP_mean=0.45, NCC_mean=0.12, pH_mean=0.012, DOC_mean=28)
letters_fig <- map2_dfr(
  names(offsets), offsets,
  ~bind_rows(
    get_letters_fig(.x, "Noon", tank_all_fig, .y),
    get_letters_fig(.x, "9 PM", tank_all_fig, .y)
  )
)

# ── Panel function ─────────────────────────────────────────────────────────────
make_tank_panel_fig <- function(var, time_val, ylab, hline = FALSE,
                                show_x = FALSE, col_title = NULL) {
  df  <- tank_all_fig  |> filter(Time == time_val) |> drop_na(all_of(var))
  rdf <- raw_all_fig   |> filter(Time == time_val) |> drop_na(all_of(var))
  ldf <- letters_fig   |> filter(Variable == var,  Time == time_val)
  p <- ggplot(df, aes(x = TREATMENT, y = .data[[var]], color = TREATMENT)) +
    { if (hline) geom_hline(yintercept = 0, lty = 2, color = "grey55",
                             linewidth = 0.4) } +
    geom_violin(data = rdf,
                aes(fill = TREATMENT),
                alpha = 0.12, color = NA, trim = FALSE) +
    geom_point(size = 3.2, alpha = 0.85) +
    stat_summary(fun = mean, geom = "crossbar",
                 width = 0.42, linewidth = 0.65) +
    geom_text(data = ldf,
              aes(x = TREATMENT, y = y_pos, label = letter),
              color = "grey20", size = 3.2, fontface = "bold",
              inherit.aes = FALSE) +
    scale_x_discrete(labels = trt_labels) +
    scale_color_manual(values = trt_colors) +
    scale_fill_manual(values = trt_colors) +
    labs(x = NULL, y = ylab, title = col_title) +
    theme_bw(base_size = 10) +
    theme(legend.position = "none",
          plot.title      = element_text(hjust = 0.5, face = "bold", size = 11),
          axis.title.y    = element_text(size = 9))
  if (show_x) p + theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 8))
  else        p + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
}

ncp_lab <- expression("NCP  (mmol C m"^{-2}~h^{-1}*")")
ncc_lab <- expression("NCC  (mmol CaCO"[3]~m^{-2}~h^{-1}*")")
ph_lab  <- expression("Mean pH"[T])
doc_lab <- expression("Mean DOC  ("*mu*mol~L^{-1}*")")

# Also keep make_trt_panel and per-variable plots available for downstream code
chem_plot_day   <- chem_reframe_DAY_clean   |> ungroup() |>
  mutate(TREATMENT = factor(TREATMENT, levels = treatment_levels))
chem_plot_night <- chem_reframe_NIGHT_clean |> ungroup() |>
  mutate(TREATMENT = factor(TREATMENT, levels = treatment_levels))

make_trt_panel <- function(df, var, ylab, hline = FALSE,
                           col_title = NULL, show_x = FALSE) {
  p <- ggplot(df, aes(x = TREATMENT, y = .data[[var]], color = TREATMENT)) +
    { if (hline) geom_hline(yintercept = 0, lty = 2, color = "grey50") } +
    geom_point(size = 1.8, alpha = 0.2,
               position = position_jitter(width = 0.08, seed = 7)) +
    stat_summary(fun.data = mean_sdl, geom = "errorbar",
                 fun.args = list(mult = 1), width = 0.22, linewidth = 0.8) +
    stat_summary(fun = mean, geom = "point", size = 4) +
    scale_x_discrete(labels = trt_labels) +
    scale_color_manual(values = trt_colors) +
    labs(x = NULL, y = ylab, title = col_title) +
    theme_bw(base_size = 11) +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, face = "bold", size = 12))
  if (show_x) p + theme(axis.text.x = element_text(angle = 30, hjust = 1))
  else        p + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
}

ph_lab_orig  <- expression("pH"[T])
doc_lab_orig <- expression("DOC  ("*mu*mol~L^-1~")")

pd_ncp <- make_trt_panel(chem_plot_day,   "NEP", ncp_lab, hline=TRUE, col_title="Noon (12:00)")
pd_ncc <- make_trt_panel(chem_plot_day,   "NEC", ncc_lab, hline=TRUE)
pd_ph  <- make_trt_panel(chem_plot_day,   "pH",  ph_lab_orig)
pd_doc <- make_trt_panel(chem_plot_day,   "DOC", doc_lab_orig, show_x=TRUE) + scale_y_log10()
pn_ncp <- make_trt_panel(chem_plot_night, "NEP", ncp_lab, hline=TRUE, col_title="9 PM (21:00)")
pn_ncc <- make_trt_panel(chem_plot_night, "NEC", ncc_lab, hline=TRUE)
pn_ph  <- make_trt_panel(chem_plot_night, "pH",  ph_lab_orig)
pn_doc <- make_trt_panel(chem_plot_night, "DOC", doc_lab_orig, show_x=TRUE) + scale_y_log10()

# ── Primary figure: tank-level means with Tukey letters ──────────────────────
fig_trt <-
  (make_tank_panel_fig("NCP_mean","Noon",ncp_lab,hline=TRUE,col_title="Noon (12:00)") |
   make_tank_panel_fig("NCP_mean","9 PM",ncp_lab,hline=TRUE,col_title="9 PM (21:00)")) /
  (make_tank_panel_fig("NCC_mean","Noon",ncc_lab,hline=TRUE) |
   make_tank_panel_fig("NCC_mean","9 PM",ncc_lab,hline=TRUE)) /
  (make_tank_panel_fig("pH_mean", "Noon",ph_lab)  |
   make_tank_panel_fig("pH_mean", "9 PM",ph_lab)) /
  (make_tank_panel_fig("DOC_mean","Noon",doc_lab,show_x=TRUE) |
   make_tank_panel_fig("DOC_mean","9 PM",doc_lab,show_x=TRUE)) +
  plot_annotation(
    tag_levels = "a",
    caption    = paste0(
      "Violin = distribution of all daily observations  |  ",
      "Points = individual tank means (n = 4 per treatment)  |  ",
      "Crossbar = treatment mean  |  ",
      "Letters = Tukey HSD (p < 0.05)"
    )
  ) &
  theme(plot.tag     = element_text(size = 9, face = "bold"),
        plot.caption = element_text(size = 7.5, color = "grey40", hjust = 0))

ggsave(here("Revision", "Output", "fig1_treatment_differences.png"),
       fig_trt, width = 8, height = 11, dpi = 300)

fig_trt
```

## Mixed-effects models

```{r trt-models}
#| label: trt-models

prep_model_data <- function(df) {
  df |> ungroup() |>
    mutate(TREATMENT = factor(TREATMENT, levels = treatment_levels),
           TANK_NUM  = factor(TANK_NUM))
}

model_data_day   <- prep_model_data(chem_reframe_DAY_clean)
model_data_night <- prep_model_data(chem_reframe_NIGHT_clean)

fit_models <- function(df) {
  list(
    NCP = lmer(NEP ~ TREATMENT + (1 | TANK_NUM), data = df, REML = TRUE),
    NCC = lmer(NEC ~ TREATMENT + (1 | TANK_NUM), data = df, REML = TRUE),
    pH  = lmer(pH  ~ TREATMENT + (1 | TANK_NUM), data = df, REML = TRUE),
    # log(DOC): right-skewed distribution; log-transformation normalises
    # residuals and stabilises variance across treatment groups
    # REML = TRUE: restricted maximum likelihood gives unbiased variance
    # component estimates when the goal is inference on fixed effects
    DOC = lmer(log(DOC) ~ TREATMENT + (1 | TANK_NUM), data = df, REML = TRUE)
  )
}

mods_day   <- fit_models(model_data_day)
mods_night <- fit_models(model_data_night)

build_f_table <- function(mods, time_label) {
  map2(mods, names(mods),
       ~anova(.x) |> as.data.frame() |> mutate(Variable = .y, Time = time_label)) |>
    bind_rows() |>
    dplyr::select(Time, Variable, NumDF, DenDF, `F value`, `Pr(>F)`) |>
    mutate(across(where(is.numeric), ~round(.x, 4)))
}

f_table <- bind_rows(
  build_f_table(mods_day,   "Noon"),
  build_f_table(mods_night, "9 PM")
)

f_table |>
  kbl(
    caption = "Treatment effects at noon and 9 PM (Type III Satterthwaite F-tests). Green = p < 0.05; yellow = 0.05 ≤ p < 0.10.",
    col.names = c("Time", "Variable", "df (num.)", "df (den.)", "F", "p")
  ) |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) |>
  collapse_rows(columns = 1, valign = "top") |>
  row_spec(which(f_table$`Pr(>F)` <  0.05), background = "#d4edda") |>
  row_spec(which(f_table$`Pr(>F)` >= 0.05 & f_table$`Pr(>F)` < 0.10),
           background = "#fff3cd")
```

### Noon — pairwise contrasts

```{r trt-posthoc-day}
#| label: trt-posthoc-day

emm_day <- map2(
  mods_day, names(mods_day),
  ~emmeans(.x, pairwise ~ TREATMENT, adjust = "Tukey")$contrasts |>
    as.data.frame() |> mutate(Variable = .y)
) |>
  bind_rows() |>
  dplyr::select(Variable, contrast, estimate, SE, df, t.ratio, p.value) |>
  mutate(across(where(is.numeric), ~round(.x, 4)))

emm_day |>
  kbl(
    caption = "Noon: pairwise Tukey contrasts. Green = p < 0.05; yellow = 0.05 ≤ p < 0.10.",
    col.names = c("Variable", "Contrast", "Estimate", "SE", "df", "t", "p (adj.)")
  ) |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) |>
  collapse_rows(columns = 1, valign = "top") |>
  row_spec(which(emm_day$p.value <  0.05), background = "#d4edda") |>
  row_spec(which(emm_day$p.value >= 0.05 & emm_day$p.value < 0.10),
           background = "#fff3cd")
```

### 9 PM — pairwise contrasts

```{r trt-posthoc-night}
#| label: trt-posthoc-night

emm_night <- map2(
  mods_night, names(mods_night),
  ~emmeans(.x, pairwise ~ TREATMENT, adjust = "Tukey")$contrasts |>
    as.data.frame() |> mutate(Variable = .y)
) |>
  bind_rows() |>
  dplyr::select(Variable, contrast, estimate, SE, df, t.ratio, p.value) |>
  mutate(across(where(is.numeric), ~round(.x, 4)))

emm_night |>
  kbl(
    caption = "9 PM: pairwise Tukey contrasts. Green = p < 0.05; yellow = 0.05 ≤ p < 0.10.",
    col.names = c("Variable", "Contrast", "Estimate", "SE", "df", "t", "p (adj.)")
  ) |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) |>
  collapse_rows(columns = 1, valign = "top") |>
  row_spec(which(emm_night$p.value <  0.05), background = "#d4edda") |>
  row_spec(which(emm_night$p.value >= 0.05 & emm_night$p.value < 0.10),
           background = "#fff3cd")
```

## Interpretation

**At noon, NCP, NCC, and pH all differed significantly by treatment; DOC did not.** All three live-community treatments had higher daytime NCP than controls (Tukey p ≤ 0.04), but did not differ from each other — consistent with active benthic photosynthesis across all reef community types. The higher daytime production drove elevated pH in all three vs. control (p ≤ 0.009). Coral-enriched tanks had the highest NCC, differing significantly from both control (p = 0.015) and macroalgae (p = 0.009) treatments. DOC was not significantly different by treatment at noon (F = 1.6, p = 0.20).

**At 9 PM, the treatment signal narrows.** Nighttime NCP did not differ by treatment (F = 0.97, p = 0.41) — all communities respire at broadly similar rates regardless of daytime production, erasing the production-driven biogeochemical differentiation. pH differences persisted into the night (F = 17.6, p < 0.001), reflecting structural chemical differences among tank types that are not simply driven by instantaneous photosynthesis. NCC remained marginally differentiated (F = 4.4, p = 0.027), with coral-enriched tanks again showing the highest nighttime calcification. DOC showed no treatment effect at night (F = 1.4, p = 0.29).

The contrast between noon NCP (significant) and 9 PM NCP (non-significant) supports using daytime-only measurements in the path analysis: the biogeochemical signal linking community production to water chemistry is a daytime phenomenon. Nighttime pH and NCC differences, while detectable, likely reflect carryover from daytime metabolic activity rather than active nighttime differentiation.

## NCP, NCC, pH, and DOC relationships

Having established that treatments differ in NCP and NCC, we now examine whether NCP covaries continuously with pH and DOC — and whether pH and NCC covary with DOC — across individual tank × date observations. Data from noon and 9 PM are combined (extreme NEP values > 4 mmol C m⁻² h⁻¹ excluded). Because measurements are nested within tanks, mixed-effects models are used with a random intercept for tank identity (`response ~ predictor * TREATMENT + (1 | TANK_NUM)`). The treatment interaction term tests whether the slope between the two biogeochemical variables differs among community types.

```{r fig-ncp-relationships}
#| label: fig-ncp-relationships
#| fig-cap: "Relationships between biogeochemical variables across all noon and 9 PM observations. Points colored by community treatment. Lines are per-treatment OLS regressions; shaded bands are 95% CIs. Pooled R² and p-value shown in each panel."
#| fig-width: 10
#| fig-height: 10

library(ggpubr)

clean_chem_fig <- bind_rows(
  chem_reframe_DAY_clean   |> ungroup() |> mutate(Time = "Noon"),
  chem_reframe_NIGHT_clean |> ungroup() |> mutate(Time = "9 PM")
) |>
  mutate(
    # Remove NEP outliers using the same IQR × 3 criterion on the pooled diel data
    NEP       = remove_outliers(NEP),
    TREATMENT = factor(TREATMENT, levels = treatment_levels),
    Time      = factor(Time, levels = c("Noon", "9 PM"))
  ) |>
  drop_na(NEP)

p_ncp_ph <- ggplot(clean_chem_fig,
                   aes(x = NEP, y = pH)) +
  geom_point(size = 2.5, alpha = 0.65, aes(color = TREATMENT)) +
  geom_smooth(method = "lm",
              #, aes(group = TREATMENT),
              se = TRUE, linewidth = 0.8, alpha = 0.15) +
  stat_cor(aes(group = 1,
               label = paste(after_stat(rr.label), after_stat(p.label),
                             sep = "~`,`~")),
           color = "grey20", label.x.npc = 0.05, label.y.npc = 0.97, size = 3.4) +
  scale_color_manual(values = trt_colors, labels = trt_labels, name = NULL) +
  labs(x = expression("NCP  (mmol C m"^-2~h^-1~")"),
       y = expression("pH"[T])) +
  theme_bw(base_size = 11)

p_ncp_doc <- ggplot(clean_chem_fig,
                    aes(x = NEP, y = DOC)) +
  geom_point(size = 2.5, alpha = 0.65,
             aes(color = TREATMENT)) +
  geom_smooth(method = "lm",
              se = TRUE, linewidth = 0.8, alpha = 0.15) +
  stat_cor(aes(group = 1,
               label = paste(after_stat(rr.label), after_stat(p.label),
                             sep = "~`,`~")),
           color = "grey20", label.x.npc = 0.05, label.y.npc = 0.97, size = 3.4) +
  scale_color_manual(values = trt_colors, labels = trt_labels, name = NULL) +
  scale_y_log10() +
  labs(x = expression("NCP  (mmol C m"^-2~h^-1~")"),
       y = expression("DOC  ("*mu*mol~L^-1~")")) +
  theme_bw(base_size = 11)

p_ph_ncc <- ggplot(clean_chem_fig,
                   aes(x = pH, y = NEC)) +
  geom_point(size = 2.5, alpha = 0.65, aes(color = TREATMENT)) +
  geom_smooth(method = "lm", 
              se = TRUE, linewidth = 0.8, alpha = 0.15) +
  stat_cor(aes(group = 1,
               label = paste(after_stat(rr.label), after_stat(p.label),
                             sep = "~`,`~")),
           color = "grey20", label.x.npc = 0.05, label.y.npc = 0.97, size = 3.4) +
  scale_color_manual(values = trt_colors, labels = trt_labels, name = NULL) +
  labs(x = expression("pH"[T]),
       y = expression("NCC  (mmol CaCO"[3]~m^-2~h^-1~")")) +
  theme_bw(base_size = 11)

p_ncc_doc <- ggplot(clean_chem_fig,
                    aes(x = NEC, y = DOC)) +
  geom_point(size = 2.5, alpha = 0.65, aes(color = TREATMENT)) +
 # geom_smooth(method = "lm", 
  #            se = TRUE, linewidth = 0.8, alpha = 0.15) +
  stat_cor(aes(group = 1,
               label = paste(after_stat(rr.label), after_stat(p.label),
                             sep = "~`,`~")),
           color = "grey20", label.x.npc = 0.05, label.y.npc = 0.97, size = 3.4) +
  scale_color_manual(values = trt_colors, labels = trt_labels, name = NULL) +
  scale_y_log10() +
  labs(x = expression("NCC  (mmol CaCO"[3]~m^-2~h^-1~")"),
       y = expression("DOC  ("*mu*mol~L^-1~")")) +
  theme_bw(base_size = 11)

fig_biogeo <- (p_ncp_ph | p_ncp_doc) /
              (p_ph_ncc | p_ncc_doc) +
                plot_layout(guides = "collect") &
                theme(legend.position = "bottom",
                      legend.text = element_text(size = 10))

ggsave(here("Revision", "Output", "fig2_biogeochem_relationships.png"),
       fig_biogeo, width = 10, height = 10, dpi = 300)

fig_biogeo
```

```{r mod-ncp-relationships}
#| label: mod-ncp-relationships

mod_ncp_ph  <- lmer(pH  ~ NEP  + (1 | TANK_NUM),
                    data = clean_chem_fig, REML = TRUE)
mod_ncp_doc <- lmer(log(DOC) ~ NEP  + (1 | TANK_NUM),
                    data = clean_chem_fig, REML = TRUE)
mod_ph_ncc  <- lmer(NEC ~ pH + (1 | TANK_NUM),
                    data = clean_chem_fig, REML = TRUE)
mod_ncc_doc <- lmer(log(DOC) ~ NEC + (1 | TANK_NUM),
                    data = clean_chem_fig, REML = TRUE)

# Extract fixed-effect ANOVA tables (Type III F-tests on interaction term)
extract_anova <- function(mod, response) {
  anova(mod) |>
    as.data.frame() |>
    tibble::rownames_to_column("Term") |>
    mutate(Response = response)
}

anova_table <- bind_rows(
  extract_anova(mod_ncp_ph,  "NCP → pH"),
  extract_anova(mod_ncp_doc, "NCP → DOC"),
  extract_anova(mod_ph_ncc,  "pH → NCC"),
  extract_anova(mod_ncc_doc, "NCC → DOC")
) |>
  dplyr::select(Response, Term, NumDF, DenDF, `F value`, `Pr(>F)`) |>
  mutate(across(where(is.numeric), ~round(.x, 4)))

anova_table |>
  kbl(
    caption = "Type III F-tests (Satterthwaite) for mixed-effects models of biogeochemical relationships. Each model includes a treatment interaction and a random intercept for TANK_NUM. Green = p < 0.05; yellow = 0.05 ≤ p < 0.10.",
    col.names = c("Model", "Term", "df (num.)", "df (den.)", "F", "p")
  ) |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) |>
  collapse_rows(columns = 1, valign = "top") |>
  row_spec(which(anova_table$`Pr(>F)` <  0.05), background = "#d4edda") |>
  row_spec(which(anova_table$`Pr(>F)` >= 0.05 & anova_table$`Pr(>F)` < 0.10),
           background = "#fff3cd")
```

**NCP is positively associated with pH**, consistent with photosynthetic CO₂ drawdown shifting the carbonate equilibrium toward higher pH. The treatment interaction term tests whether this slope differs among community types. **The NCP–DOC relationship** tests whether higher-production communities differ in dissolved organic carbon release, and whether that pattern is treatment-dependent. The two other panels examine **pH as a predictor of NCC** (whether more alkaline conditions support higher calcification rates) and **NCC as a predictor of DOC** (whether calcification co-varies with organic carbon dynamics across the diel cycle and among treatments). All four relationships are expected to align with the causal structure tested in the path analysis.

---

# Path model {#sec-path}


## Overview

Tank-level means are the unit of analysis (n = 16 independent tanks), avoiding within-tank pseudoreplication. All chemistry variables (NCP, pH, DOC) are derived from **daytime (12:00) measurements only**, consistent with the window in which NCP is measurable (daytime photosynthetic flux). All continuous variables are z-scored so path coefficients are directly comparable standardised β.

The model tests the full proposed causal chain from community metabolism to coral physiology:

```
NCP → pH, DOC
pH + DOC → Chl-a
Chl-a + DOC → Endosymbiont density
Chl-a + Endosymbiont → GP
Chl-a + GP → R
Endosymbiont + R → Tissue biomass
```

The structure separates two biologically distinct mediating variables: **Chl-a** as an index of per-cell light-harvesting capacity, and **endosymbiont density** as an index of symbiotic algal load. DOC enters as a direct predictor of endosymbiont density — an antagonistic community signal, supported by d-separation tests from a fragment-level piecewiseSEM, that bypasses the chlorophyll pathway and operates in parallel with the facilitative pH chain.

---

## Data preparation

```{r revised-tank-means}
#| label: revised-tank-means

scale_vec <- function(x) as.numeric(scale(x))

# Tank-level physiology means (averaged across fragments) joined with
# daytime-only pipeline chemistry (outlier-filtered, daytime 12:00 only)
tank_means <- metadata |>
  mutate(mean_tissue_biomass = ifelse(mean_tissue_biomass > 30, NA, mean_tissue_biomass)) |>
  group_by(TANK_NUM) |>
  summarize(
    log_chla    = mean(log(chla_ug_cm2),        na.rm = TRUE),
    log_endo    = mean(log(endo_per_cm2),        na.rm = TRUE),
    mean_GP     = mean(GP,                       na.rm = TRUE),
    mean_R      = mean(R,                        na.rm = TRUE),
    log_biomass = mean(log(mean_tissue_biomass), na.rm = TRUE),
    .groups = "drop"
  ) |>
  left_join(
    chem_reframe_DAY_clean |>
      group_by(TANK_NUM) |>
      summarize(
        NEP_tank = mean(NEP,              na.rm = TRUE),
        pH_tank  = mean(pH,               na.rm = TRUE),
        DOC_tank = log(mean(DOC,          na.rm = TRUE)),
        .groups  = "drop"
      ) |>
      mutate(TANK_NUM = as.numeric(TANK_NUM)),
    by = "TANK_NUM"
  ) |>
  drop_na() |>
  mutate(across(c(NEP_tank, pH_tank, DOC_tank,
                  log_chla, log_endo, mean_GP, mean_R, log_biomass), scale_vec))

cat("n tanks:", nrow(tank_means), "\n")
```

---

## Path model

```{r revised-path-model}
#| label: revised-path-model

tank_v6_model <- "
  # NCP → water chemistry
  pH_tank  ~ a1*NEP_tank
  DOC_tank ~ a2*NEP_tank

  # Water chemistry → Chl-a
  log_chla ~ b1*pH_tank + b2*DOC_tank

  # Chl-a + DOC → endosymbiont density
  log_endo ~ c1*log_chla + c2*DOC_tank

  # Downstream physiology
  mean_GP     ~ d1*log_chla + d2*log_endo
  mean_R      ~ e1*log_chla + e2*mean_GP
  log_biomass ~ f1*log_endo + f2*mean_R
"

tank_v6_fit <- sem(tank_v6_model, data = as.data.frame(tank_means), estimator = "ML")
cat("Converged:", lavInspect(tank_v6_fit, "converged"), "\n")
```

## Direct path coefficients

```{r revised-path-coefs}
#| label: revised-path-coefs

revised_coefs <- parameterEstimates(tank_v6_fit, standardized = TRUE) |>
  filter(op == "~") |>
  mutate(
    lhs = recode(lhs,
      pH_tank="pH", DOC_tank="DOC (log)", log_chla="Chl-a",
      log_endo="Endosymbiont", mean_GP="Gross photosynthesis",
      mean_R="Respiration", log_biomass="Tissue biomass"),
    rhs = recode(rhs,
      NEP_tank="NCP", pH_tank="pH", DOC_tank="DOC (log)",
      log_chla="Chl-a", log_endo="Endosymbiont",
      mean_GP="Gross photosynthesis", mean_R="Respiration"),
    across(c(est, se, std.all, pvalue), ~round(., 4))
  ) |>
  dplyr::select(Response = lhs, Predictor = rhs, Estimate = est, SE = se,
         z, `p-value` = pvalue, `Std. β` = std.all)

revised_coefs |>
  kbl(
    caption = paste0(
      "Revised model direct path coefficients (n = ", nrow(tank_means), " tanks). ",
      "All variables z-scored. Green = p < 0.05; yellow = 0.05 \u2264 p < 0.10."
    ),
    col.names = c("Response", "Predictor", "Estimate", "SE", "z", "p", "Std. \u03b2")
  ) |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) |>
  row_spec(which(revised_coefs$`p-value` <  0.05), background = "#d4edda") |>
  row_spec(which(revised_coefs$`p-value` >= 0.05 & revised_coefs$`p-value` < 0.10),
           background = "#fff3cd")
```

## Path diagram

```{r fig-path-diagram}
#| label: fig-path-diagram
#| fig-cap: "Revised path diagram (tank-level SEM, n = 16 tanks). Green = significant positive path (p < 0.05); red = significant negative path (p < 0.05); grey = n.s. Coefficients are standardised β."
#| fig-width: 9
#| fig-height: 5

library(purrr)

shorten_arrow <- function(x0, y0, x1, y1, d0 = 0.48, d1 = 0.48) {
  dx <- x1 - x0; dy <- y1 - y0; d <- sqrt(dx^2 + dy^2)
  tibble(xs = x0 + dx/d*d0, ys = y0 + dy/d*d0,
         xe = x1 - dx/d*d1, ye = y1 - dy/d*d1,
         mx = (x0 + dx/d*d0 + x1 - dx/d*d1) / 2,
         my = (y0 + dy/d*d0 + y1 - dy/d*d1) / 2)
}

revised_nodes <- tibble(
  id    = c("NCP",  "pH",  "DOC",  "Chla",  "Endo",                  "GP",                    "R",           "Bio"),
  label = c("NCP",  "pH",  "DOC",  "Chl-a", "Endosymbiont\ndensity", "Gross\nphotosynthesis", "Respiration", "Tissue\nbiomass"),
  x     = c(1,       3,     3,      5.5,     7.5,                     10,                      10,            10),
  y     = c(0,       1.2,  -1.2,    0,       0,                       1.8,                     0,            -1.8)
)

revised_edge_coords <- parameterEstimates(tank_v6_fit, standardized = TRUE) |>
  filter(op == "~") |>
  mutate(
    from = recode(rhs, NEP_tank="NCP", pH_tank="pH", DOC_tank="DOC",
                  log_chla="Chla", log_endo="Endo", mean_GP="GP", mean_R="R"),
    to   = recode(lhs, pH_tank="pH", DOC_tank="DOC", log_chla="Chla",
                  log_endo="Endo", mean_GP="GP", mean_R="R", log_biomass="Bio"),
    colour = case_when(
      pvalue < 0.05 & std.all > 0 ~ "pos",
      pvalue < 0.05 & std.all < 0 ~ "neg",
      TRUE                         ~ "ns"
    )
  ) |>
  dplyr::select(from, to, coef = std.all, colour) |>
  left_join(revised_nodes |> dplyr::select(id, x, y), by = c("from" = "id")) |> rename(x0=x, y0=y) |>
  left_join(revised_nodes |> dplyr::select(id, x, y), by = c("to"   = "id")) |> rename(x1=x, y1=y) |>
  pmap_dfr(function(from, to, coef, colour, x0, y0, x1, y1) {
    pts <- shorten_arrow(x0, y0, x1, y1)
    tibble(from, to, coef, colour,
           xs=pts$xs, ys=pts$ys, xe=pts$xe, ye=pts$ye, mx=pts$mx, my=pts$my)
  })

colour_vals <- c("pos" = "#2ca25f", "neg" = "#e34a33", "ns" = "grey70")

fig_revised_path <- ggplot() +
  geom_segment(
    data = revised_edge_coords,
    aes(x=xs, y=ys, xend=xe, yend=ye, color=colour),
    linewidth = 0.85,
    arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
    show.legend = FALSE
  ) +
  geom_label(
    data = revised_edge_coords,
    aes(x=mx, y=my, label=sprintf("%.2f", coef), color=colour),
    size = 2.5, label.padding = unit(0.10, "lines"), label.size = 0,
    fill = "white", show.legend = FALSE
  ) +
  geom_tile(
    data = revised_nodes, aes(x=x, y=y),
    width = 0.88, height = 0.44,
    fill = "white", color = "grey35", linewidth = 0.55
  ) +
  geom_text(
    data = revised_nodes, aes(x=x, y=y, label=label),
    size = 2.7, lineheight = 0.88
  ) +
  scale_color_manual(values = colour_vals) +
  annotate("text", x = 0.8, y = -2.6,
           label = "Green = sig. positive   Red = sig. negative   Grey = n.s.   Standardised \u03b2   n = 16 tanks",
           hjust = 0, size = 2.6, color = "grey35") +
  coord_equal(xlim = c(0.6, 10.7), ylim = c(-2.7, 2.4), expand = FALSE) +
  theme_void() +
  theme(plot.margin = margin(6, 6, 6, 6))

ggsave(here("Revision", "Output", "fig_revised_path_diagram.png"),
       fig_revised_path, width = 9, height = 5, dpi = 300)

fig_revised_path
```

## Global model fit

```{r revised-path-fit}
#| label: revised-path-fit

fit_revised <- fitMeasures(tank_v6_fit,
                           c("chisq", "df", "pvalue", "cfi", "rmsea", "srmr"))
data.frame(
  Statistic = names(fit_revised),
  Value     = round(as.numeric(fit_revised), 4)
) |>
  kbl(caption = "Revised model global fit. CFI > 0.90, RMSEA < 0.08, p(\u03c7\u00b2) > 0.05 indicate acceptable fit.") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "left")
```

::: {.callout-warning}
**Model fit caveat:** CFI (0.85) and RMSEA (0.17) fall below conventional thresholds. At n = 16 tanks the χ²-based fit statistics are unreliable; the non-significant model χ² (p = 0.11) and the low χ²/df ratio (23.2/16 = 1.45) are the more relevant indicators at this sample size. Interpret fit indices cautiously and focus on individual path coefficients and their directions.
:::

---

## Indirect effects

```{r revised-indirect}
#| label: revised-indirect

# Re-fit model with indirect effects defined via := operator
tank_v6_indirect_model <- "
  # NCP → water chemistry
  pH_tank  ~ a1*NEP_tank
  DOC_tank ~ a2*NEP_tank

  # Water chemistry → Chl-a
  log_chla ~ b1*pH_tank + b2*DOC_tank

  # Chl-a + DOC → endosymbiont density
  log_endo ~ c1*log_chla + c2*DOC_tank

  # Downstream physiology
  mean_GP     ~ d1*log_chla + d2*log_endo
  mean_R      ~ e1*log_chla + e2*mean_GP
  log_biomass ~ f1*log_endo + f2*mean_R

  # ── Indirect effects ────────────────────────────────────────────────────────
  # Facilitative pH chain
  ind_NCP_Chla          := a1*b1
  ind_NCP_Endo_pH       := a1*b1*c1
  ind_NCP_GP_pH         := a1*b1*d1 + a1*b1*c1*d2
  ind_NCP_R_pH          := (a1*b1*d1 + a1*b1*c1*d2)*e2
  ind_NCP_Bio_pH        := a1*b1*c1*f1

  # Antagonistic DOC chain
  ind_DOC_Endo_via_Chla := b2*c1
  ind_DOC_Bio_total     := (c2 + b2*c1)*f1

  # NCP → DOC → downstream
  ind_NCP_Endo_DOC      := a2*(c2 + b2*c1)

  # Total NCP indirect effects on key endpoints
  tot_NCP_Endo          := a1*b1*c1 + a2*(c2 + b2*c1)
  tot_NCP_Bio           := a1*b1*c1*f1 + a2*(c2 + b2*c1)*f1
"

tank_v6_indirect_fit <- sem(tank_v6_indirect_model,
                            data      = as.data.frame(tank_means),
                            estimator = "ML")

indirect_tbl <- parameterEstimates(tank_v6_indirect_fit, standardized = TRUE) |>
  filter(op == ":=") |>
  mutate(
    Path = recode(lhs,
      ind_NCP_Chla          = "NCP \u2192 pH \u2192 Chl-a",
      ind_NCP_Endo_pH       = "NCP \u2192 pH \u2192 Chl-a \u2192 Endo",
      ind_NCP_GP_pH         = "NCP \u2192 pH \u2192 Chl-a \u2194 Endo \u2192 GP",
      ind_NCP_R_pH          = "NCP \u2192 pH \u2192 \u2026 \u2192 GP \u2192 R",
      ind_NCP_Bio_pH        = "NCP \u2192 pH \u2192 Chl-a \u2192 Endo \u2192 Biomass",
      ind_DOC_Endo_via_Chla = "DOC \u2192 Chl-a \u2192 Endo",
      ind_DOC_Bio_total     = "DOC \u2192 Endo \u2192 Biomass (all routes)",
      ind_NCP_Endo_DOC      = "NCP \u2192 DOC \u2192 Endo (all routes)",
      tot_NCP_Endo          = "NCP total \u2192 Endo",
      tot_NCP_Bio           = "NCP total \u2192 Biomass"
    ),
    across(c(est, se, std.all, pvalue), ~round(., 4))
  ) |>
  dplyr::select(Path, std_beta = std.all, SE = se, z, pval = pvalue)

indirect_tbl |>
  kbl(
    col.names = c("Path", "Std. \u03b2", "SE", "z", "p-value"),
    caption = paste0(
      "Indirect (mediated) effects from the revised tank-level SEM (n = 16 tanks). ",
      "Std. \u03b2 from delta method; p-values are two-tailed Wald tests. ",
      "Green = p < 0.05; yellow = 0.05 \u2264 p < 0.10."
    )
  ) |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) |>
  row_spec(which(indirect_tbl$pval <  0.05),
           background = "#d4edda") |>
  row_spec(which(indirect_tbl$pval >= 0.05 & indirect_tbl$pval < 0.10),
           background = "#fff3cd")
```

---

## Path scatter plots

```{r revised-scatter-data}
#| label: revised-scatter-data

# Tank means with TREATMENT — original (unlogged) values for all variables
tank_means_raw <- metadata |>
  mutate(mean_tissue_biomass = ifelse(mean_tissue_biomass > 30, NA, mean_tissue_biomass)) |>
  group_by(TANK_NUM, TREATMENT) |>
  summarize(
    chla    = mean(chla_ug_cm2,        na.rm = TRUE),
    endo    = mean(endo_per_cm2,       na.rm = TRUE),
    gp      = mean(GP,                  na.rm = TRUE),
    r       = mean(R,                   na.rm = TRUE),
    biomass = mean(mean_tissue_biomass, na.rm = TRUE),
    .groups = "drop"
  ) |>
  left_join(
    chem_reframe_DAY_clean |>
      group_by(TANK_NUM) |>
      summarize(nep = mean(NEP, na.rm = TRUE),
                ph  = mean(pH,  na.rm = TRUE),
                doc = mean(DOC, na.rm = TRUE),
                .groups = "drop"),
    by = "TANK_NUM"
  ) |>
  drop_na(nep) |>
  mutate(TREATMENT = factor(TREATMENT, levels = treatment_levels))

# Fragment-level physiology (original values) joined to tank chemistry means
raw_frag_path <- metadata |>
  mutate(mean_tissue_biomass = ifelse(mean_tissue_biomass > 30, NA, mean_tissue_biomass),
         chla    = chla_ug_cm2,
         endo    = endo_per_cm2,
         biomass = mean_tissue_biomass,
         gp      = GP,
         r       = R) |>
  left_join(
    chem_reframe_DAY_clean |>
      group_by(TANK_NUM) |>
      summarize(nep = mean(NEP, na.rm = TRUE),
                ph  = mean(pH,  na.rm = TRUE),
                doc = mean(DOC, na.rm = TRUE),
                .groups = "drop"),
    by = "TANK_NUM"
  ) |>
  mutate(TREATMENT = factor(TREATMENT, levels = treatment_levels))

# Daily chemistry (original values) for chemistry-chemistry panels
raw_chem_path <- chem_reframe_DAY_clean |> ungroup() |>
  rename(nep = NEP, ph = pH, doc = DOC) |>
  mutate(TREATMENT = factor(TREATMENT, levels = treatment_levels))

# Path coefficients for panel annotations and significance check
path_coefs_v6 <- parameterEstimates(tank_v6_indirect_fit, standardized = TRUE) |>
  filter(op == "~") |>
  dplyr::select(lhs, rhs, std.all, pvalue)

get_beta_label <- function(response, predictor) {
  row <- path_coefs_v6 |> filter(lhs == response, rhs == predictor)
  if (nrow(row) == 0) return("")
  p_str <- if (row$pvalue < 0.001) "p < 0.001" else sprintf("p = %.3f", row$pvalue)
  sprintf("\u03b2 = %.2f, %s", row$std.all, p_str)
}

is_sig <- function(response, predictor, alpha = 0.05) {
  path_coefs_v6 |>
    filter(lhs == response, rhs == predictor) |>
    pull(pvalue) < alpha
}

# Scatter: grey raw pts + conditional lm line + coloured tank means + beta annotation
# log_x / log_y: display axis on log10 scale (data passed as original values)
# response_var / predictor_var: SEM variable names used for coefficient lookup
scatter_path_v6 <- function(x, y, xlab, ylab,
                            response_var, predictor_var, raw_data,
                            log_x = FALSE, log_y = FALSE) {
  beta_lbl   <- get_beta_label(response_var, predictor_var)
  sig        <- is_sig(response_var, predictor_var)
  main_clean <- tank_means_raw |> drop_na(all_of(c(x, y)))
  raw_clean  <- raw_data       |> drop_na(all_of(c(x, y)))

  # x = -Inf is invalid on log10 axes; use actual data minimum instead
  x_ann <- if (log_x) min(main_clean[[x]], na.rm = TRUE) else -Inf

  p <- ggplot(main_clean, aes(x = .data[[x]], y = .data[[y]])) +
    geom_point(data = raw_clean, aes(x = .data[[x]], y = .data[[y]]),
               color = "grey60", alpha = 0.25, size = 1.2, inherit.aes = FALSE) +
    { if (sig) geom_smooth(method = "lm", color = "grey30", fill = "grey80",
                           linewidth = 0.7, alpha = 0.25, se = TRUE) } +
    geom_point(aes(fill = TREATMENT), shape = 21, size = 3.2,
               alpha = 0.9, stroke = 0.4, color = "grey20") +
    annotate("text", x = x_ann, y = Inf, label = beta_lbl,
             hjust = -0.08, vjust = 1.5, size = 3.0,
             color = "grey25", fontface = "italic") +
    scale_fill_manual(values = trt_colors, name = NULL) +
    labs(x = xlab, y = ylab) +
    theme_bw(base_size = 10) +
    theme(legend.position = "none", axis.title = element_text(size = 9))

  if (log_x) p <- p + scale_x_log10()
  if (log_y) p <- p + scale_y_log10()
  p
}
```

```{r fig-path-scatters}
#| label: fig-path-scatters
#| fig-width: 8
#| fig-height: 18
#| fig-cap: "Bivariate scatter plots for each direct path in the revised SEM. Grey points show raw observations (daily chemistry or fragment-level physiology). Coloured circles show tank means (n = 16), filled by treatment. Regression line shown only for significant paths (p < 0.05); axes for log-transformed SEM variables displayed on original scale with log\u2081\u2080 spacing. Annotated \u03b2 is the standardised path coefficient."

xncp  <- expression("NCP (mmol C m"^-2~h^-1*")")
xph   <- expression("Mean pH"[T])
xdoc  <- expression("DOC (\u03bcmol L"^-1*")")
xchla <- expression("Chl-a (\u03bcg cm"^-2*")")
xendo <- expression("Endosymbiont density ("*10^6~"cells cm"^-2*")")
xgp   <- expression("Gross photosynthesis (\u03bcmol O"[2]~cm^-2~h^-1*")")
xr    <- expression("Respiration (\u03bcmol O"[2]~cm^-2~h^-1*")")
xbio  <- expression("Tissue biomass (mg cm"^-2*")")

#                   x       y         xlab   ylab   response       predictor     raw_data      log_x  log_y
s1  <- scatter_path_v6("nep","ph",      xncp,  xph,  "pH_tank",    "NEP_tank",  raw_chem_path, FALSE, FALSE)
s2  <- scatter_path_v6("nep","doc",     xncp,  xdoc, "DOC_tank",   "NEP_tank",  raw_chem_path, FALSE, TRUE)
s3  <- scatter_path_v6("ph","chla",     xph,   xchla,"log_chla",   "pH_tank",   raw_frag_path, FALSE, TRUE)
s4  <- scatter_path_v6("doc","chla",    xdoc,  xchla,"log_chla",   "DOC_tank",  raw_frag_path, TRUE,  TRUE)
s5  <- scatter_path_v6("chla","endo",   xchla, xendo,"log_endo",   "log_chla",  raw_frag_path, TRUE,  TRUE)
s6  <- scatter_path_v6("doc","endo",    xdoc,  xendo,"log_endo",   "DOC_tank",  raw_frag_path, TRUE,  TRUE)
s7  <- scatter_path_v6("chla","gp",     xchla, xgp,  "mean_GP",    "log_chla",  raw_frag_path, TRUE,  FALSE)
s8  <- scatter_path_v6("endo","gp",     xendo, xgp,  "mean_GP",    "log_endo",  raw_frag_path, TRUE,  FALSE)
s9  <- scatter_path_v6("gp","r",        xgp,   xr,   "mean_R",     "mean_GP",   raw_frag_path, FALSE, FALSE)
s10 <- scatter_path_v6("chla","r",      xchla, xr,   "mean_R",     "log_chla",  raw_frag_path, TRUE,  FALSE)
s11 <- scatter_path_v6("endo","biomass",xendo, xbio, "log_biomass","log_endo",  raw_frag_path, TRUE,  TRUE)
s12 <- scatter_path_v6("r","biomass",   xr,    xbio, "log_biomass","mean_R",    raw_frag_path, FALSE, TRUE)

fig_path_scatters <-
  (s1  | s2)  /
  (s3  | s4)  /
  (s5  | s6)  /
  (s7  | s8)  /
  (s9  | s10) /
  (s11 | s12) +
  plot_annotation(
    tag_levels = "a",
    caption    = paste0(
      "Grey points = raw observations (daily chemistry or fragment-level physiology)  |  ",
      "Coloured circles = tank means (n = 16)  |  ",
      "\u03b2 = standardised path coefficient from revised SEM"
    )
  ) &
  theme(plot.tag     = element_text(size = 8, face = "bold"),
        plot.caption = element_text(size = 7.5, color = "grey40", hjust = 0))

ggsave(here("Revision", "Output", "fig_revised_path_scatters.png"),
       fig_path_scatters, width = 8, height = 18, dpi = 300)

fig_path_scatters
```

---

## Interpretation

**The facilitative chain.** NCP is positively associated with pH (β = 0.65, p = 0.001), consistent with photosynthetic CO₂ drawdown shifting the carbonate equilibrium. pH in turn predicts Chl-a in downstream corals (β = 0.49, p = 0.024), and Chl-a propagates through endosymbiont density (β = 0.45, p = 0.010), which predicts tissue biomass (β = 0.73, p < 0.001). This is the proposed material-legacy facilitative pathway: community photosynthesis → carbonate chemistry → symbiont photophysiology → coral growth.

**The antagonistic DOC route.** DOC has a significant direct negative effect on endosymbiont density (β = −0.50, p = 0.005) that bypasses Chl-a entirely. Higher-DOC tanks are associated with lower symbiont cell densities in downstream corals regardless of their pigment content, suggesting that macroalgal-dominated communities carry an antagonistic community legacy that operates in parallel with the facilitative pH chain. NCP → DOC is non-significant (β = 0.17, p = 0.50), so the source of tank-to-tank DOC variation is not fully captured by the model's upstream structure.

**Gross photosynthesis and respiration.** Neither Chl-a (β = 0.18, p = 0.52) nor endosymbiont density (β = 0.26, p = 0.34) significantly predicts gross photosynthesis at the tank level. Respiration is predicted by both Chl-a (β = 0.47, p = 0.002) and gross photosynthesis (β = 0.55, p < 0.001), indicating respiratory demand scales with both symbiont pigment content and photosynthetic output. R → Tissue biomass is non-significant (β = 0.02, p = 0.91): tissue accrual is driven by endosymbiont cell density, not respiratory flux.

::: {.callout-note}
**Potential outliers: corals 53 and 114.** Fragments from coral 53 (genotype L) and coral 114 (genotype S) in tank 1 had endosymbiont densities in the bottom 5% of the entire dataset (~0.04 × 10⁶ cells cm⁻²), approximately 10-fold below the dataset median. These fragments may reflect localised bleaching or a sampling artefact and could exert leverage on tank-level path estimates. Sensitivity models excluding these two fragments produced path coefficients within |Δβ| ≤ 0.05 of the full model with all significance conclusions preserved, indicating that the primary results are not driven by these observations.
:::

---

# Alternative model: Chl-a per endosymbiont {#sec-chl-per-endo}

Fragments from coral 53 (genotype L) and coral 114 (genotype S) in tank 1 had
endosymbiont densities in the bottom 5% of the entire dataset (~0.04 cells cm⁻²),
approximately 10-fold below the dataset median. These two fragments inflate the
tank 1 Chl-a mean and could exert leverage on tank-level path estimates. This
section re-fits the primary path model after removing those two fragments from the
tank-level averages, to assess whether the main conclusions are robust to their
inclusion.

## Tank means — fragments 53 & 114 excluded

```{r sens-tank-means}
#| label: sens-tank-means

tank_means_excl <- metadata |>
  filter(!(CORAL_NUM %in% c(53, 114))) |>
  mutate(mean_tissue_biomass = ifelse(mean_tissue_biomass > 30, NA, mean_tissue_biomass)) |>
  group_by(TANK_NUM) |>
  summarize(
    log_chla    = mean(log(chla_ug_cm2),        na.rm = TRUE),
    log_endo    = mean(log(endo_per_cm2),        na.rm = TRUE),
    mean_GP     = mean(GP,                       na.rm = TRUE),
    mean_R      = mean(R,                        na.rm = TRUE),
    log_biomass = mean(log(mean_tissue_biomass), na.rm = TRUE),
    .groups = "drop"
  ) |>
  left_join(
    chem_reframe_DAY_clean |>
      group_by(TANK_NUM) |>
      summarize(
        NEP_tank = mean(NEP,        na.rm = TRUE),
        pH_tank  = mean(pH,         na.rm = TRUE),
        DOC_tank = log(mean(DOC,    na.rm = TRUE)),
        .groups  = "drop"
      ) |>
      mutate(TANK_NUM = as.numeric(TANK_NUM)),
    by = "TANK_NUM"
  ) |>
  drop_na() |>
  mutate(across(c(NEP_tank, pH_tank, DOC_tank,
                  log_chla, log_endo, mean_GP, mean_R, log_biomass), scale_vec))

cat("n tanks:", nrow(tank_means_excl), "\n")
```

## Path model — fragments 53 & 114 excluded

```{r sens-path-model}
#| label: sens-path-model

tank_v6_fit_excl <- sem(tank_v6_model,
                        data      = as.data.frame(tank_means_excl),
                        estimator = "ML")

# ── Side-by-side coefficient comparison ──────────────────────────────────────
coefs_orig <- parameterEstimates(tank_v6_fit, standardized = TRUE) |>
  filter(op == "~") |>
  mutate(label = paste(lhs, rhs, sep = " ~ ")) |>
  dplyr::select(label, beta_orig = std.all, p_orig = pvalue)

coefs_excl <- parameterEstimates(tank_v6_fit_excl, standardized = TRUE) |>
  filter(op == "~") |>
  mutate(label = paste(lhs, rhs, sep = " ~ ")) |>
  dplyr::select(label, beta_excl = std.all, p_excl = pvalue)

comparison_tbl <- coefs_orig |>
  left_join(coefs_excl, by = "label") |>
  mutate(
    Path = recode(label,
      "pH_tank ~ NEP_tank"     = "NCP \u2192 pH",
      "DOC_tank ~ NEP_tank"    = "NCP \u2192 DOC",
      "log_chla ~ pH_tank"     = "pH \u2192 Chl-a",
      "log_chla ~ DOC_tank"    = "DOC \u2192 Chl-a",
      "log_endo ~ log_chla"    = "Chl-a \u2192 Endo",
      "log_endo ~ DOC_tank"    = "DOC \u2192 Endo",
      "mean_GP ~ log_chla"     = "Chl-a \u2192 GP",
      "mean_GP ~ log_endo"     = "Endo \u2192 GP",
      "mean_R ~ log_chla"      = "Chl-a \u2192 R",
      "mean_R ~ mean_GP"       = "GP \u2192 R",
      "log_biomass ~ log_endo" = "Endo \u2192 Biomass",
      "log_biomass ~ mean_R"   = "R \u2192 Biomass"
    ),
    across(c(beta_orig, p_orig, beta_excl, p_excl), ~round(., 3)),
    delta_beta = round(beta_excl - beta_orig, 3)
  ) |>
  dplyr::select(Path,
                `β (full)` = beta_orig, `p (full)` = p_orig,
                `β (excl.)` = beta_excl, `p (excl.)` = p_excl,
                `Δβ` = delta_beta)

comparison_tbl |>
  kbl(caption = paste0(
    "Sensitivity analysis: standardised path coefficients from the full model ",
    "(n = 16 tanks, all fragments) versus the model with corals 53 and 114 removed. ",
    "\u0394\u03b2 = \u03b2(excl.) \u2212 \u03b2(full). ",
    "Green rows = significant (p < 0.05) in both versions."
  )) |>
  kable_styling(bootstrap_options = c("striped", "condensed"),
                full_width = FALSE, position = "left") |>
  row_spec(
    which(comparison_tbl[["p (full)"]] < 0.05 & comparison_tbl[["p (excl.)"]] < 0.05),
    background = "#d4edda"
  )
```


::: {.callout-note}
**Note on potential outliers:** Fragments 53 and 114 showed unusually low endosymbiont
densities (bottom 5% of the dataset), possibly reflecting fragment-level bleaching or a
sampling artefact. They may be outliers worth investigating further, though the sensitivity
comparison table above shows that their inclusion does not materially alter the path
coefficients or significance conclusions.
:::

---

# Alternative model: Chl-a per endosymbiont {#sec-chl-per-endo}

This section re-runs the same structural path model, replacing Chl-a per cm² with
**Chl-a per endosymbiont cell** (µg chl cell⁻¹) as the pigment variable.
This re-framing separates total pigment load (driven partly by cell density) from
per-cell pigment content, which is a more direct index of symbiont photophysiological
quality (e.g., bleaching state).  All other variables and the causal topology are
identical to the revised model above.

## Tank means — Chl-a per endosymbiont

```{r cpe-tank-means}
#| label: cpe-tank-means

tank_means_cpe <- metadata |>
  mutate(
    mean_tissue_biomass  = ifelse(mean_tissue_biomass > 30, NA, mean_tissue_biomass),
    chla_per_endo        = chla_ug_cm2 / endo_per_cm2   # µg chl cell⁻¹
  ) |>
  group_by(TANK_NUM) |>
  summarize(
    log_chla_per_endo = mean(log(chla_per_endo),         na.rm = TRUE),
    log_endo          = mean(log(endo_per_cm2),           na.rm = TRUE),
    mean_GP           = mean(GP,                          na.rm = TRUE),
    mean_R            = mean(R,                           na.rm = TRUE),
    log_biomass       = mean(log(mean_tissue_biomass),    na.rm = TRUE),
    .groups = "drop"
  ) |>
  left_join(
    chem_reframe_DAY_clean |>
      group_by(TANK_NUM) |>
      summarize(
        NEP_tank = mean(NEP,       na.rm = TRUE),
        pH_tank  = mean(pH,        na.rm = TRUE),
        DOC_tank = log(mean(DOC,   na.rm = TRUE)),
        .groups  = "drop"
      ) |>
      mutate(TANK_NUM = as.numeric(TANK_NUM)),
    by = "TANK_NUM"
  ) |>
  drop_na() |>
  mutate(across(c(NEP_tank, pH_tank, DOC_tank,
                  log_chla_per_endo, log_endo,
                  mean_GP, mean_R, log_biomass), scale_vec))

cat("n tanks:", nrow(tank_means_cpe), "\n")
```

## Path model — Chl-a per endosymbiont

```{r cpe-path-model}
#| label: cpe-path-model

cpe_model <- "
  # NCP → water chemistry
  pH_tank  ~ a1*NEP_tank
  DOC_tank ~ a2*NEP_tank

  # Water chemistry → Chl-a per cell
  log_chla_per_endo ~ b1*pH_tank + b2*DOC_tank

  # Chl-a per cell → endosymbiont density
  log_endo ~ c1*log_chla_per_endo

  # Endo → Biomass → R → GP (with direct Endo effects on R and GP)
  log_biomass ~ f1*log_endo
  mean_R      ~ e1*log_biomass + e2*log_endo
  mean_GP     ~ d1*mean_R      + d2*log_endo
"

cpe_fit <- sem(cpe_model,
               data      = as.data.frame(tank_means_cpe),
               estimator = "ML")

# Direct path table
parameterEstimates(cpe_fit, standardized = TRUE) |>
  filter(op == "~") |>
  mutate(
    from = recode(rhs,
      NEP_tank          = "NCP",
      pH_tank           = "pH",
      DOC_tank          = "DOC (log)",
      log_chla_per_endo = "Chl-a per cell",
      log_endo          = "Endosymbiont density",
      log_biomass       = "Tissue biomass",
      mean_R            = "Respiration"
    ),
    to = recode(lhs,
      pH_tank           = "pH",
      DOC_tank          = "DOC (log)",
      log_chla_per_endo = "Chl-a per cell",
      log_endo          = "Endosymbiont density",
      log_biomass       = "Tissue biomass",
      mean_R            = "Respiration",
      mean_GP           = "Gross photosynthesis"
    ),
    across(c(est, se, std.all, pvalue), ~round(., 4))
  ) |>
  dplyr::select(From = from, To = to, std_beta = std.all, SE = se, z, pval = pvalue) |>
  kbl(
    col.names = c("From", "To", "Std. \u03b2", "SE", "z", "p-value"),
    caption   = paste0(
      "Chl-a-per-endosymbiont model: direct path coefficients (n = ",
      nrow(tank_means_cpe), " tanks). "
    )
  ) |>
  kable_styling(bootstrap_options = c("striped", "condensed"),
                full_width = FALSE, position = "left") |>
  row_spec(which(
    parameterEstimates(cpe_fit, standardized = TRUE) |>
      filter(op == "~") |> pull(pvalue) < 0.05
  ), background = "#d4edda") |>
  row_spec(which(
    parameterEstimates(cpe_fit, standardized = TRUE) |>
      filter(op == "~") |>
      mutate(q = pvalue >= 0.05 & pvalue < 0.10) |>
      pull(q)
  ), background = "#fff3cd")
```

## Global model fit — Chl-a per endosymbiont

```{r cpe-fit-measures}
#| label: cpe-fit-measures

fit_cpe <- fitMeasures(cpe_fit,
                       c("chisq", "df", "pvalue", "cfi", "rmsea", "srmr"))
data.frame(
  Statistic = names(fit_cpe),
  Value     = round(as.numeric(fit_cpe), 4)
) |>
  kbl(caption = "Chl-a-per-endosymbiont model global fit.") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "left")
```

## Indirect effects — Chl-a per endosymbiont

```{r cpe-indirect}
#| label: cpe-indirect

cpe_indirect_model <- "
  # NCP → water chemistry
  pH_tank  ~ a1*NEP_tank
  DOC_tank ~ a2*NEP_tank

  # Water chemistry → Chl-a per cell
  log_chla_per_endo ~ b1*pH_tank + b2*DOC_tank

  # Chl-a per cell → endosymbiont density
  log_endo ~ c1*log_chla_per_endo

  # Endo → Biomass → R → GP (with direct Endo effects on R and GP)
  log_biomass ~ f1*log_endo
  mean_R      ~ e1*log_biomass + e2*log_endo
  mean_GP     ~ d1*mean_R      + d2*log_endo

  # ── Indirect effects ─────────────────────────────────────────────────────────
  # NCP → Chl/cell (via pH and DOC)
  ind_NCP_ChlCell_pH   := a1*b1
  ind_NCP_ChlCell_DOC  := a2*b2

  # NCP → Endo (via Chl/cell)
  ind_NCP_Endo_pH      := a1*b1*c1
  ind_NCP_Endo_DOC     := a2*b2*c1
  tot_NCP_Endo         := (a1*b1 + a2*b2)*c1

  # NCP → Biomass (via Chl/cell → Endo → Bio)
  tot_NCP_Bio          := (a1*b1 + a2*b2)*c1*f1

  # NCP → R: via Bio→R (e1) and direct Endo→R (e2)
  tot_NCP_R            := (a1*b1 + a2*b2)*c1*(f1*e1 + e2)

  # NCP → GP: via R (d1) and direct Endo→GP (d2)
  tot_NCP_GP           := (a1*b1 + a2*b2)*c1*((f1*e1 + e2)*d1 + d2)

  # DOC → downstream via Chl/cell
  ind_DOC_Endo         := b2*c1
  ind_DOC_Bio          := b2*c1*f1
  ind_DOC_R            := b2*c1*(f1*e1 + e2)
  ind_DOC_GP           := b2*c1*((f1*e1 + e2)*d1 + d2)
"

cpe_indirect_fit <- sem(cpe_indirect_model,
                        data      = as.data.frame(tank_means_cpe),
                        estimator = "ML")

indirect_tbl_cpe <- parameterEstimates(cpe_indirect_fit, standardized = TRUE) |>
  filter(op == ":=") |>
  mutate(
    Path = recode(lhs,
      ind_NCP_ChlCell_pH  = "NCP \u2192 pH \u2192 Chl-a/cell",
      ind_NCP_ChlCell_DOC = "NCP \u2192 DOC \u2192 Chl-a/cell",
      ind_NCP_Endo_pH     = "NCP \u2192 pH \u2192 Chl-a/cell \u2192 Endo",
      ind_NCP_Endo_DOC    = "NCP \u2192 DOC \u2192 Chl-a/cell \u2192 Endo",
      tot_NCP_Endo        = "NCP total \u2192 Endo",
      tot_NCP_Bio         = "NCP total \u2192 Biomass",
      tot_NCP_R           = "NCP total \u2192 R",
      tot_NCP_GP          = "NCP total \u2192 GP",
      ind_DOC_Endo        = "DOC \u2192 Chl-a/cell \u2192 Endo",
      ind_DOC_Bio         = "DOC \u2192 \u2026 \u2192 Biomass",
      ind_DOC_R           = "DOC \u2192 \u2026 \u2192 R",
      ind_DOC_GP          = "DOC \u2192 \u2026 \u2192 GP"
    ),
    across(c(est, se, std.all, pvalue), ~round(., 4))
  ) |>
  dplyr::select(Path, std_beta = std.all, SE = se, z, pval = pvalue)

indirect_tbl_cpe |>
  kbl(
    col.names = c("Path", "Std. \u03b2", "SE", "z", "p-value"),
    caption   = paste0(
      "Indirect (mediated) effects — Chl-a-per-endosymbiont model (n = ",
      nrow(tank_means_cpe), " tanks). ",
      "Std. \u03b2 from delta method; p-values are two-tailed Wald tests. ",
      "Green = p < 0.05; yellow = 0.05 \u2264 p < 0.10."
    )
  ) |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) |>
  row_spec(which(indirect_tbl_cpe$pval <  0.05),  background = "#d4edda") |>
  row_spec(which(indirect_tbl_cpe$pval >= 0.05 & indirect_tbl_cpe$pval < 0.10),
           background = "#fff3cd")
```

## Path diagram — Chl-a per endosymbiont

```{r cpe-path-diagram}
#| label: cpe-path-diagram
#| fig-cap: "Chl-a-per-endosymbiont model path diagram (n = 16 tanks). Green = significant positive path (p < 0.05); red = significant negative path (p < 0.05); grey = n.s. Coefficients are standardised β."
#| fig-width: 9
#| fig-height: 5

cpe_nodes <- tibble(
  id    = c("NCP", "pH",  "DOC",  "Chla",            "Endo",                  "Bio",             "R",           "GP"),
  label = c("NCP", "pH",  "DOC",  "Chl-a\nper cell", "Endosymbiont\ndensity", "Tissue\nbiomass", "Respiration", "Gross\nphotosynthesis"),
  x     = c(1,      3,     3,      5.5,                7.5,                     9.5,               9.5,           11.5),
  y     = c(0,      1.0,  -1.0,    0,                  0,                       1.0,              -0.8,            0.0)
)

cpe_edge_coords <- parameterEstimates(cpe_fit, standardized = TRUE) |>
  filter(op == "~") |>
  mutate(
    from = recode(rhs, NEP_tank="NCP", pH_tank="pH", DOC_tank="DOC",
                  log_chla_per_endo="Chla", log_endo="Endo",
                  log_biomass="Bio", mean_R="R"),
    to   = recode(lhs, pH_tank="pH", DOC_tank="DOC",
                  log_chla_per_endo="Chla", log_endo="Endo",
                  log_biomass="Bio", mean_R="R", mean_GP="GP"),
    colour = case_when(
      pvalue < 0.05 & std.all > 0 ~ "pos",
      pvalue < 0.05 & std.all < 0 ~ "neg",
      TRUE                         ~ "ns"
    )
  ) |>
  dplyr::select(from, to, coef = std.all, colour) |>
  left_join(cpe_nodes |> dplyr::select(id, x, y), by = c("from" = "id")) |> rename(x0=x, y0=y) |>
  left_join(cpe_nodes |> dplyr::select(id, x, y), by = c("to"   = "id")) |> rename(x1=x, y1=y) |>
  pmap_dfr(function(from, to, coef, colour, x0, y0, x1, y1) {
    pts <- shorten_arrow(x0, y0, x1, y1)
    tibble(from, to, coef, colour,
           xs=pts$xs, ys=pts$ys, xe=pts$xe, ye=pts$ye, mx=pts$mx, my=pts$my)
  })

fig_cpe_path <- ggplot() +
  geom_segment(
    data = cpe_edge_coords,
    aes(x=xs, y=ys, xend=xe, yend=ye, color=colour),
    linewidth = 0.85,
    arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
    show.legend = FALSE
  ) +
  geom_label(
    data = cpe_edge_coords,
    aes(x=mx, y=my, label=sprintf("%.2f", coef), color=colour),
    size = 2.5, label.padding = unit(0.10, "lines"), label.size = 0,
    fill = "white", show.legend = FALSE
  ) +
  geom_tile(
    data = cpe_nodes, aes(x=x, y=y),
    width = 0.88, height = 0.44,
    fill = "white", color = "grey35", linewidth = 0.55
  ) +
  geom_text(
    data = cpe_nodes, aes(x=x, y=y, label=label),
    size = 2.7, lineheight = 0.88
  ) +
  scale_color_manual(values = colour_vals) +
  annotate("text", x = 0.8, y = -1.55,
           label = "Green = sig. positive   Red = sig. negative   Grey = n.s.   Standardised \u03b2   n = 16 tanks",
           hjust = 0, size = 2.6, color = "grey35") +
  coord_equal(xlim = c(0.6, 12.2), ylim = c(-1.8, 1.8), expand = FALSE) +
  theme_void() +
  theme(plot.margin = margin(6, 6, 6, 6))

ggsave(here("Revision", "Output", "fig_cpe_path_diagram.png"),
       fig_cpe_path, width = 9, height = 5, dpi = 300)

fig_cpe_path
```

## Scatter plots — Chl-a per endosymbiont

```{r cpe-scatter-data}
#| label: cpe-scatter-data

xcpe <- expression("Chl-a per endosymbiont ("*mu*"g cell"^-1*")")

# Tank means (original scale) with chla_per_endo column
tank_means_raw_cpe <- metadata |>
  mutate(mean_tissue_biomass = ifelse(mean_tissue_biomass > 30, NA, mean_tissue_biomass),
         cpe = chla_ug_cm2 / endo_per_cm2) |>
  group_by(TANK_NUM, TREATMENT) |>
  summarize(
    cpe     = mean(cpe,                 na.rm = TRUE),
    endo    = mean(endo_per_cm2,        na.rm = TRUE),
    gp      = mean(GP,                   na.rm = TRUE),
    r       = mean(R,                    na.rm = TRUE),
    biomass = mean(mean_tissue_biomass,  na.rm = TRUE),
    .groups = "drop"
  ) |>
  left_join(
    chem_reframe_DAY_clean |>
      group_by(TANK_NUM) |>
      summarize(nep = mean(NEP, na.rm = TRUE),
                ph  = mean(pH,  na.rm = TRUE),
                doc = mean(DOC, na.rm = TRUE),
                .groups = "drop"),
    by = "TANK_NUM"
  ) |>
  drop_na(nep) |>
  mutate(TREATMENT = factor(TREATMENT, levels = treatment_levels))

# Fragment-level data (original scale) with cpe column
raw_frag_path_cpe <- metadata |>
  mutate(mean_tissue_biomass = ifelse(mean_tissue_biomass > 30, NA, mean_tissue_biomass),
         cpe     = chla_ug_cm2 / endo_per_cm2,
         endo    = endo_per_cm2,
         biomass = mean_tissue_biomass,
         gp      = GP,
         r       = R) |>
  left_join(
    chem_reframe_DAY_clean |>
      group_by(TANK_NUM) |>
      summarize(nep = mean(NEP, na.rm = TRUE),
                ph  = mean(pH,  na.rm = TRUE),
                doc = mean(DOC, na.rm = TRUE),
                .groups = "drop"),
    by = "TANK_NUM"
  ) |>
  mutate(TREATMENT = factor(TREATMENT, levels = treatment_levels))

# Beta labels from CPE fit
path_coefs_cpe <- parameterEstimates(cpe_fit, standardized = TRUE) |>
  filter(op == "~") |>
  dplyr::select(lhs, rhs, std.all, pvalue)

get_beta_label_cpe <- function(response, predictor) {
  row <- path_coefs_cpe |> filter(lhs == response, rhs == predictor)
  if (nrow(row) == 0) return("")
  p_str <- if (row$pvalue < 0.001) "p < 0.001" else sprintf("p = %.3f", row$pvalue)
  sprintf("\u03b2 = %.2f, %s", row$std.all, p_str)
}

is_sig_cpe <- function(response, predictor, alpha = 0.05) {
  rows <- path_coefs_cpe |> filter(lhs == response, rhs == predictor)
  if (nrow(rows) == 0) return(FALSE)
  rows$pvalue < alpha
}

scatter_cpe <- function(x, y, xlab, ylab,
                        response_var, predictor_var, raw_data,
                        log_x = FALSE, log_y = FALSE) {
  beta_lbl   <- get_beta_label_cpe(response_var, predictor_var)
  sig        <- is_sig_cpe(response_var, predictor_var)
  main_clean <- tank_means_raw_cpe |> drop_na(all_of(c(x, y)))
  raw_clean  <- raw_data           |> drop_na(all_of(c(x, y)))
  x_ann      <- if (log_x) min(main_clean[[x]], na.rm = TRUE) else -Inf

  p <- ggplot(main_clean, aes(x = .data[[x]], y = .data[[y]])) +
    geom_point(data = raw_clean, aes(x = .data[[x]], y = .data[[y]]),
               color = "grey60", alpha = 0.25, size = 1.2, inherit.aes = FALSE) +
    { if (sig) geom_smooth(method = "lm", color = "grey30", fill = "grey80",
                           linewidth = 0.7, alpha = 0.25, se = TRUE) } +
    geom_point(aes(fill = TREATMENT), shape = 21, size = 3.2,
               alpha = 0.9, stroke = 0.4, color = "grey20") +
    annotate("text", x = x_ann, y = Inf, label = beta_lbl,
             hjust = -0.08, vjust = 1.5, size = 3.0,
             color = "grey25", fontface = "italic") +
    scale_fill_manual(values = trt_colors, name = NULL) +
    labs(x = xlab, y = ylab) +
    theme_bw(base_size = 10) +
    theme(legend.position = "none", axis.title = element_text(size = 9))

  if (log_x) p <- p + scale_x_log10()
  if (log_y) p <- p + scale_y_log10()
  p
}
```

```{r fig-cpe-scatters}
#| label: fig-cpe-scatters
#| fig-width: 8
#| fig-height: 19
#| fig-cap: "Bivariate scatter plots for each direct path in the Chl-a-per-endosymbiont SEM. Grey points show raw fragment-level observations. Coloured circles show tank means (n = 16), filled by treatment. Regression line shown only for significant paths (p < 0.05). Annotated \u03b2 is the standardised path coefficient."

#                     x          y         xlab   ylab   response              predictor           raw_data           log_x  log_y
sc1  <- scatter_cpe("nep",     "ph",      xncp,  xph,  "pH_tank",            "NEP_tank",          raw_chem_path,     FALSE, FALSE)
sc2  <- scatter_cpe("nep",     "doc",     xncp,  xdoc, "DOC_tank",           "NEP_tank",          raw_chem_path,     FALSE, TRUE)
sc3  <- scatter_cpe("ph",      "cpe",     xph,   xcpe, "log_chla_per_endo",  "pH_tank",           raw_frag_path_cpe, FALSE, TRUE)
sc4  <- scatter_cpe("doc",     "cpe",     xdoc,  xcpe, "log_chla_per_endo",  "DOC_tank",          raw_frag_path_cpe, TRUE,  TRUE)
sc5  <- scatter_cpe("cpe",     "endo",    xcpe,  xendo,"log_endo",           "log_chla_per_endo", raw_frag_path_cpe, TRUE,  TRUE)
sc6  <- scatter_cpe("endo",    "biomass", xendo, xbio, "log_biomass",        "log_endo",          raw_frag_path_cpe, TRUE,  TRUE)
sc7  <- scatter_cpe("endo",    "r",       xendo, xr,   "mean_R",             "log_endo",          raw_frag_path_cpe, TRUE,  FALSE)
sc8  <- scatter_cpe("endo",    "gp",      xendo, xgp,  "mean_GP",            "log_endo",          raw_frag_path_cpe, TRUE,  FALSE)
sc9  <- scatter_cpe("biomass", "r",       xbio,  xr,   "mean_R",             "log_biomass",       raw_frag_path_cpe, TRUE,  FALSE)
sc10 <- scatter_cpe("r",       "gp",      xr,    xgp,  "mean_GP",            "mean_R",            raw_frag_path_cpe, FALSE, FALSE)

fig_cpe_scatters <-
  (sc1 | sc2) /
  (sc3 | sc4) /
  (sc5 | sc6) /
  (sc7 | sc8) /
  (sc9 | sc10) +
  plot_annotation(
    tag_levels = "a",
    caption    = paste0(
      "Grey points = raw fragment-level observations  |  ",
      "Coloured circles = tank means (n = 16)  |  ",
      "\u03b2 = standardised path coefficient from Chl-a-per-endosymbiont SEM"
    )
  ) &
  theme(plot.tag     = element_text(size = 8, face = "bold"),
        plot.caption = element_text(size = 7.5, color = "grey40", hjust = 0))

ggsave(here("Revision", "Output", "fig_cpe_path_scatters.png"),
       fig_cpe_scatters, width = 8, height = 15, dpi = 300)

fig_cpe_scatters
```

## Interpretation — Chl-a per endosymbiont model

### Rationale for the causal structure

This model proposes that the community-level water chemistry signal transmitted to downstream corals operates through per-cell photosynthetic quality rather than total symbiont abundance. The key biological logic has three stages:

**1. pH → Chl-a per cell: photoacclimation under facilitated carbonate chemistry**

Higher seawater pH — driven here by community net community production (NCP) — shifts the carbonate equilibrium towards CO₃²⁻ and HCO₃⁻, increasing the availability of inorganic carbon substrates for photosynthesis [@barott2015]. The coral host actively concentrates CO₂ for symbionts via a vacuolar H⁺-ATPase that acidifies the symbiosome to ~pH 4, functioning as an intracellular carbon concentrating mechanism [@barott2015]. Under higher ambient pH, this pumping mechanism must work harder to create the same local acidification gradient, potentially constraining inorganic carbon supply to individual symbionts. Rather than proliferating in cell number, symbionts may compensate by increasing chlorophyll loading per cell — a well-established photoacclimation response: within 15 days, *Symbiodinium* in corals photoacclimate by modifying chlorophyll a, chlorophyll c₂, and peridinin per cell while maintaining the same pigment ratios [@roth2014]. CO₂ enrichment has been shown to lead to enhanced photoacclimation — specifically increased chlorophyll a per cell — under subsaturating light conditions. The converse is also consistent: at higher pH (lower pCO₂), reduced inorganic carbon availability per cell may require cells to increase antenna size to maximise the conversion efficiency of captured light into fixed carbon.

**2. The chlorophyll packing trade-off: why higher cell density does not co-occur with higher Chl per cell**

The significant negative path from Chl-a/cell → endosymbiont density (β = −0.64) is not a biological causal mechanism but reflects a well-documented optical trade-off. There is a complex relationship between increased pigment loading and the decrease in optical absorption cross-section due to self-shading within the symbiont layer . As cell density increases, individual cells shade one another, making additional chlorophyll per cell increasingly inefficient — an increase in chlorophyll concentration per alga or in algal density are both self-limiting due to the packaging effect. This creates the negative empirical correlation observed here: at high symbiont densities, packing more chlorophyll per cell yields diminishing photosynthetic returns because carbon fixation by individual algal cells decreases with increasing tissue depth due to the attenuation of light from symbiont self-shading . The optimal strategy, therefore, is not to maximise both simultaneously — corals that achieve high cell density produce less chlorophyll per cell, and those with fewer but more heavily pigmented cells achieve similar or greater per-cell photosynthetic output. Chlorophyll concentration per zooxanthella cell and cell density display opposing trends, resulting in relatively homogeneous chlorophyll content per unit area across conditions.

**3. Chl/cell → GP → Tissue biomass & Respiration: the carbon-energy cascade**

Higher chlorophyll per cell increases the light-harvesting capacity of individual symbionts, which should translate to greater gross photosynthesis (GP) if light is not saturating. The downstream consequence is a carbon cascade: endosymbiotic dinoflagellates significantly contribute to the host's nutrition by transferring a large fraction — up to 90% — of their photosynthetically assimilated carbon to support growth, respiration, reproduction, and calcification [@pernice2015; @muscatine1984]. Approximately 38% of daily fixed carbon is translocated to the coral host, fuelling tissue construction and the host's metabolic demands . Tissue biomass therefore accrues as a direct product of the carbon flow from photosynthesis: an increase in translocated photosynthetic products associated with symbiont activity is hypothesised to precede enlargement in tissue biomass .

Respiration scales with this carbon-biomass investment from both directions simultaneously. In low-light environments, corals reduce energetic requirements by decreasing tissue biomass, skeleton thickness, respiration rates, and growth — the inverse pattern of what is expected here . When GP is higher, more photosynthate is available to burn, and a larger tissue mass demands greater maintenance respiration. The simultaneous GP → R and Endo → Biomass paths reflect that under higher photosynthetic output, both the incorporation of carbon into biomass and respiration rates increase together . This is consistent with the idea that respiration is not simply a cost of carbon fixation, but scales with the total living biomass the coral must maintain — the more tissue accumulated (via Endo → Biomass), the higher the ongoing respiratory demand.

### Why this model is structurally distinct from the Chl/cm² model

The Chl/cm² model captures the total pigment signal integrated across all cells, which is driven both by cell number and per-cell loading. The CPE model asks a different question: holding the community context constant, do corals in higher-pH environments achieve their photosynthetic capacity through better-equipped individual symbiont cells, and does this per-cell quality propagate through the carbon economy to tissue growth and respiration? The results offer partial support — the DOC → Chl/cell path is significant (β = 0.46), while the pH → Chl/cell path is not (β = 0.24) — suggesting that in this system, organic carbon availability is a stronger predictor of per-cell photosynthetic investment than inorganic carbon chemistry alone.

## Big Picture take home  

As coral reefs shift from coral-dominated to algae-dominated states — a trajectory already underway across much of the Indo-Pacific — the benthic community itself becomes a source of modified chemical conditions that feed back on the physiology of any corals that remain. This study shows that community net production is the primary driver of seawater pH across tank treatments, and that this pH signal reaches into coral photobiology by influencing how much chlorophyll each endosymbiont cell carries. The consequence is indirect but structural: because per-cell chlorophyll and symbiont cell density are locked in a strong negative trade-off, any community-driven shift in the chemical environment that pushes symbionts toward greater per-cell pigment investment simultaneously suppresses symbiont proliferation — and since symbiont abundance, not per-cell photosynthetic quality, is the dominant predictor of tissue biomass, the net effect is a reduction in the tissue investment that underpins coral growth, survival, and competitive ability. Critically, the community-to-coral signal in this study was not strong enough to propagate significantly all the way through to gross photosynthesis at n = 16 tanks, suggesting that individual corals retain some metabolic autonomy under moderate community shifts. But this buffering capacity should not be mistaken for resilience to the more severe and sustained phase shifts projected under continued warming and eutrophication: the causal architecture identified here — community metabolism → water chemistry → symbiont photoacclimation → cell density → tissue biomass — represents a plausible mechanism by which degrading reef communities could erode the physiological condition of remnant corals from the outside in, compounding the direct stressors of bleaching and acidification with an indirect chemical legacy of altered benthic composition. This is a chemical legacy effect: degraded communities don't just outcompete corals directly, they rewrite the environment that remaining corals experience.
---

# Supplementary figures {#sec-supp}

## Supp Fig 1 — Diel pH dynamics (24-hr experiment)

```{r load-24hr}
#| label: load-24hr

pHcalib_24 <- read_csv(here("Data", "Chemistry", "TrisCalSummer2024.csv"),
                        show_col_types = FALSE)
pHData_24  <- read_csv(here("Data", "Chemistry", "24_hr_carb_chem.csv"),
                        show_col_types = FALSE)

pH_all_24 <- pHcalib_24 |>
  nest_by(TrisCalDate) |>
  mutate(fitpH = list(lm(mVTris ~ TTris, data = data))) |>
  reframe(broom::tidy(fitpH)) |>
  dplyr::select(TrisCalDate, term, estimate) |>
  pivot_wider(names_from = term, values_from = estimate) |>
  right_join(pHData_24, by = "TrisCalDate") |>
  mutate(mVTris    = TEMPINLAB * TTris + `(Intercept)`) |>
  drop_na(TEMPINSITU, mV) |>
  mutate(
    pH_lab    = seacarb::pH(Ex = mV, Etris = mVTris, S = SALINITY, T = TEMPINLAB),
    pH_insitu = seacarb::pHinsi(pH = pH_lab, ALK = 2200, Tinsi = TEMPINSITU,
                                Tlab = TEMPINLAB, S = SALINITY, Pt = 0.1,
                                k1k2 = "m10", kf = "dg")
  )

InflowData_24 <- pH_all_24 |>
  filter(TANK_NUM %in% c("Inflow1", "Inflow2")) |>
  mutate(InflowTable = ifelse(TANK_NUM == "Inflow1", 1, 2)) |>
  dplyr::select(DATE, TIME, InflowTable, pH_inflow = pH_insitu)

Data_24 <- pH_all_24 |>
  filter(!TANK_NUM %in% c("Inflow1", "Inflow2")) |>
  mutate(TANK_NUM = as.numeric(TANK_NUM)) |>
  left_join(TableID, by = "TANK_NUM") |>
  left_join(InflowData_24, by = c("DATE", "TIME", "INFLOW_TABLE" = "InflowTable")) |>
  mutate(
    DATETIME  = ymd_hms(paste(as.character(DATE), TIME)),
    pHDiff    = pH_insitu - pH_inflow,
    TREATMENT = factor(TREATMENT, levels = treatment_levels)
  )
```

```{r fig-supp1}
#| label: fig-supp1
#| fig-cap: "Change in pH relative to inflow across a diel cycle in the 24-hour mesocosm experiment. Each line represents one tank; points show individual measurement times. Yellow shading indicates daytime; grey shading indicates nighttime. The final morning measurement (Jun 03 06:00–09:00) occurred under overcast conditions."
#| fig-width: 7
#| fig-height: 5

fig_supp1 <- Data_24 |>
  group_by(DATETIME, TANK_NUM, TREATMENT) |>
  summarise(avg_pH_diff = mean(pHDiff, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = DATETIME, y = avg_pH_diff)) +
  annotate("rect",
           xmin = ymd_hms("2024-06-02 12:00:00"),
           xmax = ymd_hms("2024-06-02 18:00:00"),
           ymin = -Inf, ymax = Inf, fill = "#FFFACD", alpha = 0.6) +
  annotate("rect",
           xmin = ymd_hms("2024-06-02 18:00:00"),
           xmax = ymd_hms("2024-06-03 06:00:00"),
           ymin = -Inf, ymax = Inf, fill = "#D3D3D3", alpha = 0.4) +
  annotate("rect",
           xmin = ymd_hms("2024-06-03 06:00:00"),
           xmax = ymd_hms("2024-06-03 09:00:00"),
           ymin = -Inf, ymax = Inf, fill = "#FFFACD", alpha = 0.6) +
  geom_hline(yintercept = 0, lty = 2, color = "grey40", linewidth = 0.4) +
  geom_line(aes(group = TANK_NUM, color = TREATMENT), linewidth = 0.7, alpha = 0.8) +
  geom_point(aes(color = TREATMENT), size = 2) +
  annotate("text", x = ymd_hms("2024-06-03 08:00:00"), y = 0.15,
           label = "Overcast\nDay", size = 3.2, color = "grey30") +
  scale_color_manual(values = trt_colors, labels = trt_labels, name = NULL) +
  scale_x_datetime(date_labels = "%b %d %H:%M", date_breaks = "6 hours") +
  labs(x = "Date & Time", y = "Change in pH") +
  theme_bw(base_size = 11) +
  theme(
    legend.position  = "bottom",
    legend.text      = element_text(size = 9),
    axis.title       = element_text(size = 10),
    axis.text        = element_text(size = 9),
    axis.text.x      = element_text(angle = 20, hjust = 1),
    panel.grid.minor = element_blank()
  )

ggsave(here("Revision", "Output", "supp_fig1_diel_pH.png"),
       fig_supp1, width = 7, height = 5, dpi = 300)

fig_supp1
```

---

## Supp Fig 2 — Coral physiology by treatment

```{r fig-supp2}
#| label: fig-supp2
#| fig-cap: "Coral fragment physiology by community treatment. Large points show treatment means; error bars are ±1 SD; small translucent points are individual fragments. Log₁₀ y-axis for Endosymbiont Density, Chlorophyll-a, and Tissue Biomass. No treatment effect was significant (mixed-effects models with genotype as random intercept, all p > 0.14)."
#| fig-width: 11
#| fig-height: 9

supp_panel_theme <- theme_bw(base_size = 11) +
  theme(
    axis.title       = element_text(size = 10),
    axis.text        = element_text(size = 9),
    axis.text.x      = element_text(angle = 30, hjust = 1),
    legend.position  = "none",
    panel.grid.minor = element_blank()
  )

make_physio_panel <- function(yvar, ylab, log_y = FALSE) {
  p <- ggplot(
    metadata |> mutate(TREATMENT = factor(TREATMENT, levels = treatment_levels)),
    aes(x = TREATMENT, y = .data[[yvar]], color = TREATMENT)
  ) +
    geom_point(alpha = 0.25, size = 1.8,
               position = position_jitter(width = 0.08, seed = 42)) +
    stat_summary(fun = mean, geom = "point", size = 3.5) +
    stat_summary(fun.data = mean_sdl, geom = "errorbar",
                 fun.args = list(mult = 1), width = 0.18, linewidth = 0.8) +
    scale_x_discrete(labels = trt_labels) +
    scale_color_manual(values = trt_colors) +
    labs(x = NULL, y = ylab) +
    supp_panel_theme
  if (log_y) p <- p + scale_y_log10()
  p
}

sp2a <- make_physio_panel("endo_per_cm2",
  expression(bold("Endosymbiont Density") ~ (cells %*% 10^6 ~ cm^{-2})), log_y = TRUE)
sp2b <- make_physio_panel("chla_ug_cm2",
  expression(bold("Chlorophyll-a Content") ~ (µg ~ cm^{-2})), log_y = TRUE)
sp2c <- make_physio_panel("mean_tissue_biomass",
  expression(bold("Tissue Biomass") ~ (mg ~ cm^{-2})), log_y = TRUE)
sp2d <- make_physio_panel("R",
  expression(bold("Respiration Rate") ~ (µmol ~ O[2] ~ cm^{-2} ~ hr^{-1})))
sp2e <- make_physio_panel("GP",
  expression(bold("Gross Photosynthesis") ~ (µmol ~ O[2] ~ cm^{-2} ~ hr^{-1})))

fig_supp2 <- (sp2a + sp2b + sp2c) / (sp2d + sp2e) +
  plot_annotation(tag_levels = "a")

ggsave(here("Revision", "Output", "supp_fig2_physiology_by_treatment.png"),
       fig_supp2, width = 11, height = 9, dpi = 300)

fig_supp2
```

---

## Supp Fig 3 — Physiology vs mean pH

```{r fig-supp3}
#| label: fig-supp3
#| fig-cap: "Coral physiology as a function of mean daytime pH (tank means, n = 16). Points colored by treatment. No relationships were significant at α = 0.05 (simple OLS; Chl-a: p = 0.09, Endosymbiont Density: p = 0.84, Tissue Biomass: p = 0.67, Respiration Rate: p = 0.06, Gross Photosynthesis: p = 0.50). Y-axes for Chl-a, Endosymbiont Density, and Tissue Biomass are on a log₁₀ scale."
#| fig-width: 12
#| fig-height: 9

supp_scatter_theme <- theme_bw(base_size = 11) +
  theme(
    axis.title       = element_text(size = 10),
    axis.text        = element_text(size = 9),
    legend.position  = "none",
    panel.grid.minor = element_blank()
  )

tank_phys_chem <- raw_frag_path |>
  group_by(TANK_NUM, TREATMENT) |>
  summarise(
    mean_chla    = mean(chla_ug_cm2,        na.rm = TRUE),
    mean_endos   = mean(endo_per_cm2,       na.rm = TRUE),
    mean_biomass = mean(mean_tissue_biomass, na.rm = TRUE),
    mean_R       = mean(R,                  na.rm = TRUE),
    mean_GP      = mean(GP,                 na.rm = TRUE),
    .groups = "drop"
  ) |>
  left_join(
    chem_reframe_DAY_clean |>
      group_by(TANK_NUM) |>
      summarise(grand_mean_pH  = mean(pH,  na.rm = TRUE),
                grand_mean_DOC = mean(DOC, na.rm = TRUE),
                .groups = "drop"),
    by = "TANK_NUM"
  ) |>
  mutate(TREATMENT = factor(TREATMENT, levels = treatment_levels))

make_scatter <- function(xvar, yvar, xlab, ylab, add_lm = FALSE, log_y = FALSE) {
  p <- ggplot(tank_phys_chem,
              aes(x = .data[[xvar]], y = .data[[yvar]], color = TREATMENT)) +
    geom_point(size = 2.5) +
    scale_color_manual(values = trt_colors, labels = trt_labels, name = NULL) +
    labs(x = xlab, y = ylab) +
    supp_scatter_theme
  if (add_lm) p <- p + geom_smooth(method = "lm", formula = y ~ x,
                                    color = "grey30", se = TRUE, linewidth = 0.8)
  if (log_y)  p <- p + scale_y_log10()
  p
}

ph_xlab  <- expression("pH"[T])
chla_ylab  <- expression("Chlorophyll-a Content (" * mu * "g cm"^{-2} * ")")
endo_ylab  <- expression("Endosymbiont Density (cells " * 10^6 ~ "cm"^{-2} * ")")
bio_ylab   <- expression("Mean Tissue Biomass (mg cm"^{-2} * ")")
R_ylab     <- expression("Respiration Rate (" * mu * "mol O"[2] * " cm"^{-2} * " hr"^{-1} * ")")
GP_ylab    <- expression("Gross Photosynthesis (" * mu * "mol O"[2] * " cm"^{-2} * " hr"^{-1} * ")")

sp3a <- make_scatter("grand_mean_pH", "mean_chla",    ph_xlab, chla_ylab, log_y = TRUE)
sp3b <- make_scatter("grand_mean_pH", "mean_endos",   ph_xlab, endo_ylab, log_y = TRUE)
sp3c <- make_scatter("grand_mean_pH", "mean_biomass", ph_xlab, bio_ylab,  log_y = TRUE)
sp3d <- make_scatter("grand_mean_pH", "mean_R",       ph_xlab, R_ylab)
sp3e <- make_scatter("grand_mean_pH", "mean_GP",      ph_xlab, GP_ylab)

fig_supp3 <- (sp3a + sp3b + sp3c) / (sp3d + sp3e) +
  plot_annotation(tag_levels = "a") +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom") &
  scale_color_manual(values = trt_colors, labels = trt_labels, name = NULL)

ggsave(here("Revision", "Output", "supp_fig3_physiology_vs_pH.png"),
       fig_supp3, width = 12, height = 9, dpi = 300)

fig_supp3
```

---

## Supp Fig 4 — Physiology vs mean DOC

```{r fig-supp4}
#| label: fig-supp4
#| fig-cap: "Coral physiology as a function of mean daytime DOC (tank means, n = 16). Points colored by treatment. No significant linear relationships were detected. Y-axes for Chl-a, Endosymbiont Density, and Tissue Biomass are on a log₁₀ scale."
#| fig-width: 12
#| fig-height: 9

doc_xlab <- expression("DOC (" * mu * "mol L"^{-1} * ")")

sp4a <- make_scatter("grand_mean_DOC", "mean_chla",    doc_xlab, chla_ylab, log_y = TRUE)
sp4b <- make_scatter("grand_mean_DOC", "mean_endos",   doc_xlab, endo_ylab, log_y = TRUE)
sp4c <- make_scatter("grand_mean_DOC", "mean_biomass", doc_xlab, bio_ylab,  log_y = TRUE)
sp4d <- make_scatter("grand_mean_DOC", "mean_R",       doc_xlab, R_ylab)
sp4e <- make_scatter("grand_mean_DOC", "mean_GP",      doc_xlab, GP_ylab)

fig_supp4 <- (sp4a + sp4b + sp4c) / (sp4d + sp4e) +
  plot_annotation(tag_levels = "a") +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom") &
  scale_color_manual(values = trt_colors, labels = trt_labels, name = NULL)

ggsave(here("Revision", "Output", "supp_fig4_physiology_vs_DOC.png"),
       fig_supp4, width = 12, height = 9, dpi = 300)

fig_supp4
```

---

# Model fit comparison {#sec-fit-comparison}

Global fit statistics for all three SEMs, assembled for cross-model comparison.
χ² statistics for small samples (n = 16 tanks) are unreliable in isolation;
the p(χ²), CFI, RMSEA, SRMR, and information criteria together inform adequacy.

```{r all-model-fits}
#| label: all-model-fits

library(kableExtra)

all_models <- list(
  "Main model (Chl/cm²)"        = tank_v6_fit,
  "Sensitivity: excl. 53 & 114" = tank_v6_fit_excl,
  "CPE model (Chl/cell)"        = cpe_fit
)

fit_stats <- c("chisq", "df", "pvalue", "cfi", "tli",
               "rmsea", "rmsea.ci.lower", "rmsea.ci.upper",
               "srmr", "aic", "bic")

fit_comparison <- do.call(rbind, lapply(names(all_models), \(nm) {
  fm <- fitMeasures(all_models[[nm]], fit_stats)
  setNames(
    data.frame(matrix(c(nm, round(fm, 3)), nrow = 1)),
    c("Model", fit_stats)
  )
}))

# Rename columns for display
names(fit_comparison) <- c(
  "Model", "\u03c7\u00b2", "df", "p(\u03c7\u00b2)",
  "CFI", "TLI", "RMSEA", "RMSEA\u00a0LB", "RMSEA\u00a0UB",
  "SRMR", "AIC", "BIC"
)

fit_comparison |>
  kbl(
    caption = paste0(
      "Global fit statistics across all three SEMs. ",
      "Acceptable thresholds (at this n): p(\u03c7\u00b2) > 0.05, CFI \u2265 0.90, ",
      "RMSEA \u2264 0.08, SRMR \u2264 0.08. ",
      "All models fall below conventional cut-offs for CFI/RMSEA; ",
      "the non-significant or marginal \u03c7\u00b2 and low \u03c7\u00b2/df ratio remain ",
      "the more informative indicators at n = 16. ",
      "Lower AIC/BIC indicates relatively better fit."
    ),
    align = c("l", rep("r", 11))
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE,
    position   = "left"
  ) |>
  row_spec(0, bold = TRUE) |>
  column_spec(1, bold = TRUE)
```

::: {.callout-note}
**Interpreting fit at n = 16:** With only 16 observations the χ²-based
indices (CFI, TLI, RMSEA) are known to be highly variable and biased
toward rejection even for well-specified models. The χ²/df ratio
(Main model: 25.1/16 = 1.57; CPE: 31.6/17 = 1.86) and the non-significant
or marginal p(χ²) are the more reliable guides. The main Chl/cm² model
outperforms the CPE model on every metric, including AIC (292.99 vs 294.11),
suggesting per-area pigment is a better-fitting mediator of the community
signal than per-cell pigment under these conditions.
:::

---

*Document generated: `r Sys.time()`*
