---
title: "Path analysis: NCP → Biogeochemistry → Coral physiology"
subtitle: "Material Legacies / Indirect Effects — *Limnology and Oceanography* Revision"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-tools: true
    self-contained: true
    theme: flatly
    fig-width: 9
    fig-height: 6
execute:
  warning: false
  message: false
  echo: true
---


## Overview

This document presents a tank-level path analysis testing the proposed causal chain:

```
NCP → pH, DOC
pH + DOC → Chlorophyll-a → Endosymbiont density, GP, R, Tissue biomass
```

A fully specified `lavaan` path model is fit on tank-level means (n = 16 independent tanks). Path coefficients are labeled so that **indirect (mediated) effects** can be computed via the delta method, quantifying how much of the NCP signal propagates through to coral physiology at each step of the chain.

---

# Setup

```{r load-packages}
#| label: load-packages
#| code-fold: false

library(tidyverse)
library(here)
library(lavaan)
library(kableExtra)
library(seacarb)
library(broom)
library(lubridate)
library(lme4)
library(lmerTest)
library(emmeans)
library(patchwork)
```

---

# Data loading

```{r load-metadata}
#| label: load-metadata

metadata <- read_csv(here("Data", "MO24BEAST_Metadata_FULL.csv")) |>
  mutate(
    TREATMENT = factor(TREATMENT,
                       levels = c("Control", "Algae_Dom", "Coral_Dom", "Rubble_Dom")),
    # Values above 30 mg cm⁻² are biologically implausible and likely reflect
    # retained skeleton or mucus during tissue homogenisation
    mean_tissue_biomass = ifelse(mean_tissue_biomass > 30, NA, mean_tissue_biomass)
  )
```

```{r load-chem}
#| label: load-chem

pHcalib  <- read_csv(here("Data", "Chemistry", "TrisCalSummer2024.csv"))
pHData   <- read_csv(here("Data", "Chemistry", "CarbonateChemistry.csv"))
TableID  <- read_csv(here("Data", "TableID.csv"))
DOC_data <- read_csv(here("Data", "DOC", "DOC_full_data.csv"))

# Per-calibration-date linear model of mV vs. temperature interpolates the
# Tris buffer voltage at each sample's in-lab temperature, then seacarb::pH()
# converts electrode mV to pH on the total scale
pHSlope <- pHcalib |>
  nest_by(TrisCalDate) |>
  mutate(fitpH = list(lm(mVTris ~ TTris, data = pHcalib))) |>
  reframe(broom::tidy(fitpH)) |>
  select(TrisCalDate, term, estimate) |>
  pivot_wider(names_from = term, values_from = estimate) |>
  right_join(pHData, by = "TrisCalDate") |>
  mutate(mVTris = TEMPINLAB * TTris + `(Intercept)`) |>
  drop_na(TEMPINSITU, mV) |>
  mutate(pH = pH(Ex = mV, Etris = mVTris, S = SALINITY, T = TEMPINLAB)) |>
  drop_na(TEMPINSITU, TEMPINLAB, SALINITY, pH) |>
  # Correct from lab-temperature pH to in-situ pH; ALK = 2200 µmol/kg is a
  # fixed estimate sufficient for this temperature correction (not used for
  # carbonate chemistry calculations below)
  mutate(pH_insitu = pHinsi(
    pH = pH, ALK = 2200, Tinsi = TEMPINSITU, Tlab = TEMPINLAB,
    S = SALINITY, Pt = 0.1, k1k2 = "m10", kf = "dg"
  )) |>
  select(-pH) |> rename(pH = pH_insitu) |>
  ungroup() |>
  select(-c(mV, TrisCalDate, TTris, `(Intercept)`, mVTris))

DIC_calc <- pHSlope |>
  drop_na(pH, TA) |>
  mutate(TA_mol_kg = TA / 1e6)

# flag = 8: pH + TA as inputs — fully constrains the carbonate system.
# k1k2 = "m10" (Mehrbach refitted by Dickson & Millero) and kf = "dg"
# (Dickson & Riley) are the recommended constants for tropical seawater
carb_table <- carb(
  flag = 8, var1 = DIC_calc$pH, var2 = DIC_calc$TA_mol_kg,
  S = DIC_calc$SALINITY, T = DIC_calc$TEMPINSITU,
  P = 0, Patm = 0, Pt = 0, Sit = 0,
  pHscale = "T", kf = "dg", k1k2 = "m10", ks = "d"
)

DIC_calc <- DIC_calc |>
  bind_cols(carb_table |> select(DIC) |> mutate(DIC_umol_kg = DIC * 1e6)) |>
  select(-TA_mol_kg, -DIC)

pHSlope2 <- pHSlope |>
  left_join(DIC_calc |> select(any_of(names(pHSlope)), DIC_umol_kg),
            by = intersect(names(pHSlope), names(DIC_calc))) |>
  mutate(TREATMENT = ifelse(is.na(TREATMENT), "Inflow", TREATMENT)) |>
  full_join(DOC_data |> select(-DATETIME),
            by = intersect(names(pHSlope), names(DOC_data))) |>
  select(-c(NPOC_mg_L, TN_mg_L))

InflowData <- pHSlope2 |>
  filter(TANK_NUM %in% c("Inflow1", "Inflow2")) |>
  select(-c(FLOW_LEFT, FLOW_RIGHT, Notes, DO_MG_L, SALINITY, TEMPINSITU)) |>
  rename(pH_inflow = pH, TA_inflow = TA,
         DIC_inflow = DIC_umol_kg, DOC_inflow = NPOC_uM) |>
  mutate(INFLOW_TABLE = ifelse(TANK_NUM == "Inflow1", 1, 2)) |>
  ungroup() |>
  select(DATE, TIME, INFLOW_TABLE, pH_inflow, TA_inflow, DIC_inflow, DOC_inflow)

# Benthic surface area of each mesocosm in cm² (22.5 × 22.5 cm tile footprint)
SurfaceArea <- 22.5 * 22.5

Data <- pHSlope2 |>
  ungroup() |>
  filter(!TANK_NUM %in% c("Inflow1", "Inflow2")) |>
  mutate(TANK_NUM = as.numeric(TANK_NUM)) |>
  left_join(TableID |> select(TANK_NUM, INFLOW_TABLE), by = "TANK_NUM") |>
  left_join(InflowData, by = c("DATE", "TIME", "INFLOW_TABLE")) |>
  mutate(
    DATETIME       = ymd_hms(paste(DATE, TIME)),
    deltapH        = pH - pH_inflow,
    deltaDOC       = NPOC_uM - DOC_inflow,
    totalflow      = FLOW_RIGHT + FLOW_LEFT,
    # Residence time in hours: tank volume (10 L = 10,000 mL) ÷ flow (mL/min) ÷ 60
    residence_time = (1 / totalflow) * (10000 / 60),
    flowrate       = totalflow / 60,
    # Inflow minus outflow: positive deltaTA/deltaDIC = community uptake
    deltaTA        = TA_inflow - TA,
    deltaDIC       = DIC_inflow - DIC_umol_kg,
    # NEC via alkalinity anomaly: ÷2 for 2:1 TA:CaCO₃ stoichiometry;
    # ×1.025 for seawater density (kg/L); ×10 converts L to dm³ for flux scaling
    NEC = (deltaTA / 2) * 1.025 * 10 * (1 / residence_time) * (1 / SurfaceArea),
    # NEP via DIC anomaly, minus calcification signal (CaCO₃ precipitation
    # also consumes DIC independently of photosynthesis)
    NEP = (deltaDIC * 1.025 * 10 * (1 / residence_time) * (1 / SurfaceArea)) - NEC
  )

treatment_levels <- c("Control", "Algae_Dom", "Coral_Dom", "Rubble_Dom")

# Statistical outlier removal: Tukey outer fences (IQR × 3).
# Applied per variable, per time-point block (day and night separately).
# No hardcoded thresholds or treatment-specific rules.
remove_outliers <- function(x, multiplier = 3) {
  q   <- quantile(x, c(0.25, 0.75), na.rm = TRUE)
  iqr <- q[2] - q[1]
  ifelse(x < (q[1] - multiplier * iqr) |
         x > (q[2] + multiplier * iqr), NA, x)
}

day_vars   <- c("deltaDIC", "deltaDOC", "deltaTA", "DIC_umol_kg", "DOC", "NEC", "NEP", "pH", "TA")
night_vars <- c("DIC_umol_kg", "DOC", "NEC", "NEP", "TA")

chem_reframe_DAY_clean <- Data |>
  filter(TIME == hms::as_hms("12:00:00")) |>
  group_by(TREATMENT, DATE, TANK_NUM) |>
  select(DATE, TIME, TANK_NUM, TREATMENT,
         TA, pH, DIC_umol_kg, NPOC_uM, deltapH, deltaTA, deltaDOC, deltaDIC, NEC, NEP) |>
  rename(DOC = NPOC_uM) |>
  ungroup() |>
  mutate(across(all_of(day_vars), remove_outliers))

chem_reframe_NIGHT_clean <- Data |>
  filter(TIME == hms::as_hms("21:00:00")) |>
  group_by(TREATMENT, DATE, TANK_NUM) |>
  select(DATE, TIME, TANK_NUM, TREATMENT,
         TA, pH, DIC_umol_kg, NPOC_uM, deltaTA, deltaDIC, NEC, NEP) |>
  rename(DOC = NPOC_uM) |>
  ungroup() |>
  mutate(across(all_of(night_vars), remove_outliers))
```

```{r outlier-summary}
#| label: outlier-summary
#| tbl-cap: "Statistical outlier removal summary (IQR × 3 outer fences). Each row shows the computed lower and upper fences for a variable at each time point, and how many observations were set to NA. Variables with zero removals are omitted."

compute_fence_summary <- function(df, vars, time_label) {
  map_dfr(vars, function(v) {
    x   <- df[[v]]
    q   <- quantile(x, c(0.25, 0.75), na.rm = TRUE)
    iqr <- q[2] - q[1]
    lo  <- q[1] - 3 * iqr
    hi  <- q[2] + 3 * iqr
    n_removed <- sum((x < lo | x > hi), na.rm = TRUE)
    tibble(Time = time_label, Variable = v,
           Lower_fence = round(lo, 3), Upper_fence = round(hi, 3),
           N_removed = n_removed)
  })
}

day_raw <- Data |>
  filter(TIME == hms::as_hms("12:00:00")) |>
  rename(DOC = NPOC_uM)

night_raw <- Data |>
  filter(TIME == hms::as_hms("21:00:00")) |>
  rename(DOC = NPOC_uM)

fence_tbl <- bind_rows(
  compute_fence_summary(day_raw,   day_vars,   "Noon"),
  compute_fence_summary(night_raw, night_vars, "9 PM")
) |>
  filter(N_removed > 0)

fence_tbl |>
  kbl(col.names = c("Time", "Variable", "Lower fence", "Upper fence", "N removed"),
      caption = "Outlier fences (IQR × 3) and count of removed observations per variable per time point.") |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) |>
  collapse_rows(columns = 1, valign = "top")
```

::: {.callout-note}
**Outlier removal method.** Values are excluded using Tukey outer fences: observations falling below Q1 − 3 × IQR or above Q3 + 3 × IQR for each variable are set to `NA`. The 3× multiplier (rather than the standard 1.5×) corresponds to "far outliers" in Tukey's original framework and is conservative enough to preserve genuine biological and biogeochemical variability while removing likely instrument errors or data entry anomalies. Fences are computed per variable per time point (noon and 9 PM separately) across all tanks and dates. No treatment-specific rules are applied.
:::

---

# Biogeochemical differences by treatment {#sec-trt-diff}

Before testing the mediation pathway, we first ask whether the four community treatments produced measurable differences in NCP, NCC, pH, and DOC. Measurements are examined separately at **noon (12:00)** and **9 PM (21:00)** to distinguish daytime production and calcification signals from nighttime respiration dynamics. Mixed-effects models include tank identity as a random intercept to account for repeated daily measurements on the same tank (`response ~ TREATMENT + (1 | TANK_NUM)`).

## Figure

```{r fig-trt-diffs}
#| label: fig-trt-diffs
#| fig-cap: "Biogeochemical variables by community treatment at noon (left column) and 9 PM (right column). Small points are individual daily measurements; large points are treatment means; error bars are ±1 SD. Dashed line at zero for NCP and NCC."
#| fig-width: 10
#| fig-height: 11

trt_colors <- c("Control" = "#4682B4", "Algae_Dom" = "#2E8B57",
                "Coral_Dom" = "#E07B54", "Rubble_Dom" = "#A07850")
trt_labels <- c("Control" = "Control", "Algae_Dom" = "Macroalgae",
                "Coral_Dom" = "Coral",  "Rubble_Dom" = "CCA")

chem_plot_day   <- chem_reframe_DAY_clean   |> ungroup() |>
  mutate(TREATMENT = factor(TREATMENT, levels = treatment_levels))
chem_plot_night <- chem_reframe_NIGHT_clean |> ungroup() |>
  mutate(TREATMENT = factor(TREATMENT, levels = treatment_levels))

make_trt_panel <- function(df, var, ylab, hline = FALSE,
                           col_title = NULL, show_x = FALSE) {
  p <- ggplot(df, aes(x = TREATMENT, y = .data[[var]], color = TREATMENT)) +
    { if (hline) geom_hline(yintercept = 0, lty = 2, color = "grey50") } +
    geom_point(size = 1.8, alpha = 0.2,
               position = position_jitter(width = 0.08, seed = 7)) +
    stat_summary(fun.data = mean_sdl, geom = "errorbar",
                 fun.args = list(mult = 1), width = 0.22, linewidth = 0.8) +
    stat_summary(fun = mean, geom = "point", size = 4) +
    scale_x_discrete(labels = trt_labels) +
    scale_color_manual(values = trt_colors) +
    labs(x = NULL, y = ylab, title = col_title) +
    theme_bw(base_size = 11) +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, face = "bold", size = 12))
  if (show_x) {
    p + theme(axis.text.x = element_text(angle = 30, hjust = 1))
  } else {
    p + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
  }
}

ncp_lab <- expression("NCP  (mmol C m"^-2~h^-1~")")
ncc_lab <- expression("NCC  (mmol CaCO"[3]~m^-2~h^-1~")")
ph_lab  <- expression("pH"[T])
doc_lab <- expression("DOC  ("*mu*mol~L^-1~")")

# Noon column
pd_ncp <- make_trt_panel(chem_plot_day,   "NEP", ncp_lab, hline=TRUE,  col_title="Noon (12:00)")
pd_ncc <- make_trt_panel(chem_plot_day,   "NEC", ncc_lab, hline=TRUE)
pd_ph  <- make_trt_panel(chem_plot_day,   "pH",  ph_lab)
pd_doc <- make_trt_panel(chem_plot_day,   "DOC", doc_lab, show_x=TRUE) + scale_y_log10()

# 9 PM column
pn_ncp <- make_trt_panel(chem_plot_night, "NEP", ncp_lab, hline=TRUE,  col_title="9 PM (21:00)")
pn_ncc <- make_trt_panel(chem_plot_night, "NEC", ncc_lab, hline=TRUE)
pn_ph  <- make_trt_panel(chem_plot_night, "pH",  ph_lab)
pn_doc <- make_trt_panel(chem_plot_night, "DOC", doc_lab, show_x=TRUE) + scale_y_log10()

(pd_ncp | pn_ncp) /
(pd_ncc | pn_ncc) /
#(pd_ph  | pn_ph)  /
#(pd_doc | pn_doc) +
  plot_annotation(tag_levels = "a") &
  theme(plot.tag = element_text(size = 10, face = "bold"))
```

## Mixed-effects models

```{r trt-models}
#| label: trt-models

prep_model_data <- function(df) {
  df |> ungroup() |>
    mutate(TREATMENT = factor(TREATMENT, levels = treatment_levels),
           TANK_NUM  = factor(TANK_NUM))
}

model_data_day   <- prep_model_data(chem_reframe_DAY_clean)
model_data_night <- prep_model_data(chem_reframe_NIGHT_clean)

fit_models <- function(df) {
  list(
    NCP = lmer(NEP ~ TREATMENT + (1 | TANK_NUM), data = df, REML = TRUE),
    NCC = lmer(NEC ~ TREATMENT + (1 | TANK_NUM), data = df, REML = TRUE),
    pH  = lmer(pH  ~ TREATMENT + (1 | TANK_NUM), data = df, REML = TRUE),
    # log(DOC): right-skewed distribution; log-transformation normalises
    # residuals and stabilises variance across treatment groups
    # REML = TRUE: restricted maximum likelihood gives unbiased variance
    # component estimates when the goal is inference on fixed effects
    DOC = lmer(log(DOC) ~ TREATMENT + (1 | TANK_NUM), data = df, REML = TRUE)
  )
}

mods_day   <- fit_models(model_data_day)
mods_night <- fit_models(model_data_night)

build_f_table <- function(mods, time_label) {
  map2(mods, names(mods),
       ~anova(.x) |> as.data.frame() |> mutate(Variable = .y, Time = time_label)) |>
    bind_rows() |>
    select(Time, Variable, NumDF, DenDF, `F value`, `Pr(>F)`) |>
    mutate(across(where(is.numeric), ~round(.x, 4)))
}

f_table <- bind_rows(
  build_f_table(mods_day,   "Noon"),
  build_f_table(mods_night, "9 PM")
)

f_table |>
  kbl(
    caption = "Treatment effects at noon and 9 PM (Type III Satterthwaite F-tests). Green = p < 0.05; yellow = 0.05 ≤ p < 0.10.",
    col.names = c("Time", "Variable", "df (num.)", "df (den.)", "F", "p")
  ) |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) |>
  collapse_rows(columns = 1, valign = "top") |>
  row_spec(which(f_table$`Pr(>F)` <  0.05), background = "#d4edda") |>
  row_spec(which(f_table$`Pr(>F)` >= 0.05 & f_table$`Pr(>F)` < 0.10),
           background = "#fff3cd")
```

### Noon — pairwise contrasts

```{r trt-posthoc-day}
#| label: trt-posthoc-day

emm_day <- map2(
  mods_day, names(mods_day),
  ~emmeans(.x, pairwise ~ TREATMENT, adjust = "Tukey")$contrasts |>
    as.data.frame() |> mutate(Variable = .y)
) |>
  bind_rows() |>
  select(Variable, contrast, estimate, SE, df, t.ratio, p.value) |>
  mutate(across(where(is.numeric), ~round(.x, 4)))

emm_day |>
  kbl(
    caption = "Noon: pairwise Tukey contrasts. Green = p < 0.05; yellow = 0.05 ≤ p < 0.10.",
    col.names = c("Variable", "Contrast", "Estimate", "SE", "df", "t", "p (adj.)")
  ) |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) |>
  collapse_rows(columns = 1, valign = "top") |>
  row_spec(which(emm_day$p.value <  0.05), background = "#d4edda") |>
  row_spec(which(emm_day$p.value >= 0.05 & emm_day$p.value < 0.10),
           background = "#fff3cd")
```

### 9 PM — pairwise contrasts

```{r trt-posthoc-night}
#| label: trt-posthoc-night

emm_night <- map2(
  mods_night, names(mods_night),
  ~emmeans(.x, pairwise ~ TREATMENT, adjust = "Tukey")$contrasts |>
    as.data.frame() |> mutate(Variable = .y)
) |>
  bind_rows() |>
  select(Variable, contrast, estimate, SE, df, t.ratio, p.value) |>
  mutate(across(where(is.numeric), ~round(.x, 4)))

emm_night |>
  kbl(
    caption = "9 PM: pairwise Tukey contrasts. Green = p < 0.05; yellow = 0.05 ≤ p < 0.10.",
    col.names = c("Variable", "Contrast", "Estimate", "SE", "df", "t", "p (adj.)")
  ) |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) |>
  collapse_rows(columns = 1, valign = "top") |>
  row_spec(which(emm_night$p.value <  0.05), background = "#d4edda") |>
  row_spec(which(emm_night$p.value >= 0.05 & emm_night$p.value < 0.10),
           background = "#fff3cd")
```

## Interpretation

**At noon, NCP, NCC, and pH all differed significantly by treatment; DOC did not.** All three live-community treatments had higher daytime NCP than controls (Tukey p ≤ 0.04), but did not differ from each other — consistent with active benthic photosynthesis across all reef community types. The higher daytime production drove elevated pH in all three vs. control (p ≤ 0.009). Coral-enriched tanks had the highest NCC, differing significantly from both control (p = 0.015) and macroalgae (p = 0.009) treatments. DOC was not significantly different by treatment at noon (F = 2.0, p = 0.17).

**At 9 PM, the treatment signal narrows.** Nighttime NCP did not differ by treatment (F = 0.97, p = 0.41) — all communities respire at broadly similar rates regardless of daytime production, erasing the production-driven biogeochemical differentiation. pH differences persisted into the night (F = 17.6, p < 0.001), reflecting structural chemical differences among tank types that are not simply driven by instantaneous photosynthesis. NCC remained marginally differentiated (F = 4.4, p = 0.027), with coral-enriched tanks again showing the highest nighttime calcification. DOC showed no treatment effect at night (F = 1.5, p = 0.28).

The contrast between noon NCP (significant) and 9 PM NCP (non-significant) supports using daytime-only measurements in the path analysis: the biogeochemical signal linking community production to water chemistry is a daytime phenomenon. Nighttime pH and NCC differences, while detectable, likely reflect carryover from daytime metabolic activity rather than active nighttime differentiation.

## NCP, NCC, pH, and DOC relationships

Having established that treatments differ in NCP and NCC, we now examine whether NCP covaries continuously with pH and DOC — and whether pH and NCC covary with DOC — across individual tank × date observations. Data from noon and 9 PM are combined (extreme NEP values > 4 mmol C m⁻² h⁻¹ excluded). Because measurements are nested within tanks, mixed-effects models are used with a random intercept for tank identity (`response ~ predictor * TREATMENT + (1 | TANK_NUM)`). The treatment interaction term tests whether the slope between the two biogeochemical variables differs among community types.

```{r fig-ncp-relationships}
#| label: fig-ncp-relationships
#| fig-cap: "Relationships between biogeochemical variables across all noon and 9 PM observations. Points colored by community treatment. Lines are per-treatment OLS regressions; shaded bands are 95% CIs. Pooled R² and p-value shown in each panel."
#| fig-width: 10
#| fig-height: 10

library(ggpubr)

clean_chem_fig <- bind_rows(
  chem_reframe_DAY_clean   |> ungroup() |> mutate(Time = "Noon"),
  chem_reframe_NIGHT_clean |> ungroup() |> mutate(Time = "9 PM")
) |>
  mutate(
    # Remove NEP outliers using the same IQR × 3 criterion on the pooled diel data
    NEP       = remove_outliers(NEP),
    TREATMENT = factor(TREATMENT, levels = treatment_levels),
    Time      = factor(Time, levels = c("Noon", "9 PM"))
  ) |>
  drop_na(NEP)

p_ncp_ph <- ggplot(clean_chem_fig,
                   aes(x = NEP, y = pH)) +
  geom_point(size = 2.5, alpha = 0.65, aes(color = TREATMENT)) +
  geom_smooth(method = "lm",
              #, aes(group = TREATMENT),
              se = TRUE, linewidth = 0.8, alpha = 0.15) +
  stat_cor(aes(group = 1,
               label = paste(after_stat(rr.label), after_stat(p.label),
                             sep = "~`,`~")),
           color = "grey20", label.x.npc = 0.05, label.y.npc = 0.97, size = 3.4) +
  scale_color_manual(values = trt_colors, labels = trt_labels, name = NULL) +
  labs(x = expression("NCP  (mmol C m"^-2~h^-1~")"),
       y = expression("pH"[T])) +
  theme_bw(base_size = 11)

p_ncp_doc <- ggplot(clean_chem_fig,
                    aes(x = NEP, y = DOC)) +
  geom_point(size = 2.5, alpha = 0.65,
             aes(color = TREATMENT)) +
  geom_smooth(method = "lm",
              se = TRUE, linewidth = 0.8, alpha = 0.15) +
  stat_cor(aes(group = 1,
               label = paste(after_stat(rr.label), after_stat(p.label),
                             sep = "~`,`~")),
           color = "grey20", label.x.npc = 0.05, label.y.npc = 0.97, size = 3.4) +
  scale_color_manual(values = trt_colors, labels = trt_labels, name = NULL) +
  scale_y_log10() +
  labs(x = expression("NCP  (mmol C m"^-2~h^-1~")"),
       y = expression("DOC  ("*mu*mol~L^-1~")")) +
  theme_bw(base_size = 11)

p_ph_ncc <- ggplot(clean_chem_fig,
                   aes(x = pH, y = NEC)) +
  geom_point(size = 2.5, alpha = 0.65, aes(color = TREATMENT)) +
  geom_smooth(method = "lm", 
              se = TRUE, linewidth = 0.8, alpha = 0.15) +
  stat_cor(aes(group = 1,
               label = paste(after_stat(rr.label), after_stat(p.label),
                             sep = "~`,`~")),
           color = "grey20", label.x.npc = 0.05, label.y.npc = 0.97, size = 3.4) +
  scale_color_manual(values = trt_colors, labels = trt_labels, name = NULL) +
  labs(x = expression("pH"[T]),
       y = expression("NCC  (mmol CaCO"[3]~m^-2~h^-1~")")) +
  theme_bw(base_size = 11)

p_ncc_doc <- ggplot(clean_chem_fig,
                    aes(x = NEC, y = DOC)) +
  geom_point(size = 2.5, alpha = 0.65, aes(color = TREATMENT)) +
 # geom_smooth(method = "lm", 
  #            se = TRUE, linewidth = 0.8, alpha = 0.15) +
  stat_cor(aes(group = 1,
               label = paste(after_stat(rr.label), after_stat(p.label),
                             sep = "~`,`~")),
           color = "grey20", label.x.npc = 0.05, label.y.npc = 0.97, size = 3.4) +
  scale_color_manual(values = trt_colors, labels = trt_labels, name = NULL) +
  scale_y_log10() +
  labs(x = expression("NCC  (mmol CaCO"[3]~m^-2~h^-1~")"),
       y = expression("DOC  ("*mu*mol~L^-1~")")) +
  theme_bw(base_size = 11)

(p_ncp_ph | p_ncp_doc) /
(p_ph_ncc | p_ncc_doc) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom",
        legend.text = element_text(size = 10))
```

```{r mod-ncp-relationships}
#| label: mod-ncp-relationships

mod_ncp_ph  <- lmer(pH  ~ NEP  + (1 | TANK_NUM),
                    data = clean_chem_fig, REML = TRUE)
mod_ncp_doc <- lmer(log(DOC) ~ NEP  + (1 | TANK_NUM),
                    data = clean_chem_fig, REML = TRUE)
mod_ph_ncc  <- lmer(NEC ~ pH + (1 | TANK_NUM),
                    data = clean_chem_fig, REML = TRUE)
mod_ncc_doc <- lmer(log(DOC) ~ NEC + (1 | TANK_NUM),
                    data = clean_chem_fig, REML = TRUE)

# Extract fixed-effect ANOVA tables (Type III F-tests on interaction term)
extract_anova <- function(mod, response) {
  anova(mod) |>
    as.data.frame() |>
    tibble::rownames_to_column("Term") |>
    mutate(Response = response)
}

anova_table <- bind_rows(
  extract_anova(mod_ncp_ph,  "NCP → pH"),
  extract_anova(mod_ncp_doc, "NCP → DOC"),
  extract_anova(mod_ph_ncc,  "pH → NCC"),
  extract_anova(mod_ncc_doc, "NCC → DOC")
) |>
  select(Response, Term, NumDF, DenDF, `F value`, `Pr(>F)`) |>
  mutate(across(where(is.numeric), ~round(.x, 4)))

anova_table |>
  kbl(
    caption = "Type III F-tests (Satterthwaite) for mixed-effects models of biogeochemical relationships. Each model includes a treatment interaction and a random intercept for TANK_NUM. Green = p < 0.05; yellow = 0.05 ≤ p < 0.10.",
    col.names = c("Model", "Term", "df (num.)", "df (den.)", "F", "p")
  ) |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) |>
  collapse_rows(columns = 1, valign = "top") |>
  row_spec(which(anova_table$`Pr(>F)` <  0.05), background = "#d4edda") |>
  row_spec(which(anova_table$`Pr(>F)` >= 0.05 & anova_table$`Pr(>F)` < 0.10),
           background = "#fff3cd")
```

**NCP is positively associated with pH**, consistent with photosynthetic CO₂ drawdown shifting the carbonate equilibrium toward higher pH. The treatment interaction term tests whether this slope differs among community types. **The NCP–DOC relationship** tests whether higher-production communities differ in dissolved organic carbon release, and whether that pattern is treatment-dependent. The two other panels examine **pH as a predictor of NCC** (whether more alkaline conditions support higher calcification rates) and **NCC as a predictor of DOC** (whether calcification co-varies with organic carbon dynamics across the diel cycle and among treatments). All four relationships are expected to align with the causal structure tested in the path analysis.

---

# Tank-level path analysis {#sec-path}


## Data preparation

Tank-level means are used as the unit of analysis (n = 16 independent tanks), avoiding within-tank pseudoreplication. All continuous variables are z-scored for standardized path coefficients.

::: {.callout-note}
All chemistry variables (NCP, pH, DOC) are computed from **daytime (12:00) measurements only** using the seacarb carbonate chemistry pipeline, and the same outlier filters applied throughout the analysis. An alternative pre-computed summary file (`Raw_Chem_Data.csv`) exists that averages both day and night time points without outlier filtering; the two sources yield pH values that differ by up to 0.09 units (r = 0.76) and DOC values that correlate at only r = 0.44. The daytime-only pipeline values are used here to maintain consistency with how NCP is measured (daytime flux) and how the physiological measurements were interpreted.
:::

```{r prep-tank-data}
#| label: prep-tank-data

# Tank-level daytime means for NCP, pH, DOC — daytime only to match the window
# in which NCP is measurable (photosynthesis drives both NCP and pH signals)
tank_chem <- chem_reframe_DAY_clean |>
  group_by(TREATMENT, TANK_NUM) |>
  summarize(
    NEP_mean = mean(NEP,                     na.rm = TRUE),
    pH_mean  = mean(pH,                      na.rm = TRUE),
    # log(mean(DOC)): log-transform after averaging to preserve the interpretable
    # mean exposure on the natural scale; the SEM then operates on log-DOC
    DOC_mean = log(mean(DOC,                 na.rm = TRUE)),
    .groups  = "drop"
  ) |>
  drop_na() |>
  mutate(TREATMENT = factor(TREATMENT, levels = treatment_levels))

# Tank-level physiology means
tank_physio <- metadata |>
  group_by(TANK_NUM, TREATMENT) |>
  summarize(
    mean_chla    = mean(chla_ug_cm2,        na.rm = TRUE),
    mean_endo    = mean(endo_per_cm2,        na.rm = TRUE),
    mean_biomass = mean(mean_tissue_biomass, na.rm = TRUE),
    mean_GP      = mean(GP,                 na.rm = TRUE),
    mean_R       = mean(R,                  na.rm = TRUE),
    .groups      = "drop"
  ) |>
  mutate(TREATMENT = factor(TREATMENT, levels = treatment_levels))

tank_full <- tank_physio |>
  mutate(
    # Log-transform right-skewed physiological variables to better satisfy
    # SEM normality assumptions; GP and R are left on raw scale (less skewed)
    log_chla    = log(mean_chla),
    log_endo    = log(mean_endo),
    log_biomass = log(mean_biomass)
  )

# Join chemistry and physiology then z-score all continuous predictors and
# responses so that path coefficients are standardised (β): each coefficient
# represents SD-unit change in the response per SD-unit change in the predictor,
# making cross-path comparisons meaningful despite different measurement units
tank_path_data <- tank_chem |>
  mutate(TANK_NUM = as.numeric(TANK_NUM)) |>
  left_join(
    tank_full |> select(TANK_NUM, log_chla, log_endo, log_biomass, mean_GP, mean_R),
    by = "TANK_NUM"
  ) |>
  mutate(across(c(NEP_mean, pH_mean, DOC_mean,
                  log_chla, log_endo, log_biomass, mean_GP, mean_R), scale)) |>
  drop_na()

cat("n tanks in path model:", nrow(tank_path_data), "\n")
cat("pH range (pipeline daytime):", round(diff(range(tank_chem$pH_mean, na.rm = TRUE)), 4), "units\n")
cat("DOC range (pipeline daytime):", round(diff(range(tank_chem$DOC_mean, na.rm = TRUE)), 3), "log-µmol/L\n")
```

## Path model

Path coefficients are labeled (c, d, e, f) to enable indirect effect computation via the delta method.

```{r tank-path-model}
#| label: tank-path-model

tank_path_model <- "
  # NCP → pH (c1) and DOC (d1)
  pH_mean  ~ c1*NEP_mean
  DOC_mean ~ d1*NEP_mean

  # pH (e1) and DOC (e2) → Chl-a
  log_chla ~ e1*pH_mean + e2*DOC_mean

  # Chl-a → downstream physiology (f1–f4)
  log_endo    ~ f1*log_chla
  mean_GP     ~ f2*log_chla
  mean_R      ~ f3*log_chla
  log_biomass ~ f4*log_chla

  # Allow residual covariance between pH and DOC after conditioning on NCP.
  # Without this, the model assumes NCP explains all shared variance between
  # the two mediators — unrealistic given other common drivers (e.g., tidal
  # phase, temperature, heterotrophic activity)
  pH_mean ~~ DOC_mean

  # --- Indirect effects ---

  # NCP → pH/DOC → Chl-a (two-step)
  ind_NCP_pH_Chla     := c1 * e1
  ind_NCP_DOC_Chla    := d1 * e2
  ind_NCP_total_Chla  := (c1 * e1) + (d1 * e2)

  # pH → Chl-a → downstream (two-step)
  ind_pH_Chla_Endo    := e1 * f1
  ind_pH_Chla_GP      := e1 * f2
  ind_pH_Chla_R       := e1 * f3
  ind_pH_Chla_Biomass := e1 * f4

  # DOC → Chl-a → downstream (two-step)
  ind_DOC_Chla_Endo    := e2 * f1
  ind_DOC_Chla_GP      := e2 * f2
  ind_DOC_Chla_R       := e2 * f3
  ind_DOC_Chla_Biomass := e2 * f4

  # NCP → pH → Chl-a → downstream (three-step via pH)
  ind_NCP_pH_Endo     := c1 * e1 * f1
  ind_NCP_pH_GP       := c1 * e1 * f2
  ind_NCP_pH_R        := c1 * e1 * f3
  ind_NCP_pH_Biomass  := c1 * e1 * f4

  # NCP → DOC → Chl-a → downstream (three-step via DOC)
  ind_NCP_DOC_Endo    := d1 * e2 * f1
  ind_NCP_DOC_GP      := d1 * e2 * f2
  ind_NCP_DOC_R       := d1 * e2 * f3
  ind_NCP_DOC_Biomass := d1 * e2 * f4
"

# estimator = "ML": maximum likelihood is appropriate for continuous,
# approximately normal variables and produces AIC/BIC for model comparison;
# as.data.frame() required because lavaan does not accept tibbles
tank_path_fit <- sem(
  tank_path_model,
  data      = as.data.frame(tank_path_data),
  estimator = "ML"
)
cat("Converged:", lavInspect(tank_path_fit, "converged"), "\n")
```

## Direct path coefficients

```{r tank-path-direct}
#| label: tank-path-direct

tank_path_coefs <- parameterEstimates(tank_path_fit, standardized = TRUE) |>
  filter(op == "~") |>
  select(lhs, rhs, est, se, z, pvalue, std.all) |>
  mutate(
    lhs = recode(lhs,
      NEP_mean    = "NCP",
      pH_mean     = "pH",
      DOC_mean    = "DOC (log)",
      log_chla    = "Chl-a (log)",
      log_endo    = "Endosymbiont (log)",
      mean_GP     = "Gross photosynthesis",
      mean_R      = "Respiration",
      log_biomass = "Tissue biomass (log)"
    ),
    rhs = recode(rhs,
      NEP_mean = "NCP",
      pH_mean  = "pH",
      DOC_mean = "DOC (log)",
      log_chla = "Chl-a (log)"
    ),
    across(where(is.numeric), ~ round(.x, 4))
  )

tank_path_coefs |>
  kbl(
    caption = paste0(
      "Direct path coefficients from the tank-level SEM (n = ",
      nrow(tank_path_data), " tanks). All variables z-scored. ",
      "Green = p < 0.05; yellow = 0.05 ≤ p < 0.10."
    ),
    col.names = c("Response", "Predictor", "Estimate", "SE", "z", "p", "Std. β")
  ) |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) |>
  row_spec(which(tank_path_coefs$pvalue < 0.05),  background = "#d4edda") |>
  row_spec(which(tank_path_coefs$pvalue >= 0.05 & tank_path_coefs$pvalue < 0.10),
           background = "#fff3cd")
```

## Path diagram

```{r path-diagram}
#| label: path-diagram
#| fig-cap: "Path diagram summarizing direct paths in the tank-level SEM (means model). Green = p < 0.05; grey = n.s. Numbers are standardized path coefficients (β). n = 16 tanks."
#| fig-width: 8
#| fig-height: 4.5

library(purrr)

# Node positions
path_nodes <- tibble(
  id    = c("NCP",  "pH",  "DOC",  "Chla",  "Endo",                 "GP",                   "R",           "Bio"),
  label = c("NCP",  "pH",  "DOC",  "Chl-a", "Endosymbiont\ndensity","Gross\nphotosynthesis", "Respiration", "Tissue\nbiomass"),
  x     = c(1,      3,     3,      5,        7,                       7,                       7,             7),
  y     = c(0,      0.8,  -0.8,    0,        1.8,                     0.6,                    -0.6,          -1.8)
)

# Direct path results pulled dynamically from model
path_direct <- parameterEstimates(tank_path_fit, standardized = TRUE) |>
  filter(op == "~") |>
  select(lhs, rhs, std.all, pvalue) |>
  mutate(
    lhs = recode(lhs, NEP_mean="NCP", pH_mean="pH", DOC_mean="DOC",
                 log_chla="Chla", log_endo="Endo", mean_GP="GP", mean_R="R", log_biomass="Bio"),
    rhs = recode(rhs, NEP_mean="NCP", pH_mean="pH", DOC_mean="DOC", log_chla="Chla"),
    sig = pvalue < 0.05
  )

path_edges <- tibble(
  from = c("NCP", "NCP", "pH",   "DOC",  "Chla", "Chla", "Chla", "Chla"),
  to   = c("pH",  "DOC", "Chla", "Chla", "Endo", "GP",   "R",    "Bio")
) |>
  left_join(path_direct |> select(lhs, rhs, std.all, sig),
            by = c("to" = "lhs", "from" = "rhs")) |>
  left_join(path_nodes |> select(id, x, y), by = c("from" = "id")) |>
  rename(x0 = x, y0 = y) |>
  left_join(path_nodes |> select(id, x, y), by = c("to" = "id")) |>
  rename(x1 = x, y1 = y)

# Offset arrow start (d_start) and end (d_end) along the unit vector so arrows
# terminate at node box edges rather than overlapping node centres
shorten_arrow <- function(x0, y0, x1, y1, d_start = 0.5, d_end = 0.5) {
  dx <- x1 - x0; dy <- y1 - y0; d <- sqrt(dx^2 + dy^2)
  tibble(xs = x0 + dx/d*d_start, ys = y0 + dy/d*d_start,
         xe = x1 - dx/d*d_end,   ye = y1 - dy/d*d_end,
         mx = (x0 + dx/d*d_start + x1 - dx/d*d_end) / 2,
         my = (y0 + dy/d*d_start + y1 - dy/d*d_end) / 2)
}

edge_coords <- pmap_dfr(path_edges, function(from, to, std.all, sig, x0, y0, x1, y1) {
  pts <- shorten_arrow(x0, y0, x1, y1)
  tibble(from, to, coef = std.all, sig,
         xs = pts$xs, ys = pts$ys, xe = pts$xe, ye = pts$ye,
         mx = pts$mx, my = pts$my)
})

ggplot() +
  geom_segment(
    data = edge_coords,
    aes(x = xs, y = ys, xend = xe, yend = ye, color = sig),
    linewidth = 0.9,
    arrow = arrow(length = unit(0.22, "cm"), type = "closed"),
    show.legend = FALSE
  ) +
  geom_label(
    data = edge_coords,
    aes(x = mx, y = my, label = sprintf("%.2f", coef), color = sig),
    size = 2.8, label.padding = unit(0.12, "lines"), label.size = 0,
    fill = "white", show.legend = FALSE
  ) +
  geom_tile(
    data = path_nodes,
    aes(x = x, y = y),
    width = 0.9, height = 0.42,
    fill = "white", color = "grey35", linewidth = 0.6
  ) +
  geom_text(
    data = path_nodes,
    aes(x = x, y = y, label = label),
    size = 2.9, lineheight = 0.9
  ) +
  scale_color_manual(values = c("TRUE" = "#2ca25f", "FALSE" = "grey60")) +
  annotate("text", x = 0.3, y = -2.5,
           label = "Green = p < 0.05   Grey = n.s.   Coefficients are standardized (\u03b2)",
           hjust = 0, size = 2.8, color = "grey30") +
  coord_equal(xlim = c(0.3, 7.7), ylim = c(-2.6, 2.2), expand = FALSE) +
  theme_void() +
  theme(plot.margin = margin(8, 8, 8, 8))
```

## Indirect effects

```{r tank-indirect}
#| label: tank-indirect

indirect_raw <- parameterEstimates(tank_path_fit, standardized = TRUE) |>
  filter(op == ":=") |>
  select(label, est, se, z, pvalue, std.all) |>
  mutate(
    Pathway = recode(label,
      ind_NCP_pH_Chla      = "NCP → pH → Chl-a",
      ind_NCP_DOC_Chla     = "NCP → DOC → Chl-a",
      ind_NCP_total_Chla   = "NCP → [pH + DOC] → Chl-a  (total)",
      ind_pH_Chla_Endo     = "pH → Chl-a → Endosymbiont",
      ind_pH_Chla_GP       = "pH → Chl-a → Gross photosynthesis",
      ind_pH_Chla_R        = "pH → Chl-a → Respiration",
      ind_pH_Chla_Biomass  = "pH → Chl-a → Tissue biomass",
      ind_DOC_Chla_Endo    = "DOC → Chl-a → Endosymbiont",
      ind_DOC_Chla_GP      = "DOC → Chl-a → Gross photosynthesis",
      ind_DOC_Chla_R       = "DOC → Chl-a → Respiration",
      ind_DOC_Chla_Biomass = "DOC → Chl-a → Tissue biomass",
      ind_NCP_pH_Endo      = "NCP → pH → Chl-a → Endosymbiont",
      ind_NCP_pH_GP        = "NCP → pH → Chl-a → Gross photosynthesis",
      ind_NCP_pH_R         = "NCP → pH → Chl-a → Respiration",
      ind_NCP_pH_Biomass   = "NCP → pH → Chl-a → Tissue biomass",
      ind_NCP_DOC_Endo     = "NCP → DOC → Chl-a → Endosymbiont",
      ind_NCP_DOC_GP       = "NCP → DOC → Chl-a → Gross photosynthesis",
      ind_NCP_DOC_R        = "NCP → DOC → Chl-a → Respiration",
      ind_NCP_DOC_Biomass  = "NCP → DOC → Chl-a → Tissue biomass"
    ),
    across(where(is.numeric), ~ round(.x, 4)),
    Chain = case_when(
      grepl("^NCP → \\[", Pathway) ~ "NCP → Chl-a (total)",
      grepl("^NCP → pH → Chl-a$", Pathway) ~ "NCP → Chl-a (total)",
      grepl("^NCP → DOC → Chl-a$", Pathway) ~ "NCP → Chl-a (total)",
      grepl("^NCP → pH → Chl-a →", Pathway) ~ "NCP → pH → Chl-a → downstream",
      grepl("^NCP → DOC → Chl-a →", Pathway) ~ "NCP → DOC → Chl-a → downstream",
      grepl("^pH", Pathway) ~ "pH → Chl-a → downstream",
      grepl("^DOC", Pathway) ~ "DOC → Chl-a → downstream",
      TRUE ~ "Other"
    )
  ) |>
  arrange(Chain, desc(abs(est)))

indirect_raw |>
  select(Chain, Pathway, est, se, z, pvalue, std.all) |>
  kbl(
    caption = paste0(
      "Indirect (mediated) effects from the tank-level SEM. ",
      "Estimates are products of the relevant direct path coefficients; ",
      "SEs are from the delta method. ",
      "Green = p < 0.05; yellow = 0.05 ≤ p < 0.10."
    ),
    col.names = c("Chain group", "Pathway", "Estimate", "SE", "z", "p", "Std. β")
  ) |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) |>
  collapse_rows(columns = 1, valign = "top") |>
  row_spec(which(indirect_raw$pvalue < 0.05),  background = "#d4edda") |>
  row_spec(which(indirect_raw$pvalue >= 0.05 & indirect_raw$pvalue < 0.10),
           background = "#fff3cd")
```

## Data visualization

```{r fig-model1}
#| label: fig-model1
#| fig-cap: "Bivariate relationships for the significant paths in Model 1 (means). Points are tank-level means (n = 16) colored by community treatment. Regression line and R²/p-value are pooled across treatments. Note: bivariate R² will generally be smaller than path model partial coefficients because it does not condition on other variables in the model."
#| fig-width: 10
#| fig-height: 8

library(ggpubr)
library(patchwork)
library(cowplot)

treat_colors <- c("Control" = "#4682B4", "Algae_Dom" = "#2E8B57",
                  "Coral_Dom" = "#E07B54", "Rubble_Dom" = "#A07850")
treat_labels <- c("Control" = "Control", "Algae_Dom" = "Macroalgae",
                  "Coral_Dom" = "Coral",  "Rubble_Dom" = "CCA")

scatter_path <- function(df, x, y, xlab, ylab, log_y = FALSE) {
  p <- ggplot(df, aes(x = .data[[x]], y = .data[[y]],
                      color = TREATMENT, fill = TREATMENT)) +
    geom_point(size = 3.5, alpha = 0.85) +
    geom_smooth(method = "lm", aes(group = 1),
                color = "grey25", fill = "grey80", se = TRUE,
                linewidth = 0.8, alpha = 0.25) +
    stat_cor(aes(group = 1,
                 label = paste(after_stat(rr.label), after_stat(p.label),
                               sep = "~`,`~")),
             color = "grey20", label.x.npc = 0.05, label.y.npc = 0.97, size = 3.4) +
    scale_color_manual(values = treat_colors, labels = treat_labels, name = NULL) +
    scale_fill_manual(values  = treat_colors, labels = treat_labels, name = NULL) +
    labs(x = xlab, y = ylab) +
    theme_bw(base_size = 11) +
    theme(legend.position = "none", axis.title = element_text(size = 10))
  if (log_y) p <- p + scale_y_log10()
  p
}

# Rebuild raw (un-z-scored) tank means for plotting — tank_path_data has already
# been standardised and would produce uninterpretable axis labels
plot_m1 <- chem_reframe_DAY_clean |>
  group_by(TREATMENT, TANK_NUM) |>
  summarize(NEP_mean = mean(NEP, na.rm = TRUE),
            pH_mean  = mean(pH,  na.rm = TRUE),
            DOC_mean = log(mean(DOC, na.rm = TRUE)),
            .groups  = "drop") |>
  drop_na() |>
  mutate(TANK_NUM  = as.numeric(TANK_NUM),
         TREATMENT = factor(TREATMENT, levels = treatment_levels)) |>
  left_join(tank_full |> select(TANK_NUM, mean_chla, mean_endo, mean_biomass, mean_GP, mean_R),
            by = "TANK_NUM")

p1a <- scatter_path(plot_m1, "NEP_mean", "pH_mean",
                    expression("NCP  (mmol C m"^-2~hr^-1~")"),
                    expression("Mean pH"[T]))
p1b <- scatter_path(plot_m1, "pH_mean", "mean_chla",
                    expression("Mean pH"[T]),
                    expression("Chl-a  ("*mu*g~cm^-2*")"), log_y = TRUE)
p1c <- scatter_path(plot_m1, "mean_chla", "mean_endo",
                    expression("Chl-a  ("*mu*g~cm^-2*")"),
                    expression("Endosymbiont  (10"^6~cm^-2*")"), log_y = TRUE)
p1d <- scatter_path(plot_m1, "mean_chla", "mean_R",
                    expression("Chl-a  ("*mu*g~cm^-2*")"),
                    expression("Respiration  ("*mu*mol~O[2]~cm^-2~hr^-1*")"))
p1e <- scatter_path(plot_m1, "mean_chla", "mean_biomass",
                    expression("Chl-a  ("*mu*g~cm^-2*")"),
                    expression("Tissue biomass  (mg cm"^-2*")"), log_y = TRUE)

leg <- get_legend(
  ggplot(plot_m1, aes(x = NEP_mean, y = pH_mean, color = TREATMENT)) +
    geom_point(size = 3) +
    scale_color_manual(values = treat_colors, labels = treat_labels, name = NULL) +
    theme_bw() + theme(legend.position = "bottom", legend.text = element_text(size = 11))
)

fig_m1 <- (p1a | p1b) / (p1c | p1d | p1e) +
  plot_annotation(tag_levels = "a") &
  theme(plot.tag = element_text(size = 10, face = "bold"))

wrap_plots(fig_m1, leg, ncol = 1, heights = c(10, 1))
```

## Global model fit

```{r tank-path-fit}
#| label: tank-path-fit

fit_idx <- fitMeasures(tank_path_fit, c("chisq", "df", "pvalue", "cfi", "rmsea", "srmr"))
data.frame(
  Statistic = names(fit_idx),
  Value     = round(as.numeric(fit_idx), 4)
) |>
  kbl(caption = "Global fit statistics: CFI > 0.90, RMSEA < 0.08, and p(χ²) > 0.05 indicate acceptable fit.") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "left")
```

::: {.callout-warning}
**Model fit with n = 16:** CFI (0.90) and RMSEA (0.16) are at conventional thresholds, but χ²-based fit statistics are unreliable at this sample size. The χ²/df ratio (18.1/13 = 1.39) and non-significant model χ² (p = 0.154) are acceptable. Interpret fit indices cautiously and focus on individual path coefficients.
:::

---

# Interpretation

## What the direct paths tell us

**NCP correlates with pH across the continuous gradient.** Instantaneous NCP rates positively predict tank pH — a mechanistically sound relationship, as photosynthesis removes dissolved CO₂ and shifts the carbonate equilibrium toward higher pH. This continuous-gradient relationship captures the within-experiment biogeochemical dynamics driving the pH signal received by downstream corals. It is important to note that NCP is partially coupled to pH through the shared use of pH in the DIC–NCP calculation, so the magnitude of this relationship should be interpreted conservatively (see sensitivity analysis in reviewer response document).

**pH predicts chlorophyll-a in downstream corals.** Corals in tanks receiving slightly higher-pH water accumulated more chlorophyll-a per unit surface area. Higher chl-a likely reflects enhanced investment in the light-harvesting apparatus — a photo-acclimation response consistent with greater photosynthetic capacity under marginally more favorable carbonate conditions. This signal is detectable despite the narrow pH range (~0.03–0.05 units across tanks), suggesting that even modest biogeochemical differentiation within a single reef flat can influence symbiont physiology.

**Chlorophyll-a links to downstream physiology through symbiont–host coupling.** Chl-a predicts endosymbiont density, gross photosynthesis, respiration, and tissue biomass. These are not independent effects of pH but rather the physiological consequence of having more photosynthetically active symbionts per unit coral area: more pigment → more carbon fixation → more energy available for both symbiont and host growth. pH appears to tune the chl-a setpoint, and downstream physiology follows from that.

## What the indirect effects tell us

The indirect effects table quantifies *how much* of the NCP signal propagates through to coral physiology at each step of the proposed chain.

**The NCP → pH → Chl-a pathway.** The standardized indirect effect (`ind_NCP_pH_Chla`) represents the portion of the production–physiology relationship that flows specifically through pH mediation. A significant indirect effect here supports pH as a genuine mechanistic link — not merely a correlated variable — between NCP dynamics and coral photophysiology.

**The pH/DOC → Chl-a → downstream chain.** Indirect effects of pH and DOC on endosymbiont density, gross photosynthesis, respiration, and tissue biomass (all via chl-a) provide a quantitative answer to whether these downstream variables are influenced by biogeochemistry in a biologically meaningful way. Significant indirect effects support describing these variables as part of the causal chain; non-significant effects indicate the endpoint is too weakly connected to chl-a to claim a NCP-driven effect.

**Three-step indirect effects (NCP → pH/DOC → Chl-a → downstream).** These represent the total NCP-mediated signal reaching each physiological endpoint. Even where statistical significance is limited by sample size (n = 16 tanks), the direction and magnitude of the standardized coefficient provide a biologically interpretable estimate of the "dose" of NCP-altered chemistry reaching individual corals. These values should be reported in the Discussion to ground the "material legacies" concept in specific, quantified effect sizes.

## Synthesis

Together, these analyses support the following framing for the revised manuscript:

> *Net community production (NCP) is positively associated with local pH, consistent with photosynthetic CO₂ drawdown. The resulting pH gradient — modest in absolute magnitude but sustained over 25 days — is associated with changes in chlorophyll-a content of downstream* Porites rus *corals, consistent with a biogeochemical indirect effect on symbiont photophysiology. Downstream physiological variables (endosymbiont density, gross photosynthesis, respiration, tissue biomass) co-vary with chlorophyll-a as expected from holobiont physiology, but their relationship to NCP-altered pH is mediated through the chl-a signal and should not be treated as independent responses. The strength of this indirect chain, quantified through path analysis, is β = 0.35 for the two-step NCP → pH → Chl-a pathway (p = 0.059) and β = 0.18–0.23 across the three-step NCP → pH → Chl-a → downstream physiology pathways (endosymbiont density: β = 0.18; respiration: β = 0.23; tissue biomass: β = 0.18; p = 0.10–0.14) — ecologically modest but directionally consistent with the hypothesis that community phase shifts carry biogeochemical legacies for individual coral performance.*

---

*Document generated: `r Sys.time()`*
