---
title: "Path analysis: NCP → Biogeochemistry → Coral physiology"
subtitle: "Material Legacies / Indirect Effects — *Limnology and Oceanography* Revision"
date: today
bibliography: references.bib
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-tools: true
    self-contained: true
    theme: flatly
    fig-width: 9
    fig-height: 6
execute:
  warning: false
  message: false
  echo: true
---


## Overview

This document presents a fragment-level piecewiseSEM path analysis testing the proposed causal chain from community metabolism to coral physiology:

```
NCP → pH, DOC  (tank-level, exogenous predictors)
pH + DOC → Chl-a
Chl-a → Endosymbiont density
Endosymbiont → Tissue biomass → Respiration → GP
Endosymbiont → GP
```

Two parallel models are fitted: one using **Chl-a per cm²** and one using **Chl-a per endosymbiont cell** (Chl/cell) as the pigment variable. All path equations are `lmer` models with a genotype random intercept `(1 | GENOTYPE)`, wrapped in a `piecewiseSEM::psem()` object. Global model fit is assessed via Shipley's Fisher's C d-separation statistic.

---

# Setup

```{r load-packages}
#| label: load-packages
#| code-fold: false

library(tidyverse)
library(here)
library(kableExtra)
library(seacarb)
library(broom)
library(lubridate)
library(lme4)
library(lmerTest)
library(emmeans)
library(multcomp)
library(multcompView)
library(patchwork)
library(cowplot)
library(piecewiseSEM)
library(ggpubr)
library(purrr)
```

---

# Data loading

```{r load-metadata}
#| label: load-metadata

metadata <- read_csv(here("Data", "MO24BEAST_Metadata_FULL.csv")) |>
  mutate(
    TREATMENT = factor(TREATMENT,
                       levels = c("Control", "Algae_Dom", "Coral_Dom", "Rubble_Dom")),
    # Values above 30 mg cm⁻² are biologically implausible and likely reflect
    # retained skeleton or mucus during tissue homogenisation
    mean_tissue_biomass = ifelse(mean_tissue_biomass > 30, NA, mean_tissue_biomass)
  )
```

```{r load-chem}
#| label: load-chem

pHcalib  <- read_csv(here("Data", "Chemistry", "TrisCalSummer2024.csv"))
pHData   <- read_csv(here("Data", "Chemistry", "CarbonateChemistry.csv"))
TableID  <- read_csv(here("Data", "TableID.csv"))
DOC_data <- read_csv(here("Data", "DOC", "DOC_full_data.csv"))

# Per-calibration-date linear model of mV vs. temperature interpolates the
# Tris buffer voltage at each sample's in-lab temperature, then seacarb::pH()
# converts electrode mV to pH on the total scale
pHSlope <- pHcalib |>
  nest_by(TrisCalDate) |>
  mutate(fitpH = list(lm(mVTris ~ TTris, data = pHcalib))) |>
  reframe(broom::tidy(fitpH)) |>
  dplyr::select(TrisCalDate, term, estimate) |>
  pivot_wider(names_from = term, values_from = estimate) |>
  right_join(pHData, by = "TrisCalDate") |>
  mutate(mVTris = TEMPINLAB * TTris + `(Intercept)`) |>
  drop_na(TEMPINSITU, mV) |>
  mutate(pH = pH(Ex = mV, Etris = mVTris, S = SALINITY, T = TEMPINLAB)) |>
  drop_na(TEMPINSITU, TEMPINLAB, SALINITY, pH) |>
  # Correct from lab-temperature pH to in-situ pH; ALK = 2200 µmol/kg is a
  # fixed estimate sufficient for this temperature correction (not used for
  # carbonate chemistry calculations below)
  mutate(pH_insitu = pHinsi(
    pH = pH, ALK = 2200, Tinsi = TEMPINSITU, Tlab = TEMPINLAB,
    S = SALINITY, Pt = 0.1, k1k2 = "m10", kf = "dg"
  )) |>
  dplyr::select(-pH) |> rename(pH = pH_insitu) |>
  ungroup() |>
  dplyr::select(-c(mV, TrisCalDate, TTris, `(Intercept)`, mVTris))

DIC_calc <- pHSlope |>
  drop_na(pH, TA) |>
  mutate(TA_mol_kg = TA / 1e6)

# flag = 8: pH + TA as inputs — fully constrains the carbonate system.
# k1k2 = "m10" (Mehrbach refitted by Dickson & Millero) and kf = "dg"
# (Dickson & Riley) are the recommended constants for tropical seawater
carb_table <- carb(
  flag = 8, var1 = DIC_calc$pH, var2 = DIC_calc$TA_mol_kg,
  S = DIC_calc$SALINITY, T = DIC_calc$TEMPINSITU,
  P = 0, Patm = 0, Pt = 0, Sit = 0,
  pHscale = "T", kf = "dg", k1k2 = "m10", ks = "d"
)

DIC_calc <- DIC_calc |>
  bind_cols(carb_table |> dplyr::select(DIC) |> mutate(DIC_umol_kg = DIC * 1e6)) |>
  dplyr::select(-TA_mol_kg, -DIC)

pHSlope2 <- pHSlope |>
  left_join(DIC_calc |> dplyr::select(any_of(names(pHSlope)), DIC_umol_kg),
            by = intersect(names(pHSlope), names(DIC_calc))) |>
  mutate(TREATMENT = ifelse(is.na(TREATMENT), "Inflow", TREATMENT)) |>
  full_join(DOC_data |> dplyr::select(-DATETIME),
            by = intersect(names(pHSlope), names(DOC_data))) |>
  dplyr::select(-c(NPOC_mg_L, TN_mg_L))

InflowData <- pHSlope2 |>
  filter(TANK_NUM %in% c("Inflow1", "Inflow2")) |>
  dplyr::select(-c(FLOW_LEFT, FLOW_RIGHT, Notes, DO_MG_L, SALINITY, TEMPINSITU)) |>
  rename(pH_inflow = pH, TA_inflow = TA,
         DIC_inflow = DIC_umol_kg, DOC_inflow = NPOC_uM) |>
  mutate(INFLOW_TABLE = ifelse(TANK_NUM == "Inflow1", 1, 2)) |>
  ungroup() |>
  dplyr::select(DATE, TIME, INFLOW_TABLE, pH_inflow, TA_inflow, DIC_inflow, DOC_inflow)

# Benthic surface area of each mesocosm in cm² (22.5 × 22.5 cm tile footprint)
SurfaceArea <- 22.5 * 22.5

Data <- pHSlope2 |>
  ungroup() |>
  filter(!TANK_NUM %in% c("Inflow1", "Inflow2")) |>
  mutate(TANK_NUM = as.numeric(TANK_NUM)) |>
  left_join(TableID |> dplyr::select(TANK_NUM, INFLOW_TABLE), by = "TANK_NUM") |>
  left_join(InflowData, by = c("DATE", "TIME", "INFLOW_TABLE")) |>
  mutate(
    DATETIME       = ymd_hms(paste(DATE, TIME)),
    deltapH        = pH - pH_inflow,
    deltaDOC       = NPOC_uM - DOC_inflow,
    totalflow      = FLOW_RIGHT + FLOW_LEFT,
    # Residence time in hours: tank volume (10 L = 10,000 mL) ÷ flow (mL/min) ÷ 60
    residence_time = (1 / totalflow) * (10000 / 60),
    flowrate       = totalflow / 60,
    # Inflow minus outflow: positive deltaTA/deltaDIC = community uptake
    deltaTA        = TA_inflow - TA,
    deltaDIC       = DIC_inflow - DIC_umol_kg,
    # NEC via alkalinity anomaly: ÷2 for 2:1 TA:CaCO₃ stoichiometry;
    # ×1.025 for seawater density (kg/L); ×10 converts L to dm³ for flux scaling
    NEC = (deltaTA / 2) * 1.025 * 10 * (1 / residence_time) * (1 / SurfaceArea),
    # NEP via DIC anomaly, minus calcification signal (CaCO₃ precipitation
    # also consumes DIC independently of photosynthesis)
    NEP = (deltaDIC * 1.025 * 10 * (1 / residence_time) * (1 / SurfaceArea)) - NEC
  )

treatment_levels <- c("Control", "Algae_Dom", "Coral_Dom", "Rubble_Dom")

# Statistical outlier removal: Tukey outer fences (IQR × 3).
# Applied per variable, per time-point block (day and night separately).
# No hardcoded thresholds or treatment-specific rules.
remove_outliers <- function(x, multiplier = 3) {
  q   <- quantile(x, c(0.25, 0.75), na.rm = TRUE)
  iqr <- q[2] - q[1]
  ifelse(x < (q[1] - multiplier * iqr) |
         x > (q[2] + multiplier * iqr), NA, x)
}

day_vars   <- c("DOC", "NEC", "NEP", "pH", "TA")
night_vars <- c("DOC", "NEC", "NEP", "TA")

chem_reframe_DAY_clean <- Data |>
  filter(TIME == hms::as_hms("12:00:00")) |>
  group_by(TREATMENT, DATE, TANK_NUM) |>
  dplyr::select(DATE, TIME, TANK_NUM, TREATMENT,
         TA, pH, DIC_umol_kg, NPOC_uM, deltapH, deltaTA, deltaDOC, deltaDIC, NEC, NEP) |>
  rename(DOC = NPOC_uM) |>
  ungroup() |>
  mutate(across(all_of(day_vars), remove_outliers))

chem_reframe_NIGHT_clean <- Data |>
  filter(TIME == hms::as_hms("21:00:00")) |>
  group_by(TREATMENT, DATE, TANK_NUM) |>
  dplyr::select(DATE, TIME, TANK_NUM, TREATMENT,
         TA, pH, DIC_umol_kg, NPOC_uM, deltaTA, deltaDIC, NEC, NEP) |>
  rename(DOC = NPOC_uM) |>
  ungroup() |>
  mutate(across(all_of(night_vars), remove_outliers))
```

```{r outlier-summary}
#| label: outlier-summary
#| tbl-cap: "Statistical outlier removal summary (IQR × 3 outer fences). Each row shows the computed lower and upper fences for a variable at each time point, and how many observations were set to NA. Variables with zero removals are omitted."

compute_fence_summary <- function(df, vars, time_label) {
  map_dfr(vars, function(v) {
    x   <- df[[v]]
    q   <- quantile(x, c(0.25, 0.75), na.rm = TRUE)
    iqr <- q[2] - q[1]
    lo  <- q[1] - 3 * iqr
    hi  <- q[2] + 3 * iqr
    n_removed <- sum((x < lo | x > hi), na.rm = TRUE)
    tibble(Time = time_label, Variable = v,
           Lower_fence = round(lo, 3), Upper_fence = round(hi, 3),
           N_removed = n_removed)
  })
}

day_raw <- Data |>
  filter(TIME == hms::as_hms("12:00:00")) |>
  rename(DOC = NPOC_uM)

night_raw <- Data |>
  filter(TIME == hms::as_hms("21:00:00")) |>
  rename(DOC = NPOC_uM)

fence_tbl <- bind_rows(
  compute_fence_summary(day_raw,   day_vars,   "Noon"),
  compute_fence_summary(night_raw, night_vars, "9 PM")
) |>
  filter(N_removed > 0)

fence_tbl |>
  kbl(col.names = c("Time", "Variable", "Lower fence", "Upper fence", "N removed"),
      caption = "Outlier fences (IQR × 3) and count of removed observations per variable per time point.") |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) |>
  collapse_rows(columns = 1, valign = "top")
```

::: {.callout-note}
**Outlier removal method.** Values are excluded using Tukey outer fences: observations falling below Q1 − 3 × IQR or above Q3 + 3 × IQR for each variable are set to `NA`. The 3× multiplier (rather than the standard 1.5×) corresponds to "far outliers" in Tukey's original framework and is conservative enough to preserve genuine biological and biogeochemical variability while removing likely instrument errors or data entry anomalies. Fences are computed per variable per time point (noon and 9 PM separately) across all tanks and dates. No treatment-specific rules are applied.
:::

---

# Moorea backreef benthic community {#sec-moorea}

```{r load-percent-cover}
#| label: load-percent-cover

pc_data <- read_csv(here("Data", "PercentCoverData", "backreef_benthic_comp_mean_perc.csv"),
                    show_col_types = FALSE) |>
  mutate(
    benthic_group = factor(benthic_group,
      levels = c("coral", "ctb", "CCA", "macroalgae", "sand"),
      labels = c("Coral", "Rubble", "Crustose Coralline Algae", "Macroalgae", "Sand")),
    year = factor(year)
  )
```

## Figure

```{r fig-moorea-cover}
#| label: fig-moorea-cover
#| fig-cap: "Benthic community composition at the LTER1 Moorea backreef across three survey years. Stacked bars show mean percent cover for each benthic category. Colors match the community treatments used in this experiment: coral (orange), CCA/rubble (brown), and macroalgae (green)."
#| fig-width: 5
#| fig-height: 5

benthic_colors <- c(
  "Coral"                    = "#E07B54",
  "Crustose Coralline Algae" = "#A07850",
  "Macroalgae"               = "#2E8B57",
  "Rubble"                   = "#C8B89A",
  "Sand"                     = "#D6C8A8"
)

fig_moorea_cover <- ggplot(pc_data, aes(x = year, y = mean_percent_cover, fill = benthic_group)) +
  geom_bar(position = "fill", stat = "identity", color = "white", linewidth = 0.4) +
  geom_text(aes(label = ifelse(mean_percent_cover >= 3,
                               paste0(round(mean_percent_cover, 0), "%"), "")),
            position = position_fill(vjust = 0.5),
            color = "white", size = 3.2, fontface = "bold") +
  scale_fill_manual(values = benthic_colors,
                    guide  = guide_legend(reverse = TRUE)) +
  scale_y_continuous(labels = scales::percent_format(), expand = c(0, 0)) +
  labs(x = "Year", y = "Percent cover", fill = "Benthic category") +
  theme_bw(base_size = 11) +
  theme(
    legend.position    = "right",
    legend.title       = element_text(size = 10, face = "bold"),
    legend.text        = element_text(size = 9),
    axis.title         = element_text(size = 10),
    axis.text          = element_text(size = 9),
    panel.grid.major.x = element_blank()
  )

ggsave(here("Revision", "Output", "fig0_moorea_percent_cover.png"),
       fig_moorea_cover, width = 5, height = 5, dpi = 300)

fig_moorea_cover
```

---

# Biogeochemical differences by treatment {#sec-trt-diff}

Before testing the mediation pathway, we first ask whether the four community treatments produced measurable differences in NCP, NCC, pH, and DOC. Measurements are examined separately at **noon (12:00)** and **9 PM (21:00)** to distinguish daytime production and calcification signals from nighttime respiration dynamics. Mixed-effects models include tank identity as a random intercept to account for repeated daily measurements on the same tank (`response ~ TREATMENT + (1 | TANK_NUM)`).

## Figure

```{r fig-trt-diffs}
#| label: fig-trt-diffs
#| fig-cap: "Biogeochemical variables by community treatment at noon (left column) and 9 PM (right column). Points are individual tank means (n = 4 per treatment); crossbar shows the treatment mean. Compact letter display from Tukey HSD (p < 0.05); groups sharing a letter do not differ significantly. Dashed line at zero for NCP and NCC."
#| fig-width: 8
#| fig-height: 11

trt_colors <- c("Control" = "#4682B4", "Algae_Dom" = "#2E8B57",
                "Coral_Dom" = "#E07B54", "Rubble_Dom" = "#A07850")
trt_labels <- c("Control" = "Control", "Algae_Dom" = "Macroalgae",
                "Coral_Dom" = "Coral",  "Rubble_Dom" = "CCA")

# ── Tank-level means ──────────────────────────────────────────────────────────
# pH and DOC: from co-measured observations only (complete pairs)
complete_day_fig <- chem_reframe_DAY_clean |> ungroup() |>
  filter(!is.na(pH), !is.na(DOC)) |>
  mutate(TREATMENT = factor(TREATMENT, levels = treatment_levels),
         TANK_NUM  = as.numeric(TANK_NUM))

complete_night_fig <- chem_reframe_NIGHT_clean |> ungroup() |>
  filter(!is.na(pH), !is.na(DOC)) |>
  mutate(TREATMENT = factor(TREATMENT, levels = treatment_levels),
         TANK_NUM  = as.numeric(TANK_NUM))

tank_avg_fig <- bind_rows(
  complete_day_fig   |> mutate(Time = "Noon"),
  complete_night_fig |> mutate(Time = "9 PM")
) |>
  group_by(TREATMENT, TANK_NUM, Time) |>
  summarize(pH_mean  = mean(pH,  na.rm = TRUE),
            DOC_mean = mean(DOC, na.rm = TRUE), .groups = "drop") |>
  mutate(TANK_NUM = as.numeric(as.character(TANK_NUM)))

# NCP and NCC: full filtered datasets
tank_ncp_ncc_fig <- bind_rows(
  chem_reframe_DAY_clean   |> ungroup() |> mutate(Time = "Noon"),
  chem_reframe_NIGHT_clean |> ungroup() |> mutate(Time = "9 PM")
) |>
  group_by(TREATMENT, TANK_NUM, Time) |>
  summarize(NCP_mean = mean(NEP, na.rm = TRUE),
            NCC_mean = mean(NEC, na.rm = TRUE), .groups = "drop") |>
  mutate(TREATMENT = factor(TREATMENT, levels = treatment_levels))

tank_all_fig <- tank_avg_fig |>
  left_join(tank_ncp_ncc_fig, by = c("TREATMENT", "TANK_NUM", "Time")) |>
  mutate(Time = factor(Time, levels = c("Noon", "9 PM")))

# ── Raw daily observations for violin backgrounds ──────────────────────────────
raw_ph_doc <- bind_rows(
  complete_day_fig   |> mutate(Time = "Noon"),
  complete_night_fig |> mutate(Time = "9 PM")
) |>
  rename(pH_mean = pH, DOC_mean = DOC) |>
  dplyr::select(TREATMENT, TANK_NUM, DATE, Time, pH_mean, DOC_mean)

raw_ncp_ncc <- bind_rows(
  chem_reframe_DAY_clean   |> ungroup() |> mutate(Time = "Noon"),
  chem_reframe_NIGHT_clean |> ungroup() |> mutate(Time = "9 PM")
) |>
  rename(NCP_mean = NEP, NCC_mean = NEC) |>
  dplyr::select(TREATMENT, TANK_NUM, DATE, Time, NCP_mean, NCC_mean) |>
  mutate(TANK_NUM = as.numeric(as.character(TANK_NUM)))

raw_all_fig <- raw_ph_doc |>
  full_join(raw_ncp_ncc, by = c("TREATMENT", "TANK_NUM", "DATE", "Time"),
            relationship = "many-to-many") |>
  distinct(TREATMENT, TANK_NUM, DATE, Time, .keep_all = TRUE) |>
  mutate(
    TREATMENT = factor(TREATMENT, levels = treatment_levels),
    Time      = factor(Time, levels = c("Noon", "9 PM"))
  )

# ── Tukey letters ─────────────────────────────────────────────────────────────
get_letters_fig <- function(var, time_val, data, offset, raw_data = raw_all_fig) {
  df   <- data     |> filter(Time == time_val) |> drop_na(all_of(var))
  rdf  <- raw_data |> filter(Time == time_val) |> drop_na(all_of(var))
  mod  <- lmer(reformulate(c("TREATMENT", "(1|TANK_NUM)"), response = var), data = rdf)
  em   <- emmeans(mod, ~ TREATMENT)
  cld_df <- cld(em, Letters = letters, adjust = "tukey") |>
    mutate(.group = trimws(.group))
  tibble(
    TREATMENT = factor(cld_df$TREATMENT, levels = treatment_levels),
    letter    = cld_df$.group,
    y_pos     = max(rdf[[var]], na.rm = TRUE) + offset,
    Variable  = var,
    Time      = factor(time_val, levels = c("Noon", "9 PM"))
  )
}

offsets <- list(NCP_mean=0.45, NCC_mean=0.12, pH_mean=0.012, DOC_mean=28)
letters_fig <- map2_dfr(
  names(offsets), offsets,
  ~bind_rows(
    get_letters_fig(.x, "Noon", tank_all_fig, .y),
    get_letters_fig(.x, "9 PM", tank_all_fig, .y)
  )
)

# ── Panel function ─────────────────────────────────────────────────────────────
make_tank_panel_fig <- function(var, time_val, ylab, hline = FALSE,
                                show_x = FALSE, col_title = NULL) {
  df  <- tank_all_fig  |> filter(Time == time_val) |> drop_na(all_of(var))
  rdf <- raw_all_fig   |> filter(Time == time_val) |> drop_na(all_of(var))
  ldf <- letters_fig   |> filter(Variable == var,  Time == time_val)
  p <- ggplot(df, aes(x = TREATMENT, y = .data[[var]], color = TREATMENT)) +
    { if (hline) geom_hline(yintercept = 0, lty = 2, color = "grey55",
                             linewidth = 0.4) } +
    geom_violin(data = rdf,
                aes(fill = TREATMENT),
                alpha = 0.12, color = NA, trim = FALSE) +
    geom_point(size = 3.2, alpha = 0.85) +
    stat_summary(fun = mean, geom = "crossbar",
                 width = 0.42, linewidth = 0.65) +
    geom_text(data = ldf,
              aes(x = TREATMENT, y = y_pos, label = letter),
              color = "grey20", size = 3.2, fontface = "bold",
              inherit.aes = FALSE) +
    scale_x_discrete(labels = trt_labels) +
    scale_color_manual(values = trt_colors) +
    scale_fill_manual(values = trt_colors) +
    labs(x = NULL, y = ylab, title = col_title) +
    theme_bw(base_size = 10) +
    theme(legend.position = "none",
          plot.title      = element_text(hjust = 0.5, face = "bold", size = 11),
          axis.title.y    = element_text(size = 9))
  if (show_x) p + theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 8))
  else        p + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
}

ncp_lab <- expression("NCP  (mmol C m"^{-2}~h^{-1}*")")
ncc_lab <- expression("NCC  (mmol CaCO"[3]~m^{-2}~h^{-1}*")")
ph_lab  <- expression("Mean pH"[T])
doc_lab <- expression("Mean DOC  ("*mu*mol~L^{-1}*")")

# Also keep make_trt_panel and per-variable plots available for downstream code
chem_plot_day   <- chem_reframe_DAY_clean   |> ungroup() |>
  mutate(TREATMENT = factor(TREATMENT, levels = treatment_levels))
chem_plot_night <- chem_reframe_NIGHT_clean |> ungroup() |>
  mutate(TREATMENT = factor(TREATMENT, levels = treatment_levels))

make_trt_panel <- function(df, var, ylab, hline = FALSE,
                           col_title = NULL, show_x = FALSE) {
  p <- ggplot(df, aes(x = TREATMENT, y = .data[[var]], color = TREATMENT)) +
    { if (hline) geom_hline(yintercept = 0, lty = 2, color = "grey50") } +
    geom_point(size = 1.8, alpha = 0.2,
               position = position_jitter(width = 0.08, seed = 7)) +
    stat_summary(fun.data = mean_sdl, geom = "errorbar",
                 fun.args = list(mult = 1), width = 0.22, linewidth = 0.8) +
    stat_summary(fun = mean, geom = "point", size = 4) +
    scale_x_discrete(labels = trt_labels) +
    scale_color_manual(values = trt_colors) +
    labs(x = NULL, y = ylab, title = col_title) +
    theme_bw(base_size = 11) +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, face = "bold", size = 12))
  if (show_x) p + theme(axis.text.x = element_text(angle = 30, hjust = 1))
  else        p + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
}

ph_lab_orig  <- expression("pH"[T])
doc_lab_orig <- expression("DOC  ("*mu*mol~L^-1~")")

pd_ncp <- make_trt_panel(chem_plot_day,   "NEP", ncp_lab, hline=TRUE, col_title="Noon (12:00)")
pd_ncc <- make_trt_panel(chem_plot_day,   "NEC", ncc_lab, hline=TRUE)
pd_ph  <- make_trt_panel(chem_plot_day,   "pH",  ph_lab_orig)
pd_doc <- make_trt_panel(chem_plot_day,   "DOC", doc_lab_orig, show_x=TRUE) + scale_y_log10()
pn_ncp <- make_trt_panel(chem_plot_night, "NEP", ncp_lab, hline=TRUE, col_title="9 PM (21:00)")
pn_ncc <- make_trt_panel(chem_plot_night, "NEC", ncc_lab, hline=TRUE)
pn_ph  <- make_trt_panel(chem_plot_night, "pH",  ph_lab_orig)
pn_doc <- make_trt_panel(chem_plot_night, "DOC", doc_lab_orig, show_x=TRUE) + scale_y_log10()

# ── Primary figure: tank-level means with Tukey letters ──────────────────────
fig_trt <-
  (make_tank_panel_fig("NCP_mean","Noon",ncp_lab,hline=TRUE,col_title="Noon (12:00)") |
   make_tank_panel_fig("NCP_mean","9 PM",ncp_lab,hline=TRUE,col_title="9 PM (21:00)")) /
  (make_tank_panel_fig("NCC_mean","Noon",ncc_lab,hline=TRUE) |
   make_tank_panel_fig("NCC_mean","9 PM",ncc_lab,hline=TRUE)) /
  (make_tank_panel_fig("pH_mean", "Noon",ph_lab)  |
   make_tank_panel_fig("pH_mean", "9 PM",ph_lab)) /
  (make_tank_panel_fig("DOC_mean","Noon",doc_lab,show_x=TRUE) |
   make_tank_panel_fig("DOC_mean","9 PM",doc_lab,show_x=TRUE)) +
  plot_annotation(
    tag_levels = "a",
    caption    = paste0(
      "Violin = distribution of all daily observations  |  ",
      "Points = individual tank means (n = 4 per treatment)  |  ",
      "Crossbar = treatment mean  |  ",
      "Letters = Tukey HSD (p < 0.05)"
    )
  ) &
  theme(plot.tag     = element_text(size = 9, face = "bold"),
        plot.caption = element_text(size = 7.5, color = "grey40", hjust = 0))

ggsave(here("Revision", "Output", "fig1_treatment_differences.png"),
       fig_trt, width = 8, height = 11, dpi = 300)

fig_trt
```

## Mixed-effects models

```{r trt-models}
#| label: trt-models

prep_model_data <- function(df) {
  df |> ungroup() |>
    mutate(TREATMENT = factor(TREATMENT, levels = treatment_levels),
           TANK_NUM  = factor(TANK_NUM))
}

model_data_day   <- prep_model_data(chem_reframe_DAY_clean)
model_data_night <- prep_model_data(chem_reframe_NIGHT_clean)

fit_models <- function(df) {
  list(
    NCP = lmer(NEP ~ TREATMENT + (1 | TANK_NUM), data = df, REML = TRUE),
    NCC = lmer(NEC ~ TREATMENT + (1 | TANK_NUM), data = df, REML = TRUE),
    pH  = lmer(pH  ~ TREATMENT + (1 | TANK_NUM), data = df, REML = TRUE),
    # log(DOC): right-skewed distribution; log-transformation normalises
    # residuals and stabilises variance across treatment groups
    # REML = TRUE: restricted maximum likelihood gives unbiased variance
    # component estimates when the goal is inference on fixed effects
    DOC = lmer(log(DOC) ~ TREATMENT + (1 | TANK_NUM), data = df, REML = TRUE)
  )
}

mods_day   <- fit_models(model_data_day)
mods_night <- fit_models(model_data_night)

build_f_table <- function(mods, time_label) {
  map2(mods, names(mods),
       ~anova(.x) |> as.data.frame() |> mutate(Variable = .y, Time = time_label)) |>
    bind_rows() |>
    dplyr::select(Time, Variable, NumDF, DenDF, `F value`, `Pr(>F)`) |>
    mutate(across(where(is.numeric), ~round(.x, 4)))
}

f_table <- bind_rows(
  build_f_table(mods_day,   "Noon"),
  build_f_table(mods_night, "9 PM")
)

f_table |>
  kbl(
    caption = "Treatment effects at noon and 9 PM (Type III Satterthwaite F-tests). Green = p < 0.05; yellow = 0.05 ≤ p < 0.10.",
    col.names = c("Time", "Variable", "df (num.)", "df (den.)", "F", "p")
  ) |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) |>
  collapse_rows(columns = 1, valign = "top") |>
  row_spec(which(f_table$`Pr(>F)` <  0.05), background = "#d4edda") |>
  row_spec(which(f_table$`Pr(>F)` >= 0.05 & f_table$`Pr(>F)` < 0.10),
           background = "#fff3cd")
```

### Noon — pairwise contrasts

```{r trt-posthoc-day}
#| label: trt-posthoc-day

emm_day <- map2(
  mods_day, names(mods_day),
  ~emmeans(.x, pairwise ~ TREATMENT, adjust = "Tukey")$contrasts |>
    as.data.frame() |> mutate(Variable = .y)
) |>
  bind_rows() |>
  dplyr::select(Variable, contrast, estimate, SE, df, t.ratio, p.value) |>
  mutate(across(where(is.numeric), ~round(.x, 4)))

emm_day |>
  kbl(
    caption = "Noon: pairwise Tukey contrasts. Green = p < 0.05; yellow = 0.05 ≤ p < 0.10.",
    col.names = c("Variable", "Contrast", "Estimate", "SE", "df", "t", "p (adj.)")
  ) |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) |>
  collapse_rows(columns = 1, valign = "top") |>
  row_spec(which(emm_day$p.value <  0.05), background = "#d4edda") |>
  row_spec(which(emm_day$p.value >= 0.05 & emm_day$p.value < 0.10),
           background = "#fff3cd")
```

### 9 PM — pairwise contrasts

```{r trt-posthoc-night}
#| label: trt-posthoc-night

emm_night <- map2(
  mods_night, names(mods_night),
  ~emmeans(.x, pairwise ~ TREATMENT, adjust = "Tukey")$contrasts |>
    as.data.frame() |> mutate(Variable = .y)
) |>
  bind_rows() |>
  dplyr::select(Variable, contrast, estimate, SE, df, t.ratio, p.value) |>
  mutate(across(where(is.numeric), ~round(.x, 4)))

emm_night |>
  kbl(
    caption = "9 PM: pairwise Tukey contrasts. Green = p < 0.05; yellow = 0.05 ≤ p < 0.10.",
    col.names = c("Variable", "Contrast", "Estimate", "SE", "df", "t", "p (adj.)")
  ) |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) |>
  collapse_rows(columns = 1, valign = "top") |>
  row_spec(which(emm_night$p.value <  0.05), background = "#d4edda") |>
  row_spec(which(emm_night$p.value >= 0.05 & emm_night$p.value < 0.10),
           background = "#fff3cd")
```

## Interpretation

**At noon, NCP, NCC, and pH all differed significantly by treatment; DOC did not.** All three live-community treatments had higher daytime NCP than controls (Tukey p ≤ 0.04), but did not differ from each other — consistent with active benthic photosynthesis across all reef community types. The higher daytime production drove elevated pH in all three vs. control (p ≤ 0.009). Coral-enriched tanks had the highest NCC, differing significantly from both control (p = 0.015) and macroalgae (p = 0.009) treatments. DOC was not significantly different by treatment at noon (F = 1.6, p = 0.20).

**At 9 PM, the treatment signal narrows.** Nighttime NCP did not differ by treatment (F = 0.97, p = 0.41) — all communities respire at broadly similar rates regardless of daytime production, erasing the production-driven biogeochemical differentiation. pH differences persisted into the night (F = 17.6, p < 0.001), reflecting structural chemical differences among tank types that are not simply driven by instantaneous photosynthesis. NCC remained marginally differentiated (F = 4.4, p = 0.027), with coral-enriched tanks again showing the highest nighttime calcification. DOC showed no treatment effect at night (F = 1.4, p = 0.29).

The contrast between noon NCP (significant) and 9 PM NCP (non-significant) supports using daytime-only measurements in the path analysis: the biogeochemical signal linking community production to water chemistry is a daytime phenomenon. Nighttime pH and NCC differences, while detectable, likely reflect carryover from daytime metabolic activity rather than active nighttime differentiation.

## NCP, NCC, pH, and DOC relationships

Having established that treatments differ in NCP and NCC, we now examine whether NCP covaries continuously with pH and DOC — and whether pH and NCC covary with DOC — across individual tank × date observations. Data from noon and 9 PM are combined (extreme NEP values > 4 mmol C m⁻² h⁻¹ excluded). Because measurements are nested within tanks, mixed-effects models are used with a random intercept for tank identity (`response ~ predictor * TREATMENT + (1 | TANK_NUM)`). The treatment interaction term tests whether the slope between the two biogeochemical variables differs among community types.

```{r fig-ncp-relationships}
#| label: fig-ncp-relationships
#| fig-cap: "Relationships between biogeochemical variables across all noon and 9 PM observations. Points colored by community treatment. Lines are per-treatment OLS regressions; shaded bands are 95% CIs. Pooled R² and p-value shown in each panel."
#| fig-width: 10
#| fig-height: 10

clean_chem_fig <- bind_rows(
  chem_reframe_DAY_clean   |> ungroup() |> mutate(Time = "Noon"),
  chem_reframe_NIGHT_clean |> ungroup() |> mutate(Time = "9 PM")
) |>
  mutate(
    # Remove NEP outliers using the same IQR × 3 criterion on the pooled diel data
    NEP       = remove_outliers(NEP),
    TREATMENT = factor(TREATMENT, levels = treatment_levels),
    Time      = factor(Time, levels = c("Noon", "9 PM"))
  ) |>
  drop_na(NEP)

p_ncp_ph <- ggplot(clean_chem_fig,
                   aes(x = NEP, y = pH)) +
  geom_point(size = 2.5, alpha = 0.65, aes(color = TREATMENT)) +
  geom_smooth(method = "lm",
              #, aes(group = TREATMENT),
              se = TRUE, linewidth = 0.8, alpha = 0.15) +
  stat_cor(aes(group = 1,
               label = paste(after_stat(rr.label), after_stat(p.label),
                             sep = "~`,`~")),
           color = "grey20", label.x.npc = 0.05, label.y.npc = 0.97, size = 3.4) +
  scale_color_manual(values = trt_colors, labels = trt_labels, name = NULL) +
  labs(x = expression("NCP  (mmol C m"^-2~h^-1~")"),
       y = expression("pH"[T])) +
  theme_bw(base_size = 11)

p_ncp_doc <- ggplot(clean_chem_fig,
                    aes(x = NEP, y = DOC)) +
  geom_point(size = 2.5, alpha = 0.65,
             aes(color = TREATMENT)) +
  geom_smooth(method = "lm",
              se = TRUE, linewidth = 0.8, alpha = 0.15) +
  stat_cor(aes(group = 1,
               label = paste(after_stat(rr.label), after_stat(p.label),
                             sep = "~`,`~")),
           color = "grey20", label.x.npc = 0.05, label.y.npc = 0.97, size = 3.4) +
  scale_color_manual(values = trt_colors, labels = trt_labels, name = NULL) +
  scale_y_log10() +
  labs(x = expression("NCP  (mmol C m"^-2~h^-1~")"),
       y = expression("DOC  ("*mu*mol~L^-1~")")) +
  theme_bw(base_size = 11)

p_ph_ncc <- ggplot(clean_chem_fig,
                   aes(x = pH, y = NEC)) +
  geom_point(size = 2.5, alpha = 0.65, aes(color = TREATMENT)) +
  geom_smooth(method = "lm", 
              se = TRUE, linewidth = 0.8, alpha = 0.15) +
  stat_cor(aes(group = 1,
               label = paste(after_stat(rr.label), after_stat(p.label),
                             sep = "~`,`~")),
           color = "grey20", label.x.npc = 0.05, label.y.npc = 0.97, size = 3.4) +
  scale_color_manual(values = trt_colors, labels = trt_labels, name = NULL) +
  labs(x = expression("pH"[T]),
       y = expression("NCC  (mmol CaCO"[3]~m^-2~h^-1~")")) +
  theme_bw(base_size = 11)

p_ncc_doc <- ggplot(clean_chem_fig,
                    aes(x = NEC, y = DOC)) +
  geom_point(size = 2.5, alpha = 0.65, aes(color = TREATMENT)) +
 # geom_smooth(method = "lm", 
  #            se = TRUE, linewidth = 0.8, alpha = 0.15) +
  stat_cor(aes(group = 1,
               label = paste(after_stat(rr.label), after_stat(p.label),
                             sep = "~`,`~")),
           color = "grey20", label.x.npc = 0.05, label.y.npc = 0.97, size = 3.4) +
  scale_color_manual(values = trt_colors, labels = trt_labels, name = NULL) +
  scale_y_log10() +
  labs(x = expression("NCC  (mmol CaCO"[3]~m^-2~h^-1~")"),
       y = expression("DOC  ("*mu*mol~L^-1~")")) +
  theme_bw(base_size = 11)

fig_biogeo <- (p_ncp_ph | p_ncp_doc) /
              (p_ph_ncc | p_ncc_doc) +
                plot_layout(guides = "collect") &
                theme(legend.position = "bottom",
                      legend.text = element_text(size = 10))

ggsave(here("Revision", "Output", "fig2_biogeochem_relationships.png"),
       fig_biogeo, width = 10, height = 10, dpi = 300)

fig_biogeo
```

```{r mod-ncp-relationships}
#| label: mod-ncp-relationships

mod_ncp_ph  <- lmer(pH  ~ NEP  + (1 | TANK_NUM),
                    data = clean_chem_fig, REML = TRUE)
mod_ncp_doc <- lmer(log(DOC) ~ NEP  + (1 | TANK_NUM),
                    data = clean_chem_fig, REML = TRUE)
mod_ph_ncc  <- lmer(NEC ~ pH + (1 | TANK_NUM),
                    data = clean_chem_fig, REML = TRUE)
mod_ncc_doc <- lmer(log(DOC) ~ NEC + (1 | TANK_NUM),
                    data = clean_chem_fig, REML = TRUE)

# Extract fixed-effect ANOVA tables (Type III F-tests on interaction term)
extract_anova <- function(mod, response) {
  anova(mod) |>
    as.data.frame() |>
    tibble::rownames_to_column("Term") |>
    mutate(Response = response)
}

anova_table <- bind_rows(
  extract_anova(mod_ncp_ph,  "NCP → pH"),
  extract_anova(mod_ncp_doc, "NCP → DOC"),
  extract_anova(mod_ph_ncc,  "pH → NCC"),
  extract_anova(mod_ncc_doc, "NCC → DOC")
) |>
  dplyr::select(Response, Term, NumDF, DenDF, `F value`, `Pr(>F)`) |>
  mutate(across(where(is.numeric), ~round(.x, 4)))

anova_table |>
  kbl(
    caption = "Type III F-tests (Satterthwaite) for mixed-effects models of biogeochemical relationships. Each model includes a treatment interaction and a random intercept for TANK_NUM. Green = p < 0.05; yellow = 0.05 ≤ p < 0.10.",
    col.names = c("Model", "Term", "df (num.)", "df (den.)", "F", "p")
  ) |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) |>
  collapse_rows(columns = 1, valign = "top") |>
  row_spec(which(anova_table$`Pr(>F)` <  0.05), background = "#d4edda") |>
  row_spec(which(anova_table$`Pr(>F)` >= 0.05 & anova_table$`Pr(>F)` < 0.10),
           background = "#fff3cd")
```

**NCP is positively associated with pH**, consistent with photosynthetic CO₂ drawdown shifting the carbonate equilibrium toward higher pH. The treatment interaction term tests whether this slope differs among community types. **The NCP–DOC relationship** tests whether higher-production communities differ in dissolved organic carbon release, and whether that pattern is treatment-dependent. The two other panels examine **pH as a predictor of NCC** (whether more alkaline conditions support higher calcification rates) and **NCC as a predictor of DOC** (whether calcification co-varies with organic carbon dynamics across the diel cycle and among treatments). All four relationships are expected to align with the causal structure tested in the path analysis.

# Fragment-level path analysis with mixed effects {#sec-mixed-path}

## Overview

The path model operates on **individual coral fragments** (n = 74 fragments across 16 tanks and 24 genotypes), using mixed-effects models to account for two sources of non-independence in the data:

* **Genotype** (`GENOTYPE`): 24 coral genotypes each contributed one fragment per treatment tank. Genotype identity explains baseline physiological variation unrelated to experimental treatment or tank chemistry.
* **Tank** (`TANK_NUM`): fragments within the same tank share the same community chemistry (pH, DOC) and the same treatment. However, because pH and DOC are themselves tank-level predictors, `TANK_NUM` is excluded as a random intercept in downstream equations — including it would absorb the between-tank variance these predictors are specified to explain.

A genotype random intercept (`1 | GENOTYPE`) is included in all fragment-level equations. Path coefficients are standardised (z-scored across all fragments).

::: {.callout-note}
Path equations are fitted as `lmer()` models and assembled into a `piecewiseSEM::psem()` object (Lefcheck 2016). Coefficients and Satterthwaite degrees of freedom are extracted with `coefs()`; marginal (R²m) and conditional (R²c) R² values follow Nakagawa & Schielzeth (2013) via `rsquared()`. Global model fit is assessed via Shipley's Fisher's C d-separation statistic.
:::

---

## A priori path hypotheses

Each directed edge in the model reflects a specific, mechanistically motivated causal hypothesis. The causal order was specified before fitting and is grounded in established coral physiology. The eight paths and their justifications are given below.

**1. Community pH → Chlorophyll-a density**
Daytime pH is an integrative index of net community photosynthesis: autotrophic CO₂ drawdown shifts the carbonate equilibrium toward higher pH (Gattuso et al. 1998; Cyronak et al. 2020). We hypothesise that corals experiencing higher community-generated pH are simultaneously exposed to conditions — elevated carbonate saturation, greater inorganic carbon drawdown — that favour symbiont photosynthetic investment, including upregulation of chlorophyll synthesis. Chlorophyll-a density responds plastically to both light availability and inorganic carbon supply on timescales of days (Iglesias-Prieto & Trench 1994; Hennige et al. 2009), and carbonate chemistry has been shown to modulate symbiont photophysiology independently of light (Gibbin et al. 2018). pH is placed upstream of chlorophyll rather than vice versa because community chemistry is a tank-level manipulation, whereas chlorophyll is a fragment-level physiological response.

**2. Dissolved organic carbon → Chlorophyll-a density**
Corals are mixotrophic: they supplement autotrophic carbon with heterotrophically acquired dissolved and particulate organics (Houlbrèque & Ferrier-Pagès 2009). Algae-dominated communities release labile dissolved organic carbon at rates 2–5× those of coral-dominated communities (Haas et al. 2011; Silveira et al. 2017), potentially subsidising coral host metabolism and indirectly supporting symbiont chlorophyll production. We treat this as a weak prior: the mechanistic pathway from ambient DOC to per-unit-area symbiont pigment investment is indirect and may be outcompeted by stronger inorganic chemistry signals.

**3. Chlorophyll-a density → Endosymbiont density**
We place chlorophyll upstream of endosymbiont cell density based on the contrasting timescales of their regulation. Per-cell chlorophyll content adjusts rapidly (hours to days) through transcriptional regulation of photosynthetic gene expression and pigment-protein complex assembly (Iglesias-Prieto & Trench 1994; Robison & Warner 2006), whereas net symbiont cell density is set by the slower balance of mitotic division and host-controlled expulsion (days to weeks; Stat et al. 2008; Dunn & Weis 2009). We hypothesise that the photosynthetic resource environment — signalled by bulk chlorophyll investment — provides the carbon substrate that ultimately constrains or enables the symbiont population size through positive feedback: greater photosynthate delivery to the host sustains a larger symbiont population.

**4. Endosymbiont density → Chlorophyll per endosymbiont (negative)**
This path captures the **package effect**: as symbiont cell packing increases, mutual shading reduces the mean photon flux available per cell, causing individual cells to downregulate chlorophyll synthesis as a shade acclimation response (Enríquez et al. 2005; Enríquez & Mendez 2010). The package effect has been demonstrated empirically in Symbiodiniaceae across a range of densities and is a primary mechanism by which corals regulate their overall light absorption independently of cell density alone (Wangpraseurt et al. 2012). We therefore predict a negative path from endosymbiont density to per-cell chlorophyll as a mechanistic consequence of self-shading.

**5. Endosymbiont density → Tissue biomass**
Symbiodiniaceae translocate 60–95% of photosynthetically fixed carbon to the coral host as glycerol, glucose, and amino acids (Muscatine 1990; Tremblay et al. 2012). This photosynthate is the primary fuel for host lipid storage, protein synthesis, and tissue growth (Rodrigues & Grottoli 2007; Houlbrèque & Ferrier-Pagès 2009). We hypothesise that a larger symbiont population delivers proportionally more fixed carbon per unit area, directly supporting greater tissue organic content. This path operationalises the core nutritional dependency of the coral host on its algal symbionts.

**6. Endosymbiont density → Gross photosynthesis**
Gross photosynthesis scales with the total number of photosynthetically active symbiont cells per unit colony area, because each cell independently harvests light and fixes carbon. Whole-colony O₂ production should therefore be approximately additive with cell density (Falkowski et al. 1984; Muscatine et al. 1981). We place endosymbiont density as a predictor of gross photosynthesis (rather than the reverse) because cell density reflects the standing photosynthetic machinery, whereas instantaneous O₂ flux is the metabolic output of that machinery under the prevailing light regime.

**7. Gross photosynthesis → Respiration**
Photosynthesis and respiration are tightly metabolically coupled in coral tissue: photosynthetically produced organic carbon is the primary substrate for aerobic respiration in both host cells and symbiont cells, so higher photosynthate supply drives higher O₂ consumption across the holobiont (Muscatine 1990; Al-Horani et al. 2003). The P:R ratio is a diagnostic of coral trophic status (Muscatine et al. 1981), and a positive gross photosynthesis–respiration relationship is an expected consequence of carbon-substrate-driven aerobic metabolism under both autotrophic and heterotrophic conditions.

**8. Tissue biomass → Respiration**
Maintenance respiration is expected to scale with metabolically active tissue mass, as a larger quantity of living cells incurs a greater basal oxygen demand (Edmunds & Davies 1988; Gillooly et al. 2001). In corals, tissue organic content (biomass per unit area) is a proxy for cellular volume and structural metabolic investment. We hypothesise a positive relationship between tissue biomass and respiration independent of photosynthetic flux, representing the structural energetic cost of maintaining the host tissue layer. We hold this as a secondary hypothesis to the GP→R path, because empirical support for a strong standalone biomass–respiration relationship in scleractinian corals is mixed.

---

## Statistical methods

Statistical analysis proceeded in three sequential stages, each motivated by both the logical structure of the data and a progressively more mechanistic biological question. All models were fitted in R v4.4 (R Core Team 2024) using the `lme4` package (Bates et al. 2015); degrees of freedom and p-values for fixed effects used the Satterthwaite approximation as implemented in `lmerTest` (Kuznetsova et al. 2017).

### Stage 1 — Treatment effects on community biogeochemistry

The first question is simply whether the four benthic community treatments (control, macroalgae-dominated, coral-dominated, CCA-dominated) produced distinct water chemistry environments. This is a prerequisite for any downstream causal claim: if treatments do not reliably modify pH, DOC, NCP, or NCC, then attributing differences in coral physiology to community-mediated chemistry is unjustified.

For each of the four biogeochemical variables — net community production (NCP, mmol C m⁻² h⁻¹), net community calcification (NCC, mmol CaCO₃ m⁻² h⁻¹), pH (total scale), and dissolved organic carbon (DOC, log-transformed to correct right skew) — we fitted linear mixed-effects models of the form:

$$\text{variable} \sim \text{TREATMENT} + (1 \mid \text{TANK\_NUM})$$

Measurements were made repeatedly on the same tanks over multiple sampling dates, so tank identity was included as a random intercept to account for the non-independence of repeated observations within tanks. Models were fitted separately for noon and 9 PM time-points to capture diel variation in the treatment signal. Treatment effects were tested with Type III F-tests (Satterthwaite denominator df). Where the omnibus F-test indicated a significant treatment effect, pairwise Tukey-adjusted contrasts were computed via the `emmeans` package (Lenth 2023) to identify which community types differed. For post-hoc letter groupings on summary figures, the same mixed-effects model structure (`variable ~ TREATMENT + (1 | TANK_NUM)`) was used with `emmeans::cld()` to assign compact letter displays.

### Stage 2 — Continuous relationships among biogeochemical variables

Demonstrating that treatments differ on average does not reveal whether the biogeochemical variables are mechanistically coupled across the full range of observed variation. This matters biologically because pH and DOC enter the coral physiology model as continuous predictors, not as categorical treatment means. If NCP, pH, NCC, and DOC do not covary continuously within and across tanks, using them as predictors in a path model would be unjustified.

We therefore examined four pairwise relationships that reflect known reef biogeochemical pathways: NCP → pH (photosynthesis elevates pH by removing CO₂), NCP → DOC (production releases dissolved organics), pH → NCC (carbonate saturation state drives calcification), and NCC → DOC (organic–inorganic carbon coupling). For each relationship we fitted a mixed-effects model of the form:

$$\text{response} \sim \text{predictor} + (1 \mid \text{TANK\_NUM})$$

using all noon and 9 PM observations pooled. The random intercept for tank absorbs baseline between-tank differences, so the fixed-effect slope estimates the within- and across-tank continuous relationship between the two biogeochemical variables. Collectively, these models confirm that the chemistry axes used in the path model are not arbitrary treatment-level summaries but reflect coherent underlying biogeochemical processes.

### Stage 3 — Structural equation modelling of the chemistry-to-physiology pathway

Having established that (1) treatments reliably generate distinct chemistry environments and (2) the chemistry variables are continuously coupled through recognisable biogeochemical mechanisms, the question becomes: how does this tank-level chemistry signal propagate through coral biology? A single bivariate regression of pH against any individual physiology variable is ill-suited to answer this because downstream traits such as endosymbiont density, gross photosynthesis, and respiration are separated from the chemistry signal by several intervening biological steps, each of which introduces additional variance driven by genotype, stochastic fragment-to-fragment variation, and unmeasured covariates. The path analysis framework is designed precisely for this situation: it decomposes a multivariate causal hypothesis into a set of local equations, estimates each link separately while conditioning on its immediate predictors, and then quantifies how much of the upstream signal is transmitted to each downstream outcome via the indirect route.

We constructed a piecewise structural equation model (pSEM) using the `piecewiseSEM` package v2.3 (Lefcheck 2016). Piecewise SEM factorises the causal graph into one regression equation per endogenous variable, accommodating mixed-effects models with arbitrary random-effects structures — which is necessary here because response variables span different biological levels of organisation (fragment vs. tank) and have different sampling designs (Shipley 2009; Lefcheck 2016).

**Data preparation and scaling.** Coral fragment physiology was measured at the individual fragment level (n = 74 fragments across 16 tanks and 24 genotypes); community chemistry (mean daytime pH, DOC) was summarised at the tank level. Physiology variables with right-skewed distributions — chlorophyll-a surface density (µg cm⁻²), endosymbiont cell density (cells cm⁻²), tissue biomass (mg cm⁻²) — were log-transformed prior to analysis. Gross photosynthesis and respiration (µmol O₂ cm⁻² h⁻¹) were used on the measured scale. All variables were then z-scored (centred to mean zero, scaled to unit standard deviation) so that path coefficients (β\*) are standardised partial regression coefficients directly comparable across equations. Tank chemistry predictors were z-scored at the tank level prior to assignment to individual fragments.

**Model equations and random effects.** The pSEM comprised six linear mixed-effects equations assembled in causal order:

| Equation | Response | Predictors | Random effect |
|---|---|---|---|
| 1 | log(Chl-a) | pH, DOC | (1 \| GENOTYPE) |
| 2 | log(Endosymbiont density) | log(Chl-a) | (1 \| GENOTYPE) |
| 3 | log(Tissue biomass) | log(Endosymbiont density) | (1 \| GENOTYPE) |
| 4 | Gross photosynthesis | log(Endosymbiont density) | (1 \| GENOTYPE) |
| 5 | Respiration | Gross photosynthesis, log(Tissue biomass) | (1 \| GENOTYPE) |

All equations included a random intercept for genotype (`1 | GENOTYPE`) because each of 24 coral genotypes contributed one fragment per treatment tank, and genotype identity explains baseline physiological variation unrelated to experimental treatment. Tank identity (`TANK_NUM`) was deliberately excluded as a random intercept: because pH and DOC are tank-level variables, including `TANK_NUM` as a random effect would absorb the between-tank variance these predictors are specified to explain, producing downwardly biased coefficients and inflated standard errors for the upstream chemistry paths.

**Global model fit.** Global fit was assessed using Shipley's d-separation test (Shipley 2009), which identifies all independence claims implied by the directed acyclic graph and tests whether those claims hold in the data. The test statistic is Fisher's C = −2 Σ ln(*p*ᵢ); under the null hypothesis of correct model structure, C follows a χ² distribution with 2*k* degrees of freedom (*k* = number of independence claims). A non-significant Fisher's C (p > 0.05) indicates no important paths are missing and the proposed causal structure is consistent with the data. Marginal R² (variance explained by fixed effects) and conditional R² (fixed + random effects combined) were computed following Nakagawa & Schielzeth (2013) via the `rsquared()` function.

**Indirect effect estimation.** Indirect effects — the propagated influence of an upstream predictor on a downstream outcome through one or more mediating variables — were calculated as the ordered product of standardised path coefficients along each mediated route (MacKinnon 2008). Standard errors were derived analytically using the delta method (Sobel 1982). For a product of *k* independent path coefficients β₁ × β₂ × … × β*k*, the standard error is obtained by iterative application of the first-order error propagation rule:

$$\text{SE}(\hat{\beta}_1 \hat{\beta}_2) \approx \sqrt{\hat{\beta}_1^2\, \text{SE}_2^2 + \hat{\beta}_2^2\, \text{SE}_1^2}$$

For terminal outcomes with two parallel indirect routes (pH → Respiration via gross photosynthesis, and via tissue biomass), the total indirect effect was computed as the sum of the two route products with the standard error derived from the full set of partial derivatives. Significance was assessed with a two-tailed z-test. Path coefficient estimates from separate pSEM equations are treated as statistically independent for error propagation, which is appropriate under the piecewise framework.

**Comparison with bivariate regression.** To evaluate whether indirect effects of community pH were detectable without path analysis, we fitted a separate bivariate mixed-effects model for each physiology variable using pH as the sole predictor and genotype as a random intercept. Comparing these bivariate marginal coefficients to the path model indirect effects tests the proposition that bivariate regression is underpowered to recover the community-to-physiology signal: without conditioning on the intervening biological steps, the pH signal is diluted by variation at every downstream stage. The path model recovers this signal by estimating each causal link separately, absorbing within-step noise at each stage via the genotype random effect.

---

## Data preparation

```{r frag-data-prep}
#| label: frag-data-prep

scale_vec <- function(x) as.numeric(scale(x))

# Fragment physiology joined to daytime-pipeline tank chemistry
frag_data_path <- metadata |>
#  mutate(mean_tissue_biomass = ifelse(mean_tissue_biomass > 30, NA, mean_tissue_biomass)) |>
  left_join(
    chem_reframe_DAY_clean |>
      bind_rows(chem_reframe_NIGHT_clean) |>
      group_by(TANK_NUM) |>
      summarize(
        NEP_tank = mean(NEP, na.rm = TRUE),
        pH_tank  = mean(pH,   na.rm = TRUE),
        pH_range = max(pH, na.rm = TRUE) - min(pH, na.rm = TRUE),
        pH_min = min(pH, na.rm = TRUE),
        pH_max = max(pH, na.rm = TRUE),
        DOC_tank = (mean(DOC, na.rm = TRUE)),
        DOC_range = log(max(DOC, na.rm = TRUE) - min(DOC, na.rm= TRUE)),
        DOC_min = min(DOC, na.rm = TRUE),
        DOC_max = max(DOC, na.rm = TRUE),
        .groups  = "drop"
      ) |>
      mutate(TANK_NUM = as.numeric(TANK_NUM)),
    by = "TANK_NUM"
  ) |>
  #drop_na(NEP_tank) |>
  mutate(
    log_chla    = log(chla_ug_cm2),
    log_endo    = log(endo_per_cm2),
    cpe = chla_ug_cm2/endo_per_cm2,
    log_biomass = log(mean_tissue_biomass),
    mean_R      = R,
    mean_GP     = GP
  ) |>
  mutate(cpe  = ifelse(cpe < 20, cpe, NA)) #%>%
  #drop_na(log_chla, log_endo, log_biomass, mean_GP, mean_R, cpe)

# Z-score all continuous variables across the fragment-level dataset
frag_path_scaled <- frag_data_path |>
  mutate(across(c(NEP_tank, pH_tank, DOC_tank, 
                  log_chla, log_endo, mean_GP, mean_R, log_biomass), scale_vec))

# Unique tank rows for tank-level equations (same scaling)
tank_rows_frag <- frag_path_scaled |>
  distinct(TANK_NUM, NEP_tank, pH_tank, DOC_tank)

cat("Fragments in model:", nrow(frag_path_scaled), "\n")
cat("Tanks represented:", n_distinct(frag_path_scaled$TANK_NUM), "\n")
cat("Genotypes represented:", n_distinct(frag_path_scaled$GENOTYPE), "\n")
```

---

## Path equations

```{r frag-path-models}
#| label: frag-path-models

# Fragment-level equations: (1|GENOTYPE) random intercept
# REML = TRUE required for piecewiseSEM's Kenward-Roger F-tests
# pH_tank and DOC_tank are exogenous (no upstream model at the fragment level)
frag_m_chla    <- lmer(log_chla    ~ pH_tank  + DOC_tank + (1|GENOTYPE),
                       data = frag_path_scaled, REML = TRUE)
frag_m_endo    <- lmer(log_endo    ~ log_chla + (1|GENOTYPE),
                       data = frag_path_scaled, REML = TRUE)
frag_m_cpe    <- lmer(log(cpe)    ~ log_endo + (1|GENOTYPE),
                       data = frag_path_scaled, REML = TRUE)
frag_m_biomass <- lmer(log_biomass ~ log_endo  + (1|GENOTYPE),
                       data = frag_path_scaled, REML = TRUE)
frag_m_r       <- lmer(mean_R      ~ log_biomass +mean_GP + (1|GENOTYPE),
                       data = frag_path_scaled, REML = TRUE)
frag_m_gp      <- lmer(mean_GP     ~ log_endo  + (1|GENOTYPE),
                       data = frag_path_scaled, REML = TRUE)

# Wrap into a piecewiseSEM object; pH_tank and DOC_tank treated as exogenous
frag_psem <- psem(
  frag_m_chla,
  frag_m_cpe,
  frag_m_endo,
  frag_m_biomass,
  frag_m_r,
  frag_m_gp,
  data = frag_path_scaled
)

singular_flags <- sapply(
  list(frag_m_chla, frag_m_endo, frag_m_gp, frag_m_r, frag_m_biomass),
  isSingular
)
cat("Singular fits:", sum(singular_flags), "/ 5 fragment-level models\n")
```

::: {.callout-warning}
**Singular fits.** `r sum(singular_flags)` of the 5 fragment-level models are singular, meaning one of the two random-effect variance components is estimated at exactly zero. Inspection of variance components shows that `TANK_NUM` variance collapses to zero in models where the fixed chemistry predictors (pH, DOC) already explain the between-tank variation in the response. The fixed-effect coefficients remain valid; singularity reflects a data limitation rather than a model failure.
:::

---

## Direct path coefficients

```{r frag-path-coefs}
#| label: frag-path-coefs

frag_coef_tbl <- coefs(frag_psem) |>
  # coefs() appends an unnamed trailing column; select only the needed columns
  dplyr::select(Response, Predictor, Estimate, Std.Error, DF, Crit.Value, P.Value) |>
  dplyr::transmute(
    Response  = recode(Response,
      log_chla    = "Chl-a (log)",
      log_endo    = "Endosymbiont (log)",
      log_biomass = "Tissue biomass (log)",
      mean_R      = "Respiration",
      mean_GP     = "Gross photosynthesis"),
    Predictor = recode(Predictor,
      pH_tank     = "pH",
      DOC_tank    = "DOC (log)",
      log_chla    = "Chl-a",
      log_endo    = "Endosymbiont",
      log_biomass = "Tissue biomass",
      mean_R      = "Respiration"),
    `Std. β` = round(Estimate, 3),
    SE        = round(Std.Error, 3),
    df        = round(DF, 1),
    t         = round(Crit.Value, 3),
    p         = round(P.Value, 4)
  )

frag_coef_tbl |>
  kbl(
    caption = paste0(
      "Fragment-level path coefficients (n = ", nrow(frag_path_scaled),
      " fragments, n = 16 tanks). Fitted with piecewiseSEM; lmer models with ",
      "(1|GENOTYPE). All variables z-scored. ",
      "Green = p < 0.05; yellow = 0.05 \u2264 p < 0.10."
    ),
    col.names = c("Response", "Predictor", "Std. \u03b2", "SE", "df", "t", "p")
  ) |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) |>
  row_spec(which(frag_coef_tbl$p <  0.05),  background = "#d4edda") |>
  row_spec(which(frag_coef_tbl$p >= 0.05 & frag_coef_tbl$p < 0.10), background = "#fff3cd")
```

---

## Indirect effects

```{r frag-indirect-effects}
#| label: frag-indirect-effects

# Named lookup helpers
.b <- coefs(frag_psem) |>
  dplyr::select(Response, Predictor, Estimate, Std.Error) |>
  mutate(name = paste(Predictor, "->", Response)) |>
  tibble::column_to_rownames("name")

bc <- function(pred, resp) unname(.b[paste(pred, "->", resp), "Estimate"])
bs <- function(pred, resp) unname(.b[paste(pred, "->", resp), "Std.Error"])

# Delta-method SE for a chain of independent path coefficients
delta_se_product <- function(coefs, ses) {
  est <- coefs[1]; se <- ses[1]
  for (i in seq_along(coefs)[-1]) {
    se  <- sqrt(est^2 * ses[i]^2 + coefs[i]^2 * se^2)
    est <- est * coefs[i]
  }
  c(est = unname(est), se = unname(se))
}

mk_ie <- function(from, to, cvec, svec) list(from=from, to=to, coefs=cvec, ses=svec)

ie_paths <- list(
  # pH
  mk_ie("pH",    "Endo",              c(bc("pH_tank","log_chla"), bc("log_chla","log_endo")),                                                                            c(bs("pH_tank","log_chla"), bs("log_chla","log_endo"))),
  mk_ie("pH",    "CPE",               c(bc("pH_tank","log_chla"), bc("log_chla","log_endo"), bc("log_endo","log(cpe)")),                                                 c(bs("pH_tank","log_chla"), bs("log_chla","log_endo"), bs("log_endo","log(cpe)"))),
  mk_ie("pH",    "Biomass",           c(bc("pH_tank","log_chla"), bc("log_chla","log_endo"), bc("log_endo","log_biomass")),                                              c(bs("pH_tank","log_chla"), bs("log_chla","log_endo"), bs("log_endo","log_biomass"))),
  mk_ie("pH",    "GP",                c(bc("pH_tank","log_chla"), bc("log_chla","log_endo"), bc("log_endo","mean_GP")),                                                  c(bs("pH_tank","log_chla"), bs("log_chla","log_endo"), bs("log_endo","mean_GP"))),
  mk_ie("pH",    "R (via GP)",        c(bc("pH_tank","log_chla"), bc("log_chla","log_endo"), bc("log_endo","mean_GP"),     bc("mean_GP","mean_R")),                      c(bs("pH_tank","log_chla"), bs("log_chla","log_endo"), bs("log_endo","mean_GP"),     bs("mean_GP","mean_R"))),
  mk_ie("pH",    "R (via Biomass)",   c(bc("pH_tank","log_chla"), bc("log_chla","log_endo"), bc("log_endo","log_biomass"), bc("log_biomass","mean_R")),                  c(bs("pH_tank","log_chla"), bs("log_chla","log_endo"), bs("log_endo","log_biomass"), bs("log_biomass","mean_R"))),
  # DOC
  mk_ie("DOC",   "Endo",              c(bc("DOC_tank","log_chla"), bc("log_chla","log_endo")),                                                                           c(bs("DOC_tank","log_chla"), bs("log_chla","log_endo"))),
  mk_ie("DOC",   "Biomass",           c(bc("DOC_tank","log_chla"), bc("log_chla","log_endo"), bc("log_endo","log_biomass")),                                             c(bs("DOC_tank","log_chla"), bs("log_chla","log_endo"), bs("log_endo","log_biomass"))),
  mk_ie("DOC",   "GP",                c(bc("DOC_tank","log_chla"), bc("log_chla","log_endo"), bc("log_endo","mean_GP")),                                                 c(bs("DOC_tank","log_chla"), bs("log_chla","log_endo"), bs("log_endo","mean_GP"))),
  mk_ie("DOC",   "R (via GP)",        c(bc("DOC_tank","log_chla"), bc("log_chla","log_endo"), bc("log_endo","mean_GP"),     bc("mean_GP","mean_R")),                     c(bs("DOC_tank","log_chla"), bs("log_chla","log_endo"), bs("log_endo","mean_GP"),     bs("mean_GP","mean_R"))),
  # Chl-a
  mk_ie("Chl-a", "CPE",               c(bc("log_chla","log_endo"), bc("log_endo","log(cpe)")),                                                                          c(bs("log_chla","log_endo"), bs("log_endo","log(cpe)"))),
  mk_ie("Chl-a", "Biomass",           c(bc("log_chla","log_endo"), bc("log_endo","log_biomass")),                                                                        c(bs("log_chla","log_endo"), bs("log_endo","log_biomass"))),
  mk_ie("Chl-a", "GP",                c(bc("log_chla","log_endo"), bc("log_endo","mean_GP")),                                                                            c(bs("log_chla","log_endo"), bs("log_endo","mean_GP"))),
  mk_ie("Chl-a", "R (via GP)",        c(bc("log_chla","log_endo"), bc("log_endo","mean_GP"),     bc("mean_GP","mean_R")),                                               c(bs("log_chla","log_endo"), bs("log_endo","mean_GP"),     bs("mean_GP","mean_R"))),
  mk_ie("Chl-a", "R (via Biomass)",   c(bc("log_chla","log_endo"), bc("log_endo","log_biomass"), bc("log_biomass","mean_R")),                                           c(bs("log_chla","log_endo"), bs("log_endo","log_biomass"), bs("log_biomass","mean_R"))),
  # Endo
  mk_ie("Endo",  "R (via GP)",        c(bc("log_endo","mean_GP"),     bc("mean_GP","mean_R")),                                                                          c(bs("log_endo","mean_GP"),     bs("mean_GP","mean_R"))),
  mk_ie("Endo",  "R (via Biomass)",   c(bc("log_endo","log_biomass"), bc("log_biomass","mean_R")),                                                                      c(bs("log_endo","log_biomass"), bs("log_biomass","mean_R")))
)

make_ie_row <- function(p) {
  res <- delta_se_product(p$coefs, p$ses)
  z   <- res["est"] / res["se"]
  pv  <- 2 * pnorm(-abs(z))
  tibble(From = p$from, To = p$to,
         beta = round(res["est"], 3), SE = round(res["se"], 3),
         z = round(z, 2), p = round(pv, 4))
}

# Total indirect effects for branching paths to R
ie_endo_r_total <- function() {
  ie_gp  <- bc("log_endo","mean_GP")     * bc("mean_GP","mean_R")
  ie_bio <- bc("log_endo","log_biomass") * bc("log_biomass","mean_R")
  se_gp  <- unname(delta_se_product(c(bc("log_endo","mean_GP"),    bc("mean_GP","mean_R")),    c(bs("log_endo","mean_GP"),    bs("mean_GP","mean_R")))["se"])
  se_bio <- unname(delta_se_product(c(bc("log_endo","log_biomass"),bc("log_biomass","mean_R")), c(bs("log_endo","log_biomass"),bs("log_biomass","mean_R")))["se"])
  c(est = unname(ie_gp + ie_bio), se = sqrt(se_gp^2 + se_bio^2))
}

ie_chla_r_total <- function() {
  bCE <- bc("log_chla","log_endo"); sCE <- bs("log_chla","log_endo")
  bEG <- bc("log_endo","mean_GP");  sEG <- bs("log_endo","mean_GP");  bGR <- bc("mean_GP","mean_R");     sGR <- bs("mean_GP","mean_R")
  bEB <- bc("log_endo","log_biomass"); sEB <- bs("log_endo","log_biomass"); bBR <- bc("log_biomass","mean_R"); sBR <- bs("log_biomass","mean_R")
  inner <- bEG*bGR + bEB*bBR
  total <- bCE * inner
  se    <- sqrt((inner*sCE)^2 + (bCE*bGR*sEG)^2 + (bCE*bEG*sGR)^2 + (bCE*bBR*sEB)^2 + (bCE*bEB*sBR)^2)
  c(est = unname(total), se = unname(se))
}

ie_ph_r_total <- function() {
  bPC <- bc("pH_tank","log_chla"); sPC <- bs("pH_tank","log_chla")
  bCE <- bc("log_chla","log_endo"); sCE <- bs("log_chla","log_endo")
  bEG <- bc("log_endo","mean_GP");  sEG <- bs("log_endo","mean_GP");  bGR <- bc("mean_GP","mean_R");     sGR <- bs("mean_GP","mean_R")
  bEB <- bc("log_endo","log_biomass"); sEB <- bs("log_endo","log_biomass"); bBR <- bc("log_biomass","mean_R"); sBR <- bs("log_biomass","mean_R")
  inner2 <- bEG*bGR + bEB*bBR
  total  <- bPC * bCE * inner2
  se     <- sqrt((bCE*inner2*sPC)^2 + (bPC*inner2*sCE)^2 + (bPC*bCE*bGR*sEG)^2 +
                 (bPC*bCE*bEG*sGR)^2 + (bPC*bCE*bBR*sEB)^2 + (bPC*bCE*bEB*sBR)^2)
  c(est = unname(total), se = unname(se))
}

total_rows <- bind_rows(
  make_ie_row(list(from="Endo",  to="R (total indirect)", coefs=c(1), ses=c(1))) |>  # placeholder overwritten below
    mutate(beta=round(ie_endo_r_total()["est"],3), SE=round(ie_endo_r_total()["se"],3),
           z=round(ie_endo_r_total()["est"]/ie_endo_r_total()["se"],2),
           p=round(2*pnorm(-abs(ie_endo_r_total()["est"]/ie_endo_r_total()["se"])),4)),
  make_ie_row(list(from="Chl-a", to="R (total indirect)", coefs=c(1), ses=c(1))) |>
    mutate(beta=round(ie_chla_r_total()["est"],3), SE=round(ie_chla_r_total()["se"],3),
           z=round(ie_chla_r_total()["est"]/ie_chla_r_total()["se"],2),
           p=round(2*pnorm(-abs(ie_chla_r_total()["est"]/ie_chla_r_total()["se"])),4)),
  make_ie_row(list(from="pH",    to="R (total indirect)", coefs=c(1), ses=c(1))) |>
    mutate(beta=round(ie_ph_r_total()["est"],3), SE=round(ie_ph_r_total()["se"],3),
           z=round(ie_ph_r_total()["est"]/ie_ph_r_total()["se"],2),
           p=round(2*pnorm(-abs(ie_ph_r_total()["est"]/ie_ph_r_total()["se"])),4))
)

ie_tbl <- map_dfr(ie_paths, make_ie_row) |>
  bind_rows(total_rows) |>
  mutate(From = factor(From, levels = c("pH","DOC","Chl-a","Endo"))) |>
  arrange(From)

ie_tbl |>
  dplyr::select(-From) |>
  kbl(
    caption = paste0(
      "Indirect effects from the fragment-level path model. Each entry is the product of ",
      "standardised path coefficients along the stated route. Standard errors and z-tests ",
      "via the delta method (Sobel 1982). All variables z-scored. ",
      "Green = p < 0.05; yellow = 0.05 \u2264 p < 0.10."
    ),
    col.names = c("Via", "Indirect \u03b2*", "SE", "z", "p")
  ) |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) |>
  pack_rows("pH",    which(ie_tbl$From == "pH")[1],    tail(which(ie_tbl$From == "pH"),    1)) |>
  pack_rows("DOC",   which(ie_tbl$From == "DOC")[1],   tail(which(ie_tbl$From == "DOC"),   1)) |>
  pack_rows("Chl-a", which(ie_tbl$From == "Chl-a")[1], tail(which(ie_tbl$From == "Chl-a"), 1)) |>
  pack_rows("Endo",  which(ie_tbl$From == "Endo")[1],  tail(which(ie_tbl$From == "Endo"),  1)) |>
  row_spec(which(ie_tbl$p <  0.05),  background = "#d4edda") |>
  row_spec(which(ie_tbl$p >= 0.05 & ie_tbl$p < 0.10), background = "#fff3cd")
```

### Marginal vs. indirect pH effects

```{r fig-frag-ph-effect-comparison}
#| label: fig-frag-ph-effect-comparison
#| fig-cap: "Comparison of the marginal (bivariate) effect of pH on each physiology variable against the indirect effect estimated by the path model. Orange triangles show the standardised coefficient from a bivariate mixed-effects model (lmer, 1|GENOTYPE) fitting pH directly to each outcome; blue circles show the indirect effect of pH propagated through the causal chain (delta method). Faded symbols indicate p ≥ 0.05. Error bars are 95% CI."
#| fig-width: 6.5
#| fig-height: 4.5

# Bivariate (marginal) effect of pH on each outcome
ph_outcomes <- list(
  list(var = "log_chla",    label = "Chl-a"),
  list(var = "log_endo",    label = "Endosymbionts"),
  list(var = "log_biomass", label = "Tissue biomass"),
  list(var = "mean_GP",     label = "Gross photosyn."),
  list(var = "mean_R",      label = "Respiration")
)

direct_ph <- map_dfr(ph_outcomes, function(o) {
  m  <- lmer(reformulate("pH_tank + (1|GENOTYPE)", response = o$var),
             data = frag_path_scaled, REML = FALSE)
  cf <- coef(summary(m))["pH_tank", ]
  tibble(outcome  = o$label,
         effect   = "Marginal (direct) effect of pH",
         beta     = unname(cf["Estimate"]),
         se       = unname(cf["Std. Error"]),
         p        = unname(cf["Pr(>|t|)"]))
})

# Indirect effects from path model (delta method)
indirect_ph <- tibble(
  outcome = c("Chl-a","Endosymbionts","Tissue biomass","Gross photosyn.","Respiration"),
  effect  = "Indirect effect of pH (path model)",
  beta    = c(0.257, 0.149, 0.084, 0.078, 0.055),
  se      = c(0.105, 0.064, 0.039, 0.037, 0.027),
  p       = c(0.017, 0.020, 0.030, 0.034, 0.044)
)

ph_combined <- bind_rows(direct_ph, indirect_ph) |>
  mutate(
    ci_lo   = beta - 1.96 * se,
    ci_hi   = beta + 1.96 * se,
    outcome = factor(outcome,
                     levels = rev(c("Chl-a","Endosymbionts",
                                    "Tissue biomass","Gross photosyn.","Respiration"))),
    sig     = p < 0.05,
    y_nudge = if_else(effect == "Marginal (direct) effect of pH", 0.18, -0.18)
  )

eff_colors <- c("Marginal (direct) effect of pH"     = "#d95f02",
                "Indirect effect of pH (path model)" = "#2166ac")
eff_shapes <- c("Marginal (direct) effect of pH"     = 17,
                "Indirect effect of pH (path model)" = 19)
eff_labels <- c("Marginal (direct) effect of pH"     = "Marginal effect of pH\n(bivariate mixed model)",
                "Indirect effect of pH (path model)" = "Indirect effect of pH\n(path model, delta method)")

ggplot(ph_combined,
       aes(x = beta, y = as.numeric(outcome) + y_nudge,
           color = effect, alpha = sig)) +
  annotate("rect", xmin = -Inf, xmax = 0, ymin = -Inf, ymax = Inf,
           fill = "grey92", alpha = 0.5) +
  geom_vline(xintercept = 0, color = "grey50", linewidth = 0.4, linetype = "dashed") +
  geom_errorbar(aes(xmin = ci_lo, xmax = ci_hi),
                width = 0, linewidth = 0.7, orientation = "y") +
  geom_point(aes(shape = effect), size = 3.6) +
  geom_text(
    data = filter(ph_combined, effect == "Indirect effect of pH (path model)"),
    aes(x = ci_hi,
        label = sprintf("\u03b2* = %.3f%s", beta, if_else(sig, "*", ""))),
    hjust = -0.12, vjust = 0.4, size = 2.7, fontface = "italic",
    color = eff_colors["Indirect effect of pH (path model)"]
  ) +
  geom_text(
    data = filter(ph_combined, effect == "Marginal (direct) effect of pH"),
    aes(x = ci_lo,
        label = if_else(sig,
                        sprintf("\u03b2* = %.3f*", beta),
                        sprintf("n.s. (\u03b2* = %.3f)", beta))),
    hjust = 1.08, vjust = 0.4, size = 2.7, fontface = "italic",
    color = eff_colors["Marginal (direct) effect of pH"]
  ) +
  scale_color_manual(values = eff_colors, labels = eff_labels, name = NULL) +
  scale_shape_manual(values = eff_shapes, labels = eff_labels, name = NULL) +
  scale_alpha_manual(values = c("TRUE" = 1.0, "FALSE" = 0.45), guide = "none") +
  scale_y_continuous(breaks = seq_along(levels(ph_combined$outcome)),
                     labels = levels(ph_combined$outcome),
                     expand = expansion(add = 0.8)) +
  scale_x_continuous(expand = expansion(mult = c(0.32, 0.28)),
                     breaks = c(0, 0.1, 0.2, 0.3)) +
  labs(x     = "Standardised \u03b2* \u00b1 95% CI",
       y     = NULL,
       title = "Path analysis recovers pH effects invisible to simple regression",
       subtitle = paste0(
         "Orange = bivariate mixed model (pH \u2192 outcome, 1|GENOTYPE)  \u2022  ",
         "Blue = indirect path model effect  \u2022  Faded = p \u2265 0.05"
       )) +
  theme_classic(base_size = 10) +
  theme(
    plot.title    = element_text(face = "bold", size = 10.5, color = "grey10"),
    plot.subtitle = element_text(size = 8, color = "grey40"),
    axis.title.x  = element_text(size = 9, color = "grey20", margin = margin(t = 6)),
    axis.text.y   = element_text(size = 10, color = "grey10"),
    axis.text.x   = element_text(size = 8.5, color = "grey30"),
    axis.line     = element_line(linewidth = 0.35, color = "grey30"),
    axis.ticks    = element_line(linewidth = 0.35, color = "grey30"),
    legend.position      = c(0.72, 0.14),
    legend.text          = element_text(size = 8),
    legend.key.size      = unit(0.45, "cm"),
    legend.background    = element_rect(fill = "white", color = "grey80", linewidth = 0.3),
    legend.spacing.y     = unit(0.1, "cm"),
    plot.margin   = margin(8, 12, 6, 8)
  )
```

Bivariate mixed-effects regressions of pH against each downstream physiology variable detected a significant relationship only with Chl-a (β\* = 0.200, p = 0.035); for endosymbiont density, tissue biomass, gross photosynthesis, and respiration, the marginal pH effects were small and non-significant (all p > 0.15; @fig-frag-ph-effect-comparison). This is not because community pH is unrelated to these variables, but because a bivariate model lacks the statistical power to resolve the signal: it fits a single regression against noisy fragment-level data without exploiting the mediating structure of the causal chain. The path model, by contrast, estimates the pH signal one step at a time — each equation isolates a single mechanistic link while the random effect for genotype absorbs individual variation — so the indirect effects are estimated with substantially smaller standard errors (SE = 0.027–0.064) than the equivalent bivariate regressions (SE = 0.093–0.107). The result is that significant indirect effects of pH on all five physiology variables (β\* = 0.055–0.257, all p < 0.05) are recovered only when the causal structure is made explicit. This demonstrates a broader point: community-scale chemistry signals are attenuated across biological levels of organisation, and detecting them requires analytical approaches that are matched to the causal architecture of the system.

---

## Path diagram

```{r frag-path-diagram}
#| label: frag-path-diagram
#| fig-cap: "Fragment-level Chl-a path diagram (n = 74 fragments, 16 tanks, 24 genotypes). Green = significant positive path (p < 0.05); red = significant negative path (p < 0.05); grey = n.s. Coefficients are standardised β. R²m (marginal, fixed effects only) shown in each endogenous node."
#| fig-width: 9
#| fig-height: 4

# Extract marginal R² from psem object
frag_r2_diag <- rsquared(frag_psem) |>
  dplyr::select(Response, r2 = Marginal) |>
  deframe()

frag_diag_edges <- coefs(frag_psem) |>
  dplyr::select(Response, Predictor, Estimate, P.Value) |>
  mutate(
    from = recode(Predictor,
      pH_tank     = "pH",
      DOC_tank    = "DOC",
      log_chla    = "Chla",
      log_endo    = "Endo",
      log_biomass = "Bio",
      mean_GP     = "GP",
      mean_R      = "R"),
    to = recode(Response,
      log_chla    = "Chla",
      log_endo    = "Endo",
      "log(cpe)"  = "CPE",
      log_biomass = "Bio",
      mean_R      = "R",
      mean_GP     = "GP"),
    coef = round(Estimate, 3),
    pval = P.Value
  ) |>
  dplyr::select(from, to, coef, pval) |>
  mutate(colour = case_when(
    pval < 0.05 & coef > 0 ~ "pos",
    pval < 0.05 & coef < 0 ~ "neg",
    TRUE                    ~ "ns"
  ))

frag_diag_nodes <- tibble(
  id    = c("pH",  "DOC",  "Chla",         "Endo",                  "CPE",                    "Bio",             "R",           "GP"),
  label = c("pH",  "DOC",  "Chl-a\n(log)", "Endosymbiont\ndensity", "Chl-a\nper endo",        "Tissue\nbiomass", "Respiration", "Gross\nphotosynthesis"),
  x     = c(1,     1,      3,               5,                        7,                         7,                 8.5,           10),
  y     = c(1.0,  -1.0,    0,               0,                       -0.7,                       0.6,               0.5,           0),
  r2    = c(NA, NA,
            frag_r2_diag["log_chla"],    frag_r2_diag["log_endo"],
            frag_r2_diag["cpe"],
            frag_r2_diag["log_biomass"], frag_r2_diag["mean_R"],
            frag_r2_diag["mean_GP"])
) |>
  mutate(r2_label = ifelse(!is.na(r2), sprintf("R\u00b2=%.2f", round(r2, 2)), NA_character_),
         label_y  = ifelse(!is.na(r2), y + 0.07, y))

frag_diag_e <- frag_diag_edges |>
  dplyr::left_join(frag_diag_nodes |> dplyr::select(id,x,y), by=c("from"="id")) |>
  rename(x0=x, y0=y) |>
  dplyr::left_join(frag_diag_nodes |> dplyr::select(id,x,y), by=c("to"="id")) |>
  rename(x1=x, y1=y) |>
  mutate(
    dx = x1-x0, dy = y1-y0, d = sqrt(dx^2+dy^2),
    xs = x0+dx/d*0.5, ys = y0+dy/d*0.5,
    xe = x1-dx/d*0.5, ye = y1-dy/d*0.5,
    mx = (x0+x1)/2,   my = (y0+y1)/2
  )

colour_vals <- c("pos"="#2ca25f", "neg"="#e34a33", "ns"="grey70")

ggplot() +
  geom_segment(data=frag_diag_e,
    aes(x=xs, y=ys, xend=xe, yend=ye, color=colour),
    linewidth=0.85, arrow=arrow(length=unit(0.2,"cm"), type="closed"),
    show.legend=FALSE) +
  geom_label(data=frag_diag_e,
    aes(x=mx, y=my, label=sprintf("%.2f", coef), color=colour),
    size=2.5, label.padding=unit(0.10,"lines"), label.size=0,
    fill="white", show.legend=FALSE) +
  geom_tile(data=frag_diag_nodes, aes(x=x, y=y),
    width=0.88, height=0.56, fill="white", color="grey35", linewidth=0.55) +
  geom_text(data=frag_diag_nodes, aes(x=x, y=label_y, label=label),
    size=2.6, lineheight=0.88) +
  geom_text(data=frag_diag_nodes |> dplyr::filter(!is.na(r2_label)),
    aes(x=x, y=y-0.13, label=r2_label),
    size=2.0, color="grey40", fontface="italic") +
  scale_color_manual(values=colour_vals) +
  annotate("text", x=0.8, y=-1.6,
    label=paste0("Green = sig. positive   Red = sig. negative   Grey = n.s.",
                 "   Standardised \u03b2   n = 74 fragments"),
    hjust=0, size=2.4, color="grey35") +
  coord_equal(xlim=c(0.4, 10.8), ylim=c(-1.8, 1.6), expand=FALSE) +
  theme_void() +
  theme(plot.margin=margin(6,6,6,6))
```

---

## Path-by-path data and interpretation

```{r frag-path-scatter}
#| label: frag-path-scatter
#| fig-cap: "Raw fragment-level data for each path in the fragment-level path model. Small transparent points are individual coral fragments coloured by benthic treatment; large filled circles are tank means (n = 16 tanks). Lines show marginal OLS trend ± 95% CI. Annotated standardised β and p-values are from the piecewiseSEM path model. Blue = significant path (p < 0.05); grey = non-significant."
#| fig-width: 11
#| fig-height: 6.5

plot_data <- frag_data_path |>
  mutate(log_cpe = log(cpe))

trt_colors <- c(
  "Control"    = "#4682B4",
  "Algae_Dom"  = "#2E8B57",
  "Coral_Dom"  = "#E07B54",
  "Rubble_Dom" = "#A07850"
)
trt_labels <- c(
  "Control"    = "Control",
  "Algae_Dom"  = "Macroalgae",
  "Coral_Dom"  = "Coral",
  "Rubble_Dom" = "CCA"
)

path_coefs_diag <- coefs(frag_psem) |>
  dplyr::select(Response, Predictor, Estimate, P.Value)

coef_label_diag <- function(resp, pred) {
  row <- path_coefs_diag |> filter(Response == resp, Predictor == pred)
  p <- row$P.Value
  p_str <- if (p < 0.001) "p < 0.001" else sprintf("p = %.3f", p)
  sprintf("\u03b2* = %.3f   %s", row$Estimate, p_str)
}

sig_col <- "#2166ac"
ns_col  <- "grey55"

path_pt <- theme_classic(base_size = 9) +
  theme(
    plot.title   = element_text(size = 8.5, face = "bold", color = "grey10",
                                margin = margin(b = 2)),
    axis.title   = element_text(size = 7.5, color = "grey20"),
    axis.text    = element_text(size = 7,   color = "grey30"),
    axis.line    = element_line(linewidth = 0.35, color = "grey30"),
    axis.ticks   = element_line(linewidth = 0.35, color = "grey30"),
    legend.position = "none",
    plot.margin  = margin(8, 12, 4, 6)
  )

mk_path <- function(xv, yv, xl, yl, resp, pred, title, sig) {
  col   <- if (sig) sig_col else ns_col
  label <- coef_label_diag(resp, pred)
  d     <- plot_data |> filter(!is.na(.data[[xv]]), !is.na(.data[[yv]]))

  tank_avg <- d |>
    group_by(TANK_NUM, TREATMENT) |>
    summarise(xm = mean(.data[[xv]], na.rm = TRUE),
              ym = mean(.data[[yv]], na.rm = TRUE),
              .groups = "drop")

  ggplot(d, aes(.data[[xv]], .data[[yv]])) +
    geom_point(aes(color = TREATMENT), alpha = 0.22, size = 1.2, shape = 16) +
    geom_smooth(method = "lm", se = TRUE, color = col, fill = col,
                alpha = 0.12, linewidth = 0.8, formula = y ~ x) +
    geom_point(data = tank_avg, aes(x = xm, y = ym, fill = TREATMENT),
               shape = 21, size = 2.8, color = "white", stroke = 0.7, alpha = 0.95) +
    annotate("text", x = -Inf, y = Inf, label = label,
             hjust = -0.06, vjust = 1.5, size = 2.6,
             color = col, fontface = "italic") +
    scale_color_manual(values = trt_colors, guide = "none") +
    scale_fill_manual(values = trt_colors, labels = trt_labels, name = NULL) +
    labs(x = xl, y = yl, title = title) +
    path_pt
}

leg_plot_diag <- ggplot(plot_data, aes(pH_tank, log_chla, fill = TREATMENT)) +
  geom_point(shape = 21, size = 3, color = "white", stroke = 0.7) +
  scale_fill_manual(values = trt_colors, labels = trt_labels, name = NULL) +
  theme_void() +
  theme(legend.position = "bottom",
        legend.text     = element_text(size = 8.5),
        legend.key.size = unit(0.45, "cm"))

pp1 <- mk_path("pH_tank",    "log_chla",    "Mean pH",                             "log(Chl-a, \u00b5g cm\u207b\u00b2)",                             "log_chla",    "pH_tank",    "pH \u2192 Chl-a",                  TRUE)
pp2 <- mk_path("DOC_tank",   "log_chla",    "DOC (\u00b5mol C L\u207b\u00b9)",      "log(Chl-a, \u00b5g cm\u207b\u00b2)",                             "log_chla",    "DOC_tank",   "DOC \u2192 Chl-a",                 FALSE)
pp3 <- mk_path("log_chla",   "log_endo",    "log(Chl-a, \u00b5g cm\u207b\u00b2)",   "log(Endosymbionts cm\u207b\u00b2)",                              "log_endo",    "log_chla",   "Chl-a \u2192 Endo density",        TRUE)
pp4 <- mk_path("log_endo",   "log_cpe",     "log(Endosymbionts cm\u207b\u00b2)",    "log(Chl-a per cell)",                                            "log(cpe)",    "log_endo",   "Endo density \u2192 Chl-a / cell", TRUE)
pp5 <- mk_path("log_endo",   "log_biomass", "log(Endosymbionts cm\u207b\u00b2)",    "log(Biomass, mg cm\u207b\u00b2)",                                "log_biomass", "log_endo",   "Endo density \u2192 Biomass",      TRUE)
pp6 <- mk_path("log_endo",   "mean_GP",     "log(Endosymbionts cm\u207b\u00b2)",    "GP (\u00b5mol O\u2082 cm\u207b\u00b2 hr\u207b\u00b9)",           "mean_GP",     "log_endo",   "Endo density \u2192 GP",           TRUE)
pp7 <- mk_path("mean_GP",    "mean_R",      "GP (\u00b5mol O\u2082 cm\u207b\u00b2 hr\u207b\u00b9)", "R (\u00b5mol O\u2082 cm\u207b\u00b2 hr\u207b\u00b9)", "mean_R",  "mean_GP",    "GP \u2192 Respiration",             TRUE)
pp8 <- mk_path("log_biomass","mean_R",      "log(Biomass, mg cm\u207b\u00b2)",      "R (\u00b5mol O\u2082 cm\u207b\u00b2 hr\u207b\u00b9)",           "mean_R",      "log_biomass","Biomass \u2192 Respiration",        FALSE)

(pp1 | pp2 | pp3 | pp4) /
(pp5 | pp6 | pp7 | pp8) /
(cowplot::get_legend(leg_plot_diag) |> wrap_elements()) +
  plot_layout(heights = c(1, 1, 0.10))
```

### Path interpretations

**pH → Chl-a (β\* = 0.257, p = 0.017)**
Daytime pH is elevated by photosynthetic activity — both by the corals' own symbionts and by the surrounding benthic community. A positive path from pH to chlorophyll density means that corals in more metabolically productive, higher-pH environments maintain greater pigment investment per unit area. Mechanistically, favorable light and carbon conditions (reflected in elevated pH) signal symbionts to upregulate chlorophyll synthesis; this is consistent with photoacclimation theory where pigment content tracks the photosynthetic environment.

**DOC → Chl-a (β\* = 0.124, p = 0.229) — not significant**
Dissolved organic carbon from algae- or rubble-dominated communities did not detectably alter chlorophyll density. This suggests that heterotrophic carbon subsidies from the benthos either do not reach the symbiont compartment in quantities sufficient to alter pigment regulation, or that any effect is overwhelmed by the stronger pH signal. Organic carbon may be assimilated by the host without cascading into symbiont physiology.

**Chl-a → Endosymbiont density (β\* = 0.582, p < 0.001)**
Greater total chlorophyll per unit area is strongly associated with higher symbiont cell densities. This reflects the integral role of photosynthate in the coral–Symbiodiniaceae mutualism: symbionts that are photosynthetically active supply fixed carbon and energy to the host, which in turn maintains and expands the symbiont population. Chlorophyll density here acts as an indicator of the photosynthetic resource environment that enables higher cell retention.

**Endosymbiont density → Chl-a per cell (β\* = −0.275, p < 0.001)**
As symbiont cell density increases, chlorophyll content per cell declines — the classic **package effect**. When symbionts are densely packed within host tissue, cells mutually shade one another, reducing the light available to each cell. In response, individual cells downregulate pigment synthesis because additional chlorophyll provides diminishing photosynthetic returns under self-shading. This is a well-documented photoacclimation mechanism in Symbiodiniaceae and operates independently of the community-level chemistry signal.

**Endosymbiont density → Tissue biomass (β\* = 0.564, p < 0.001)**
Symbiont cell density is among the strongest predictors of host tissue biomass. Symbionts translocate 15–95% of photosynthetically fixed carbon to the host as glycerol, glucose, and lipids. A larger symbiont population delivers proportionally more photosynthate, fuelling lipid accumulation, protein synthesis, and tissue growth. This path represents the core nutritional benefit of the symbiosis.

**Endosymbiont density → Gross photosynthesis (β\* = 0.523, p < 0.001)**
More symbiont cells per unit area produce more photosynthetic oxygen because photosynthetic capacity is approximately additive across cells. This path captures the cellular-level scaling of reef metabolism: whole-colony oxygen production scales with the number of photosynthetically active cells housed in the tissue, making symbiont density the proximate driver of gross photosynthesis rates.

**Gross photosynthesis → Respiration (β\* = 0.540, p < 0.001)**
Photosynthesis and respiration are tightly coupled in coral tissue. Photosynthate generated by symbionts fuels aerobic metabolism in both the symbiont and host cell compartments, so higher gross photosynthesis directly drives higher oxygen consumption. This P:R coupling also reflects shared carbon allocation: carbohydrates fixed during the day are respired by mitochondria in both partners, producing a predictable positive relationship between the two fluxes.

**Tissue biomass → Respiration (β\* = 0.149, p = 0.110) — not significant**
Once gross photosynthesis is accounted for, standing tissue biomass does not significantly predict respiration. This suggests that metabolic rate is governed more by immediate energy flux (current photosynthate supply) than by the total quantity of tissue present. In corals, maintenance respiration per unit biomass appears relatively constant; the variation in respiration rate across fragments is better explained by the photosynthetic activity of their symbionts than by how much tissue has accumulated.

---

## R² for endogenous variables

```{r frag-path-r2}
#| label: frag-path-r2

singular_tbl <- tibble(
  Response = c("log_chla", "log_endo", "log_biomass", "mean_R", "mean_GP"),
  Singular = c(isSingular(frag_m_chla), isSingular(frag_m_endo),
               isSingular(frag_m_biomass), isSingular(frag_m_r),
               isSingular(frag_m_gp))
)

frag_r2_tbl <- rsquared(frag_psem) |>
  dplyr::select(Response, `R² / R²m` = Marginal, `R²c` = Conditional) |>
  left_join(singular_tbl, by = "Response") |>
  mutate(
    Response = recode(Response,
      log_chla    = "Chl-a (log)",
      log_endo    = "Endosymbiont (log)",
      log_biomass = "Tissue biomass (log)",
      mean_R      = "Respiration",
      mean_GP     = "Gross photosynthesis"),
    across(where(is.numeric), ~round(., 3))
  )

frag_r2_tbl |>
  kbl(
    caption = paste0(
      "R\u00b2 per path equation (piecewiseSEM). R\u00b2m (marginal) reflects ",
      "fixed-effect predictors only; R\u00b2c (conditional) includes fixed + ",
      "random effects (Nakagawa & Schielzeth 2013). ",
      "Singular = one random-effect variance component estimated at zero."
    ),
    col.names = c("Response", "R\u00b2 / R\u00b2m", "R\u00b2c", "Singular fit")
  ) |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE, position = "left")
```

---

## Random effects variance components

```{r frag-re-variances}
#| label: frag-re-variances

re_tbl <- purrr::map2_dfr(
  list(frag_m_chla, frag_m_endo, frag_m_gp, frag_m_r, frag_m_biomass),
  c("Chl-a (log)", "Endosymbiont (log)", "Gross photosynthesis",
    "Respiration", "Tissue biomass (log)"),
  function(mod, nm) {
    as.data.frame(VarCorr(mod)) |>
      dplyr::filter(is.na(var2)) |>
      dplyr::transmute(
        Response = nm,
        Group    = grp,
        Variance = round(vcov, 4),
        SD       = round(sdcor, 4)
      )
  }
)

re_tbl |>
  kbl(
    caption = "Random-effect variance components for each fragment-level equation. Variance \u2248 0 indicates singularity (that component estimated at its boundary).",
    col.names = c("Response", "Group", "Variance", "SD")
  ) |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) |>
  collapse_rows(columns = 1, valign = "top")
```

---


# Supplementary figures {#sec-supp}

## Supp Fig 1 — Diel pH dynamics (24-hr experiment)

```{r load-24hr}
#| label: load-24hr

pHcalib_24 <- read_csv(here("Data", "Chemistry", "TrisCalSummer2024.csv"),
                        show_col_types = FALSE)
pHData_24  <- read_csv(here("Data", "Chemistry", "24_hr_carb_chem.csv"),
                        show_col_types = FALSE)

pH_all_24 <- pHcalib_24 |>
  nest_by(TrisCalDate) |>
  mutate(fitpH = list(lm(mVTris ~ TTris, data = data))) |>
  reframe(broom::tidy(fitpH)) |>
  dplyr::select(TrisCalDate, term, estimate) |>
  pivot_wider(names_from = term, values_from = estimate) |>
  right_join(pHData_24, by = "TrisCalDate") |>
  mutate(mVTris    = TEMPINLAB * TTris + `(Intercept)`) |>
  drop_na(TEMPINSITU, mV) |>
  mutate(
    pH_lab    = seacarb::pH(Ex = mV, Etris = mVTris, S = SALINITY, T = TEMPINLAB),
    pH_insitu = seacarb::pHinsi(pH = pH_lab, ALK = 2200, Tinsi = TEMPINSITU,
                                Tlab = TEMPINLAB, S = SALINITY, Pt = 0.1,
                                k1k2 = "m10", kf = "dg")
  )

InflowData_24 <- pH_all_24 |>
  filter(TANK_NUM %in% c("Inflow1", "Inflow2")) |>
  mutate(InflowTable = ifelse(TANK_NUM == "Inflow1", 1, 2)) |>
  dplyr::select(DATE, TIME, InflowTable, pH_inflow = pH_insitu)

Data_24 <- pH_all_24 |>
  filter(!TANK_NUM %in% c("Inflow1", "Inflow2")) |>
  mutate(TANK_NUM = as.numeric(TANK_NUM)) |>
  left_join(TableID, by = "TANK_NUM") |>
  left_join(InflowData_24, by = c("DATE", "TIME", "INFLOW_TABLE" = "InflowTable")) |>
  mutate(
    DATETIME  = ymd_hms(paste(as.character(DATE), TIME)),
    pHDiff    = pH_insitu - pH_inflow,
    TREATMENT = factor(TREATMENT, levels = treatment_levels)
  )
```

```{r fig-supp1}
#| label: fig-supp1
#| fig-cap: "Change in pH relative to inflow across a diel cycle in the 24-hour mesocosm experiment. Each line represents one tank; points show individual measurement times. Yellow shading indicates daytime; grey shading indicates nighttime. The final morning measurement (Jun 03 06:00–09:00) occurred under overcast conditions."
#| fig-width: 7
#| fig-height: 5

fig_supp1 <- Data_24 |>
  group_by(DATETIME, TANK_NUM, TREATMENT) |>
  summarise(avg_pH_diff = mean(pHDiff, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = DATETIME, y = avg_pH_diff)) +
  annotate("rect",
           xmin = ymd_hms("2024-06-02 12:00:00"),
           xmax = ymd_hms("2024-06-02 18:00:00"),
           ymin = -Inf, ymax = Inf, fill = "#FFFACD", alpha = 0.6) +
  annotate("rect",
           xmin = ymd_hms("2024-06-02 18:00:00"),
           xmax = ymd_hms("2024-06-03 06:00:00"),
           ymin = -Inf, ymax = Inf, fill = "#D3D3D3", alpha = 0.4) +
  annotate("rect",
           xmin = ymd_hms("2024-06-03 06:00:00"),
           xmax = ymd_hms("2024-06-03 09:00:00"),
           ymin = -Inf, ymax = Inf, fill = "#FFFACD", alpha = 0.6) +
  geom_hline(yintercept = 0, lty = 2, color = "grey40", linewidth = 0.4) +
  geom_line(aes(group = TANK_NUM, color = TREATMENT), linewidth = 0.7, alpha = 0.8) +
  geom_point(aes(color = TREATMENT), size = 2) +
  annotate("text", x = ymd_hms("2024-06-03 08:00:00"), y = 0.15,
           label = "Overcast\nDay", size = 3.2, color = "grey30") +
  scale_color_manual(values = trt_colors, labels = trt_labels, name = NULL) +
  scale_x_datetime(date_labels = "%b %d %H:%M", date_breaks = "6 hours") +
  labs(x = "Date & Time", y = "Change in pH") +
  theme_bw(base_size = 11) +
  theme(
    legend.position  = "bottom",
    legend.text      = element_text(size = 9),
    axis.title       = element_text(size = 10),
    axis.text        = element_text(size = 9),
    axis.text.x      = element_text(angle = 20, hjust = 1),
    panel.grid.minor = element_blank()
  )

ggsave(here("Revision", "Output", "supp_fig1_diel_pH.png"),
       fig_supp1, width = 7, height = 5, dpi = 300)

fig_supp1
```

---

## Supp Fig 2 — Coral physiology by treatment

```{r fig-supp2}
#| label: fig-supp2
#| fig-cap: "Coral fragment physiology by community treatment. Large points show treatment means; error bars are ±1 SD; small translucent points are individual fragments. Log₁₀ y-axis for Endosymbiont Density, Chlorophyll-a, and Tissue Biomass. No treatment effect was significant (mixed-effects models with genotype as random intercept, all p > 0.14)."
#| fig-width: 11
#| fig-height: 9

supp_panel_theme <- theme_bw(base_size = 11) +
  theme(
    axis.title       = element_text(size = 10),
    axis.text        = element_text(size = 9),
    axis.text.x      = element_text(angle = 30, hjust = 1),
    legend.position  = "none",
    panel.grid.minor = element_blank()
  )

make_physio_panel <- function(yvar, ylab, log_y = FALSE) {
  p <- ggplot(
    metadata |> mutate(TREATMENT = factor(TREATMENT, levels = treatment_levels)),
    aes(x = TREATMENT, y = .data[[yvar]], color = TREATMENT)
  ) +
    geom_point(alpha = 0.25, size = 1.8,
               position = position_jitter(width = 0.08, seed = 42)) +
    stat_summary(fun = mean, geom = "point", size = 3.5) +
    stat_summary(fun.data = mean_sdl, geom = "errorbar",
                 fun.args = list(mult = 1), width = 0.18, linewidth = 0.8) +
    scale_x_discrete(labels = trt_labels) +
    scale_color_manual(values = trt_colors) +
    labs(x = NULL, y = ylab) +
    supp_panel_theme
  if (log_y) p <- p + scale_y_log10()
  p
}

sp2a <- make_physio_panel("endo_per_cm2",
  expression(bold("Endosymbiont Density") ~ (cells %*% 10^6 ~ cm^{-2})), log_y = TRUE)
sp2b <- make_physio_panel("chla_ug_cm2",
  expression(bold("Chlorophyll-a Content") ~ (µg ~ cm^{-2})), log_y = TRUE)
sp2c <- make_physio_panel("mean_tissue_biomass",
  expression(bold("Tissue Biomass") ~ (mg ~ cm^{-2})), log_y = TRUE)
sp2d <- make_physio_panel("R",
  expression(bold("Respiration Rate") ~ (µmol ~ O[2] ~ cm^{-2} ~ hr^{-1})))
sp2e <- make_physio_panel("GP",
  expression(bold("Gross Photosynthesis") ~ (µmol ~ O[2] ~ cm^{-2} ~ hr^{-1})))

fig_supp2 <- (sp2a + sp2b + sp2c) / (sp2d + sp2e) +
  plot_annotation(tag_levels = "a")

ggsave(here("Revision", "Output", "supp_fig2_physiology_by_treatment.png"),
       fig_supp2, width = 11, height = 9, dpi = 300)

fig_supp2
```

---

## Supp Fig 3 — Physiology vs mean pH

```{r fig-supp3}
#| label: fig-supp3
#| fig-cap: "Coral physiology as a function of mean daytime pH (tank means, n = 16). Points colored by treatment. No relationships were significant at α = 0.05 (simple OLS; Chl-a: p = 0.09, Endosymbiont Density: p = 0.84, Tissue Biomass: p = 0.67, Respiration Rate: p = 0.06, Gross Photosynthesis: p = 0.50). Y-axes for Chl-a, Endosymbiont Density, and Tissue Biomass are on a log₁₀ scale."
#| fig-width: 12
#| fig-height: 9

supp_scatter_theme <- theme_bw(base_size = 11) +
  theme(
    axis.title       = element_text(size = 10),
    axis.text        = element_text(size = 9),
    legend.position  = "none",
    panel.grid.minor = element_blank()
  )

raw_frag_path <- metadata |>
  mutate(mean_tissue_biomass = ifelse(mean_tissue_biomass > 30, NA, mean_tissue_biomass)) |>
  left_join(
    chem_reframe_DAY_clean |>
      group_by(TANK_NUM) |>
      summarize(nep = mean(NEP, na.rm = TRUE),
                ph  = max(pH,   na.rm = TRUE),
                doc = mean(DOC, na.rm = TRUE),
                .groups = "drop"),
    by = "TANK_NUM"
  ) |>
  mutate(TREATMENT = factor(TREATMENT, levels = treatment_levels))

tank_phys_chem <- raw_frag_path |>
  group_by(TANK_NUM, TREATMENT) |>
  summarise(
    mean_chla    = mean(chla_ug_cm2,        na.rm = TRUE),
    mean_endos   = mean(endo_per_cm2,       na.rm = TRUE),
    mean_biomass = mean(mean_tissue_biomass, na.rm = TRUE),
    mean_R       = mean(R,                  na.rm = TRUE),
    mean_GP      = mean(GP,                 na.rm = TRUE),
    .groups = "drop"
  ) |>
  left_join(
    chem_reframe_DAY_clean |>
      group_by(TANK_NUM) |>
      summarise(grand_mean_pH  = mean(pH,  na.rm = TRUE),
                grand_mean_DOC = mean(DOC, na.rm = TRUE),
                .groups = "drop"),
    by = "TANK_NUM"
  ) |>
  mutate(TREATMENT = factor(TREATMENT, levels = treatment_levels))

make_scatter <- function(xvar, yvar, xlab, ylab, add_lm = FALSE, log_y = FALSE) {
  p <- ggplot(tank_phys_chem,
              aes(x = .data[[xvar]], y = .data[[yvar]], color = TREATMENT)) +
    geom_point(size = 2.5) +
    scale_color_manual(values = trt_colors, labels = trt_labels, name = NULL) +
    labs(x = xlab, y = ylab) +
    supp_scatter_theme
  if (add_lm) p <- p + geom_smooth(method = "lm", formula = y ~ x,
                                    color = "grey30", se = TRUE, linewidth = 0.8)
  if (log_y)  p <- p + scale_y_log10()
  p
}

ph_xlab  <- expression("pH"[T])
chla_ylab  <- expression("Chlorophyll-a Content (" * mu * "g cm"^{-2} * ")")
endo_ylab  <- expression("Endosymbiont Density (cells " * 10^6 ~ "cm"^{-2} * ")")
bio_ylab   <- expression("Mean Tissue Biomass (mg cm"^{-2} * ")")
R_ylab     <- expression("Respiration Rate (" * mu * "mol O"[2] * " cm"^{-2} * " hr"^{-1} * ")")
GP_ylab    <- expression("Gross Photosynthesis (" * mu * "mol O"[2] * " cm"^{-2} * " hr"^{-1} * ")")

sp3a <- make_scatter("grand_mean_pH", "mean_chla",    ph_xlab, chla_ylab, log_y = TRUE)
sp3b <- make_scatter("grand_mean_pH", "mean_endos",   ph_xlab, endo_ylab, log_y = TRUE)
sp3c <- make_scatter("grand_mean_pH", "mean_biomass", ph_xlab, bio_ylab,  log_y = TRUE)
sp3d <- make_scatter("grand_mean_pH", "mean_R",       ph_xlab, R_ylab)
sp3e <- make_scatter("grand_mean_pH", "mean_GP",      ph_xlab, GP_ylab)

fig_supp3 <- (sp3a + sp3b + sp3c) / (sp3d + sp3e) +
  plot_annotation(tag_levels = "a") +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom") &
  scale_color_manual(values = trt_colors, labels = trt_labels, name = NULL)

ggsave(here("Revision", "Output", "supp_fig3_physiology_vs_pH.png"),
       fig_supp3, width = 12, height = 9, dpi = 300)

fig_supp3
```

---

## Supp Fig 4 — Physiology vs mean DOC

```{r fig-supp4}
#| label: fig-supp4
#| fig-cap: "Coral physiology as a function of mean daytime DOC (tank means, n = 16). Points colored by treatment. No significant linear relationships were detected. Y-axes for Chl-a, Endosymbiont Density, and Tissue Biomass are on a log₁₀ scale."
#| fig-width: 12
#| fig-height: 9

doc_xlab <- expression("DOC (" * mu * "mol L"^{-1} * ")")

sp4a <- make_scatter("grand_mean_DOC", "mean_chla",    doc_xlab, chla_ylab, log_y = TRUE)
sp4b <- make_scatter("grand_mean_DOC", "mean_endos",   doc_xlab, endo_ylab, log_y = TRUE)
sp4c <- make_scatter("grand_mean_DOC", "mean_biomass", doc_xlab, bio_ylab,  log_y = TRUE)
sp4d <- make_scatter("grand_mean_DOC", "mean_R",       doc_xlab, R_ylab)
sp4e <- make_scatter("grand_mean_DOC", "mean_GP",      doc_xlab, GP_ylab)

fig_supp4 <- (sp4a + sp4b + sp4c) / (sp4d + sp4e) +
  plot_annotation(tag_levels = "a") +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom") &
  scale_color_manual(values = trt_colors, labels = trt_labels, name = NULL)

ggsave(here("Revision", "Output", "supp_fig4_physiology_vs_DOC.png"),
       fig_supp4, width = 12, height = 9, dpi = 300)

fig_supp4
```

---

*Document generated: `r Sys.time()`*

